---
title: "MATR3-specific editing sites identification"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
author:
  - name: "Kim Nguyen"
format:
  html:
    code-fold: TRUE
    code-overflow: scroll
    code-summary: "Show code"
    code-tools: true
    code-line-numbers: true
    toc: true
    toc-depth: 5
    toc-location: right
    toc-expand: true
    number-sections: true
    theme: sandstone
    fontsize: 11pt
    linestretch: 1.5
    fig-format: png
    cap-location: margin
    body-width: 900px
    sidebar-width: 300px
    page-layout: full
    crossref:
      fig-title: Fig
    embed-resources: true
    link-external-newwindow: true
    smooth-scroll: true
    css: css.css
    execute:
      echo: true
      warning: false
---
# Overview: 
This **HyperTRIBE** analysis pipeline walks through for identifying RNA editing sites that are specific to an RNA-binding protein (RBP) of interest, as opposed to background ADAR activity.

# Libraries
Genomic ranges: https://biodatascience.github.io/compbio/bioc/ranges.html 
```{r libraries, message=FALSE}
#BiocManager::install("BindingSiteFinder")
#install.packages("ComplexUpset")
# --- bioConductor packages & Genomic infrastructure ---
library(GenomicRanges)
library(GenomicFeatures)
library(GenomeInfoDb)
library(rtracklayer)
library(Biostrings)
library(AnnotationDbi)
library(txdbmaker)      
library(VariantAnnotation)

# --- Mus musculus mm39 ---
library(org.Mm.eg.db)
library(biomaRt)

# --- Functional enrichment ---
library(clusterProfiler)
library(gprofiler2)

# --- Tidyverse & plot visualization ---
library(tidyverse)
library(ggrepel)        # Prevent label overlap
library(ggrastr)        # Rasterize heavy plots (essential for large CLIP/RNA-seq datasets)
library(ggpointdensity) # High-density scatter plots
library(ggforce)        # Specialized geoms
library(ggpp)           # Annotations
library(shadowtext)     # Better text readability

# Venn/UpSet for intersection visual
library(ggvenn)
library(eulerr)
library(UpSetR)
#library(ComplexUpset)

# Color palettes
library(RColorBrewer)
library(colorRamp2)

# Plot
# library(networkD3)
# library(ggalluvial)
# library(corrplot)
# library(GGally)

# Layout
library(gridExtra)
library(patchwork)    
library(magick)        
library(kableExtra)     
library(plotly)         
library(markdown)
```

# Input data integration and Reference genome annotation

## Import BedGraph files

Each .bg file contains editing calls per nucleotide position. The code imports each as a GRanges, deduplicates by site ID (NA..19), then collapses every range to exactly 1 bp by setting start = end = end(gr) — isolating the precise editing coordinate.
```{r setup-bg}
# Folder path
folder <- "/Users/tinkienoob/Master_thesis/MATR3/HT_BedGraph_files"

# List all .bg files
files <- list.files(folder, pattern = "\\.bg$", full.names = TRUE)

# Import all bedGraph files into a named list
bg_list <- lapply(files, function(f) {
  gr <- import(f, format = "bedGraph")
  # Filter out duplicates and keep only single-nucleotide resolution
  gr <- gr[!duplicated(gr$NA..19)]
  # Standardize to 1bp site (the actual editing coordinate)
  ranges(gr) <- IRanges(start = end(gr), end = end(gr))
  return(gr)
})
names(bg_list) <- gsub("\\.txt\\.bg$", "", basename(files))

#n_distinct(unique(bg_list$Silke_HT_1_A2G$NA..19))
```

## Import TXT files (containing coverage/score metrics)
```{r setup-txt}
# List all .txt files 
txt_files <- list.files(folder, pattern = "\\.txt$", full.names = TRUE)

txt_list <- lapply(txt_files, function(f) {
  read.table(f, header = TRUE) %>%
    dplyr::rename(start = Edit_coord) %>%
    mutate(end = start, strand = "*") %>%
    makeGRangesFromDataFrame(keep.extra.columns = TRUE)
})

names(txt_list) <- gsub("\\.txt$", "", basename(txt_files))
```

## Build reference annotation

### Make annotation database from gtf file

The annotation file was downloaded from [GENCODE mouse release M37](https://www.gencodegenes.org/mouse/release_M37.html). The gencode.vM37.basic.annotation.gtf file corresponds to the GRCm39 (mm39) mouse genome assembly and was released in May 2025. Like other GENCODE "basic" releases, it provides a curated, high-confidence subset of transcript models restricted to reference chromosomes.

`txdbmaker::makeTxDbFromGFF()` builds a structural map containing only the coordinates strictly required to define gene structure — chromosome, start, end, strand, and feature IDs. 
`rtracklayer::import()`, in contrast, reads the complete set of descriptive attributes in the GTF file (e.g., gene_id, gene_name, gene_biotype, tag), with each attribute stored as its own column in the metadata (mcols). I combine both objects below: the TxDb supplies exact gene coordinates, and the GTF-derived metadata supplies the descriptive fields used downstream.
```{r annoDB-Info, message=FALSE}
annoFile <- "/Users/tinkienoob/Master_thesis/MATR3/mm39_anno/gencode.vM37.basic.annotation.gtf" 

# Build TxDb from GFF3: structural gene model (coordinates only, needed for genomic queries)
#annoDb <- txdbmaker::makeTxDbFromGFF(file = annoFile, format = "gtf")

# Save the TxDb object to file
#AnnotationDbi::saveDb(annoDb, "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/annoDb.Db")

# Import full annotation for metadata (gene_id, gene_name, gene_biotype, tag, etc.)
annoInfo <- rtracklayer::import(annoFile, format = "gtf")
```

```{r gns-build, message=FALSE}
# Load TxDb object from file
annoDb <- AnnotationDbi::loadDb("/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/annoDb.Db")

# Extract genes as GRanges, then enriched with metadata from annoInfo via match() on gene IDs.
gns <- genes(annoDb)

# Strip version suffixes from GENCODE IDs (e.g., .18) ---
#gns$gene_id <- sub("\\..*", "", gns$gene_id)
#annoInfo$gene_id <- sub("\\..*", "", annoInfo$gene_id)

# Match gene IDs between TxDb and Gtf
idx <- match(gns$gene_id, annoInfo$gene_id)

# Combine metadata from TxDb + Gtf
meta <- cbind(elementMetadata(gns),
              elementMetadata(annoInfo)[idx,])
meta <- meta[,!duplicated(colnames(meta))]

elementMetadata(gns) <- meta
```

## Build transcript region map 
```{r region-anno}
# Define regions from TxDb
# Count the overlaps of each binding site for each region of the transcript. 
  cdseq <- cds(annoDb) 
  intrns <- unlist(intronsByTranscript(annoDb), use.names=TRUE)  
  utrs3 <- unlist(threeUTRsByTranscript(annoDb), use.names=TRUE) 
  utrs5 <- unlist(fiveUTRsByTranscript(annoDb), use.names=TRUE) 
  regions_list <- GRangesList(UTR5 = utrs5, UTR3 = utrs3, CDS = cdseq, Intron = intrns)
  seqlevelsStyle(regions_list) <- "UCSC"

#view(as.data.frame(regions_list))

#saveRDS(regions_list, file = "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/mm39_region_list.rds")
```

## Build intron table
intrns lacks of transcript/ gene id information. To annotate introns with their corresponding transcript and gene IDs, I extract the transcript metadata from the annotation database and then join it with the intron GRanges based on the transcript IDs. 
```{r intrn-anno}
# Extract all transcripts (as GRanges)
transcripts_id <- transcripts(annoDb, columns = c("tx_id", "tx_name", "gene_id"))
seqlevelsStyle(transcripts_id) <- "UCSC"

tx_df <- tibble::tibble(
  tx_id   = as.character(mcols(transcripts_id)$tx_id),
  tx_name = as.character(mcols(transcripts_id)$tx_name),
  gene_id = as.character(unlist(mcols(transcripts_id)$gene_id))) %>%
  distinct()

# recover tx_id from names(intrns)
intrns_df <- tibble::tibble(
  seqnames = as.character(seqnames(intrns)),
  start    = start(intrns),
  end      = end(intrns),
  strand   = as.character(strand(intrns)),
  width    = width(intrns),
  tx_id    = as.character(names(intrns))) %>%
  left_join(tx_df, by = "tx_id")

# Convert into Granges
introns_gr <- GRanges(
  seqnames = intrns_df$seqnames,
  ranges   = IRanges(start = intrns_df$start, end = intrns_df$end),
  strand   = intrns_df$strand,
  tx_id    = intrns_df$tx_id,
  tx_name  = intrns_df$tx_name,
  gene_id  = intrns_df$gene_id)

#saveRDS(introns_gr, file = "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/mm39_introns_gr.rds")
```

## Annotate strand for each txt GRanges
BedGraph carries genomic coordinates; meanwhile TXT carries coverage statistics. We match them by coordinate (equal overlap), then look up strand from the gene
annotation.

```{r merge-bg-txt}
# Merge site-specific scores in txt to genomic coordinates in bg
gr_list <- lapply(names(txt_list), function(nm) {
  txt_gr <- txt_list[[nm]]
  bg_gr  <- bg_list[[nm]]
  
  # Find overlap coordinates
  hits <- findOverlaps(txt_gr, bg_gr, type = "equal")
  res <- txt_gr[queryHits(hits)]
  res$score <- bg_gr$score[subjectHits(hits)]
  res$Ranges_coord <- bg_gr$NA..19[subjectHits(hits)]
  
  # Annotate strand from GTF
  strand_hits <- findOverlaps(res, gns, ignore.strand = TRUE)
  # Assign strand of the first gene hit (typical for HyperTRIBE)
  strand(res)[queryHits(strand_hits)] <- strand(gns)[subjectHits(strand_hits)]
  
  return(res)
})

# Keep same names
names(gr_list) <- names(txt_list)

# checking unique genomic ranges (chrom, start, end, strand)
#table(duplicated(granges(gr_list$Silke_DRRM_HT_1_A2G)))

#saveRDS(gr_list, file = "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/GRanges_M3_HT_allsamples.rds")
```

## Custom function to save high-quality plots for thesis
```{r}
# ── Plot theme ──

theme_thesis <- function(base_size = 14) {
  theme_bw(base_size = base_size) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = base_size + 2),
      axis.title = element_text(face = "bold", size = base_size),
      axis.text = element_text(color = "black", size = base_size - 2),
      legend.title = element_text(size = base_size, face = "bold"),
      legend.text = element_text(size = base_size - 2),
      legend.position = "right")}


# ── Plot export ──

save_thesis_plot <- function(plot, filename,
                             width = 10, height = 6.5,
                             dpi = 300, bg = "white") {
  ggsave(
    filename = filename,
    plot = plot,
    width = width,
    height = height,
    units = "in",
    dpi = dpi,
    bg = bg)}
```

# Strategy to identify true MATR3-dependent editing from background/noise events.

Thresholds are **optimised empirically** before applying them.

```{r}
condition_map <- c(
  "Silke_M3_HT_1_A2G" = "MATR3", 
  "Silke_M3_HT_2_A2G" = "MATR3", 
  "Silke_M3_HT_3_A2G" = "MATR3", 
  "Silke_M3_HT_4_A2G" = "MATR3",
  "Silke_M3d_HT_1_A2G" = "ΔRRMcontrol", 
  "Silke_M3d_HT_2_A2G" = "ΔRRMcontrol", 
  "Silke_M3d_HT_3_A2G" = "ΔRRMcontrol", 
  "Silke_M3d_HT_4_A2G" = "ΔRRMcontrol",
  "Silke_HT_1_A2G" = "ADARcontrol", 
  "Silke_HT_2_A2G" = "ADARcontrol", 
  "Silke_HT_3_A2G" = "ADARcontrol", 
  "Silke_HT_4_A2G" = "ADARcontrol")
```

## 1st GLOBAL CUTOFF: Technical global filter

### Checking read distribution

Most genes have very low counts, but a few genes have very high counts (up to 16k).
That’s why filtering by a threshold is important—it removes a lot of weak/noisy signals at the bottom tail.
```{r coverage-quantile-check}
# Plot the quantile distribution of Total_count per sample to guide the coverage cutoff choice.
#quantile(df_list$Silke_HT_1_A2G$Total_count, probs = seq(0,1, by = .1))

par(mfrow=c(2,2)) # Loop through all data frames

for (x in names(gr_list)) {
  
  df <- gr_list[[x]]
  probs <- seq(0, 100, 10) # percentiles for x-axis: 0,10,…100 
  
  q <- quantile(df$Total_count, probs = seq(0, 1, 0.1))
  
  plot(probs, q, type = "b", pch = 16, lwd = 2, col = "#0E54B6FF", # type="b": Plots both points and lines connecting them.
       main = paste("Quantile distribution of count:", x),
       xlab = "Percentile (%)", ylab = "read counts",
       ylim = c(0, max(q)))
  
  grid(col = "grey80", lty = 3)
  text(probs, q, labels = round(q, 2), pos = 3, cex = 1, col = "#BD241EFF")
  abline(h = q["20%"], col = "#BD241EFF", lty = 2, lwd = 2)
}
```


```{r}
# ECDF of read coverage per condition.
coverage_df <- imap_dfr(gr_list, ~{
  as.data.frame(.x) %>%
    mutate(Sample = .y)
}) %>%
  mutate(
    Condition = condition_map[Sample],
    Replicate = sub(".*_([1-4])_A2G$", "rep\\1", Sample),
    Condition = factor(Condition, levels = c("MATR3", "ΔRRMcontrol", "ADARcontrol")),
    Replicate = factor(Replicate, levels = c("rep1", "rep2", "rep3", "rep4")))

rep_colors <- setNames(
  colorRampPalette(c("#08306B", "#C6DBEF"))(4),
  c("rep1", "rep2", "rep3", "rep4"))

conditions <- unique(coverage_df$Condition)

for (cond in conditions) {
  
  cat("### ", cond, "\n\n", sep = "")
  
  p <- ggplot(
    coverage_df %>% filter(Condition == cond),
    aes(x = Total_count, color = Replicate)) +
    stat_ecdf(geom = "point", size = 0.8) +
    geom_vline(xintercept = c(5, 10, 15), linetype = "dashed", color = "grey50") +
    scale_color_manual(values = rep_colors, drop = FALSE) +
    theme_bw() +
    labs(
      title = cond,
      x = "Read coverage",
      y = "Cumulative fraction of editing sites",
      color = "Replicates") +
    coord_cartesian(xlim = c(0, 100)) +
    theme_thesis()
  print(p)
  
  cat("\n\n")
}
```

### Filter editing sites with sufficient coverages

keep only sites with sufficient coverage (e.g. ≥ 5 reads = cut 20% dataset off)
```{r}
count_events <- function(gr_list, condition_map) {

# create empty df   
count_events <- data.frame(
    Condition = character(),
    Total_unique_events = numeric(),
    stringsAsFactors = FALSE)

# for conditions in condition_map
for (cond in unique(condition_map)) {
  # Each name of grange in the list = 1 replicate 
    reps <- names(condition_map[condition_map == cond])
    
    # Extract data, convert to DF, and bind rows 
    combined_before <- dplyr::bind_rows(lapply(reps, function(rep) {
      as.data.frame(gr_list[[rep]])
    }))
    
    # Count unique sites (based on Ranges_coord)
    Total_unique_events <- dplyr::n_distinct(combined_before$Ranges_coord)
    
    # Store results
    count_events <- rbind(count_events, data.frame(
      Condition = cond,
      Total_unique_events = Total_unique_events))}
  
  return(count_events)
}

count_events_df <- count_events(gr_list, condition_map)
```


### Loop for different cut-off levels for coverage (reads)
```{r coverage-cutoff}
cond_order <- c("MATR3", "ADARcontrol", "ΔRRMcontrol")

# Test different coverage thresholds
coverage_cutoff <- c(0, 5, 10, 15)

# Create lists to store filtered data and counts
coverage_cutoff_list <- list()
coverage_count_list  <- list()

for (k in coverage_cutoff) {
  
  # Filter each replicate based on coverage cutoff
  df_list_filtered <- lapply(names(gr_list), function(x) {
    gr <- gr_list[[x]]
    df <- as.data.frame(gr)
    
    if (k == 0) {
      df_filtered <- df
    } else {
      df_filtered <- df %>%
        dplyr::filter(Total_count >= k)}
    
    df_filtered %>%
      dplyr::select(Name, strand, Type, width, Ranges_coord, G_count, C_count, Total_count, score) %>%
      dplyr::distinct()
  })
  
  names(df_list_filtered) <- names(gr_list)
  coverage_cutoff_list[[as.character(k)]] <- df_list_filtered
  
  # Count total unique sites per condition
  counts <- count_events(df_list_filtered, condition_map)
  
  # Add stage label
  counts$Stage <- ifelse(
    k == 0,
    "No coverage cutoff",
    paste0("Coverage ≥ ", k))
  
  coverage_count_list[[as.character(k)]] <- counts
}

# Combine all results
coverage_count_list1 <- do.call(rbind, coverage_count_list)

# Order condition classes
coverage_count_list1$Condition <- factor(
  coverage_count_list1$Condition,
  levels = cond_order)

# Order stages for plotting
coverage_count_list1$Stage <- factor(
  coverage_count_list1$Stage,
  levels = c(
    "No coverage cutoff",
    "Coverage ≥ 5",
    "Coverage ≥ 10",
    "Coverage ≥ 15"))
```


```{r coverage-cutoff}
# Build plot object
coverage_optimize <- ggplot(
  coverage_count_list1,
  aes(x = Condition, y = Total_unique_events, fill = Stage)
) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.85) +
  geom_text(
    aes(label = scales::comma(Total_unique_events)),
    position = position_dodge(width = 0.9),
    vjust = 0,
    hjust = 0,
    size = 4.5,
    angle = 65
  ) +
  scale_fill_manual(values = c(
    "No coverage cutoff" = "#9E9E9E",
    "Coverage ≥ 5" = "#08306B",
    "Coverage ≥ 10" = "#6785AB",
    "Coverage ≥ 15" = "#96AFCB")) +
  labs(
    title = "Editing sites across coverage cutoffs",
    x = "",
    y = "Number of editing sites",
    fill = "Coverage cutoff"
  ) +
  theme_bw(base_size = 14) +
  theme_thesis()

# Show plot
coverage_optimize

```

## 2nd GLOBAL CUTOFF: Editing percentage/ score
How is score (editing score) per site calculated from bigWig coverage? 
Editing percentage/ score = (G/C count depend on +/- strand)/ Total count (coverage at that genomic position) * 100.

### Editing score bins by condition {.tabset}
```{r} 
# Collect events after 1st cutoff
df_list_tc5 <- coverage_cutoff_list[["5"]]

score_df <- imap_dfr(df_list_tc5, ~{
  .x %>%
    mutate(Sample = .y)
}) %>%
  mutate(
    Condition = condition_map[Sample],
    Replicate = sub(".*_([1-4])_A2G$", "rep\\1", Sample),
    Condition = factor(Condition, levels = c("MATR3", "ΔRRMcontrol", "ADARcontrol")),
    Replicate = factor(Replicate, levels = c("rep1", "rep2", "rep3", "rep4")))

conditions <- levels(score_df$Condition)

for (cond in conditions) {
  
  cat("### ", cond, "\n\n", sep = "")
  
  p <- ggplot(
  score_df %>% filter(Condition == cond),
  aes(x = score, color = Replicate)) +
  stat_ecdf(geom = "point", size = 0.8, alpha = 0.6) +
  geom_vline(xintercept = c(1, 5, 10), linetype = "dashed", color = "grey50") +
  scale_color_manual(values = rep_colors, drop = FALSE) +
  scale_x_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 10)) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = cond,
    x = "Editing score (%)",
    y = "Cumulative fraction of sites",
    color = "Replicates") +
  theme_bw(base_size =20) +
  theme_thesis()
  
  print(p)

  cat("\n\n")
}
```

### Loop for different cut-off levels
```{r}
edit_cutoff <- c(0, 1, 3, 5, 7, 9)

# Create lists to store filtered data and counts
edit_cutoff_list <- list()
edit_count_list  <- list()

for (k in edit_cutoff) {
  
  # Filter each replicate based on editing score cutoff
  df_list_filtered <- lapply(names(df_list_tc5), function(x) {
    gr <- df_list_tc5[[x]]
    df <- as.data.frame(gr)
    
    if (k == 0) {
      df_filtered <- df
    } else {
      df_filtered <- df %>%
        dplyr::filter(score >= k)}
    
    df_filtered %>%
      dplyr::distinct()
  })
  
  # use the correct names
  names(df_list_filtered) <- names(df_list_tc5)
  
  # store filtered replicate-level data
  edit_cutoff_list[[as.character(k)]] <- df_list_filtered
  
  # count total unique events per condition
  counts <- count_events(df_list_filtered, condition_map)
  
  # add stage label
  counts$Stage <- ifelse(
    k == 0,
    "No editing score cutoff",
    paste0("Score ≥ ", k, "%"))
  
  # store summary counts
  edit_count_list[[as.character(k)]] <- counts
}

# Combine all count tables
edit_count_list1 <- do.call(rbind, edit_count_list)

# Set factor order for conditions
edit_count_list1$Condition <- factor(
  edit_count_list1$Condition,
  levels = cond_order)

# Set factor order for stages
edit_count_list1$Stage <- factor(
  edit_count_list1$Stage,
  levels = c(
    "No editing score cutoff",
    "Score ≥ 1%",
    "Score ≥ 3%",
    "Score ≥ 5%",
    "Score ≥ 7%",
    "Score ≥ 9%"))

# Plot
editscore_cutoff <- ggplot(
  edit_count_list1,
  aes(x = Condition, y = Total_unique_events, fill = Stage)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = scales::comma(Total_unique_events)),
    position = position_dodge(width = 0.9),
    vjust = 0,
    hjust = 0,
    size = 4.5,
    angle = 65) +
  scale_fill_manual(values = c(
    "No editing score cutoff" = "#9E9E9E",
    "Score ≥ 1%" = "#08306B",
    "Score ≥ 3%" = "#375A8B",
    "Score ≥ 5%" = "#6785AB",
    "Score ≥ 7%" = "#96AFCB",
    "Score ≥ 9%" = "#C6DBEF")) +
  theme_bw(base_size = 14) +
  labs(
    title = "Editing sites after coverage ≥ 5 filtering across editing score cutoffs",
    x = "",
    y = "Number of editing sites",
    fill = "Cutoffs") +
  theme_thesis()

# df_list_tc5_e1 <- edit_cutoff_list[["1"]]
df_list_tc5_e5 <- edit_cutoff_list[["5"]]

editscore_cutoff
```

# Mean editing score of reproducible events
I calculate this using rowSums(!is.na(...)). This counts how many replicates actually detected the site. This becomes our 'Reproducibility Score'—if a site has a frequency of 1, it’s a singleton/noise. If it’s 4, it’s highly robust.

Mean Score: We calculate the average editing percentage across the active replicates. This gives us a single representative editing value for that genomic location

```{r}
ADAR_ctrl_list <- df_list_tc5_e5[grep("Silke_HT_", names(df_list_tc5_e5))]
MATR3_list     <- df_list_tc5_e5[grep("Silke_M3_HT_", names(df_list_tc5_e5))]
ΔRRM_list      <- df_list_tc5_e5[grep("Silke_M3d_HT_", names(df_list_tc5_e5))]
```


```{r}
# Reproducibility and scoring pipeline 
calculate_reproducibility <- function(gr_list, prefix) {
  # Calculate mean scores per replicate
  scores <- purrr::imap(gr_list, ~{
    .x %>%
      dplyr::select(Ranges_coord, score) %>%
      group_by(Ranges_coord) %>%
      summarise(score = mean(score, na.rm = TRUE), .groups = "drop") %>%
      mutate(Condition = paste0(prefix, "_rep", gsub(".*_([1-4])_A2G", "\\1", .y)))
  }) %>% bind_rows()

  # Pivot to wider format and calculate frequency
  scores %>%
    pivot_wider(
      id_cols = Ranges_coord,
      names_from = Condition,
      values_from = score) %>%
    mutate(
      Frequency = rowSums(!is.na(across(starts_with(paste0(prefix, "_rep"))))),
      Mean = rowMeans(across(starts_with(paste0(prefix, "_rep"))), na.rm = TRUE))
}

```

## ADAR control
```{r adar-score}
ADARctrl_scores_count <- calculate_reproducibility(ADAR_ctrl_list, "ADARcontrol")
ADARctrl_scores_count
```

## MATR3 WT
```{r matr3-score}
MATR3_wt_scores_count <- calculate_reproducibility(MATR3_list, "Matr3_wt")
MATR3_wt_scores_count
```

## Deplete RRM
```{r drrm-score}
ΔRRM_scores_count <- calculate_reproducibility(ΔRRM_list, "ΔRRM")
ΔRRM_scores_count
```

## How many events reproduce in more than 2 replicates?
```{r}
# Custom function to generate high-quality UpSet plots with color-coded reproducibility
plot_reproducible_upset <- function(upset_list, main_label, highlight_color = "#2171B5") {
  
  # Transform list to UpSet format
  upset_data <- UpSetR::fromList(upset_list)
  set_names <- names(upset_list)
  
  # highlight any intersection involving 2 or more replicates.
  combos <- unlist(lapply(2:length(set_names), function(x) {
    combn(set_names, x, simplify = FALSE)
  }), recursive = FALSE)
  
  # Create queries for UpSetR
  highlight_queries <- lapply(combos, function(combo) {
    list(query = UpSetR::intersects, 
         params = combo, 
         color = highlight_color, 
         active = TRUE)
  })

  # to dynamically color the bars representing reproducible events (>= 2 replicates)
  UpSetR::upset(
    upset_data,
    nsets = length(set_names),
    order.by = "freq",
    mainbar.y.label = main_label,
    sets.x.label = "Editing sites / Replicate",
    sets.bar.color = "#08306B",
    main.bar.color = "#08306B",
    queries = highlight_queries,
    text.scale = c(2.0, 1.6, 1.6, 1.4, 2.0, 2.2),
    number.angles = 60)
}

```

### Control (HT)
```{r upset-plots}
ADARctrl_upsetlist <- list(
  ADARctrl_r1 = unique(ADAR_ctrl_list$Silke_HT_1_A2G$Ranges_coord),
  ADARctrl_r2 = unique(ADAR_ctrl_list$Silke_HT_2_A2G$Ranges_coord),
  ADARctrl_r3 = unique(ADAR_ctrl_list$Silke_HT_3_A2G$Ranges_coord),
  ADARctrl_r4 = unique(ADAR_ctrl_list$Silke_HT_4_A2G$Ranges_coord))

plot_reproducible_upset(ADARctrl_upsetlist, "Shared editing sites (HT)")
```


## MATR3 WT
```{r}
MATR3_upsetlist <- list(
        MATR3_r1 = unique(MATR3_list$Silke_M3_HT_1_A2G$Ranges_coord),
        MATR3_r2 = unique(MATR3_list$Silke_M3_HT_2_A2G$Ranges_coord),
        MATR3_r3 = unique(MATR3_list$Silke_M3_HT_3_A2G$Ranges_coord),
        MATR3_r4 = unique(MATR3_list$Silke_M3_HT_4_A2G$Ranges_coord))

plot_reproducible_upset(MATR3_upsetlist, "Shared editing sites (M3_HT)")
```

### Depleted RRM (M3d_HT)
```{r}
ΔRRM_upsetlist <- list(ΔRRM_r1 = unique(ΔRRM_list$Silke_M3d_HT_1_A2G$Ranges_coord),
                          DRRM_r2 = unique(ΔRRM_list$Silke_M3d_HT_2_A2G$Ranges_coord),
                          ΔRRM_r3 = unique(ΔRRM_list$Silke_M3d_HT_3_A2G$Ranges_coord),
                          ΔRRM_r4 = unique(ΔRRM_list$Silke_M3d_HT_4_A2G$Ranges_coord))

plot_reproducible_upset(ΔRRM_upsetlist, "Shared editing sites (M3d_HT)")
```

## Apply reproducibility cutoff
```{r fn-reprod-event}
get_reproducible_events <- function(df) {
  
  df_reproducible <- df[df$Frequency >= 2,]
  
  return(df_reproducible)}

ADARctrl_reproducible <- get_reproducible_events(ADARctrl_scores_count) 
MATR3_wt_reproducible <- get_reproducible_events(MATR3_wt_scores_count)
ΔRRM_reproducible <- get_reproducible_events(ΔRRM_scores_count)

reprod_list <- list(
  ADARctrl = ADARctrl_reproducible,
  MATR3 = MATR3_wt_reproducible,
  ΔRRMctrl = ΔRRM_reproducible)
```


# Count events after each cut-off 
```{r}
count_events <- function(gr_list, condition_map) {
  
count_events <- data.frame(
    Condition = character(),
    Total_unique_events = numeric(),
    stringsAsFactors = FALSE
  )

for (cond in unique(condition_map)) {
    reps <- names(condition_map[condition_map == cond])
    combined_before <- data.frame()
    
    for (rep in reps) {
      gr <- gr_list[[rep]]
      df <- as.data.frame(gr)
      combined_before <- dplyr::bind_rows(combined_before, df)
    }
    
    # Count unique sites (based on Ranges_coord)
    Total_unique_events <- dplyr::n_distinct(combined_before$Ranges_coord)
    
    # Store results
    count_events <- rbind(count_events, data.frame(
      Condition = cond,
      Total_unique_events = Total_unique_events
    ))
  }
  
  return(count_events)
}

reprod_conds <- c("ADARctrl" = "ADARcontrol",
                  "MATR3" = "MATR3",
                  "ΔRRMctrl" = "ΔRRMcontrol")
                  
total_counts <- count_events(gr_list, condition_map)

# calculate all events after 1st cutoff
counts_after_tc5 <- count_events(df_list_tc5, condition_map)

# editing cutoffs ≥5%
counts_after_e5 <- count_events(df_list_tc5_e5, condition_map)

# reprod cutoff ≥2 replicates 
counts_after_r2 <- count_events(reprod_list, reprod_conds)

HT_colors <- setNames(
  c("#BD241EFF", "#F5B355", "cornflowerblue"),
  c("ADARcontrol", "MATR3", "ΔRRMcontrol")
)
```

```{r}
total_counts <- count_events(gr_list, condition_map)

# calculate all events after 1st cutoff
counts_after_tc5 <- count_events(df_list_tc5, condition_map)

# editing cutoffs ≥5%
counts_after_e5 <- count_events(df_list_tc5_e5, condition_map)

# reprod cutoff ≥2 replicates 
counts_after_r2 <- count_events(reprod_list, reprod_conds)
```

```{r}
barplot_count <- bind_rows(
  total_counts     %>% mutate(Stage = "No cutoff"),
  counts_after_tc5 %>% mutate(Stage = "Coverage ≥ 5"),
  counts_after_e5  %>% mutate(Stage = "Score ≥ 5%"),
  counts_after_r2  %>% mutate(Stage = "≥ 2 replicates")
)

barplot_count$Stage <- factor(
  barplot_count$Stage,
  levels = c(
    "No cutoff",
    "Coverage ≥ 5",
    "Score ≥ 5%",
    "≥ 2 replicates"))

barplot_count$Condition <- factor(
  barplot_count$Condition,
  levels = cond_order)

compare_cutoff <- ggplot(
  barplot_count,
  aes(x = Condition, y = Total_unique_events, fill = Stage)) +
  geom_bar(
    stat = "identity",
    position = "dodge",
    width = 0.8) +
  geom_text(
    aes(label = scales::comma(Total_unique_events)),
    position = position_dodge(width = 0.9),
    vjust = 0,
    hjust = 0,
    size = 5.5,
    angle = 65) +
  scale_fill_manual(values = c(
    "No cutoff" = "#9E9E9E",
    "Coverage ≥ 5" = "#08306B",
    "Score ≥ 5%" = "#6785AB",
    "≥ 2 replicates" = "#96AFCB")) +
  scale_y_continuous(
    limits = c(0, max(barplot_count$Total_unique_events) * 1.4)) +
  theme_bw(base_size = 15) +
  labs(
    title = "Editing sites retained across sequential filtering",
    x = "",
    y = "Number of editing sites",
    fill = "Filtering step") +
  theme_thesis()

# save_thesis_plot(
#   compare_cutoff,
#   filename = "cutoff_comparison_barplot.png",
#   width = 8.5,
#   height = 6)

compare_cutoff
```

# Identify specific MATR3-dependent editing sites

## Remove sites look alike background noise to collect high confident sites

The negative control (no MATR3) captures non-specific signal (background): mapping artifacts, sequencing errors, contaminating gDNA, or editing triggered by Hypertribe. If a site appears robustly in the negative control, it’s likely not specifically induced by MATR3. Therefore, removing negative-control and overlapping among negative control and WT and dRRM sites reduces false positives.

```{r}
ggven_reproducible_events <- list(
  ADARcontrol = ADARctrl_reproducible$Ranges_coord, 
  MATR3_wt = MATR3_wt_reproducible$Ranges_coord,
  ΔRRMcontrol = ΔRRM_reproducible$Ranges_coord)

plot(eulerr::euler(ggven_reproducible_events),
       fills = list(fill = c("#BD241EFF", "#2171B5", "#F5B355"), alpha = 0.5),
       quantities = TRUE,
       stroke_size = 1, # control border size
       set_name_size = 1, # set name size
       text_size = 2)

```


```{r}
custom_colors <- c(
  "ADARcontrol" = "grey",      
  "MATR3_wt" = "#2171B5",           # highlight WT unique
  "ΔRRMcontrol" = "grey",         
  "ADARcontrol&MATR3_wt" = "grey",
  "ADARcontrol&ΔRRMcontrol" = "grey",
  "MATR3_wt&ΔRRMcontrol" = "grey", 
  "ADARcontrol&MATR3_wt&ΔRRMcontrol" = "grey")

plot(
  eulerr::euler(ggven_reproducible_events),
  fills = list(
    fill  = custom_colors,
    alpha = 0.5),
  labels = list(
    fontsize = 16),
  quantities = list(
    fontsize = 16,
    fontface = "bold"),
  stroke_size = 1)

```

## Define unique and overlapping events among 3 conditions
```{r}
# Get overlap among all 3 datasets
overlap_upon_all <- Reduce(intersect, list(ADARctrl_reproducible$Ranges_coord, 
                                    MATR3_wt_reproducible$Ranges_coord,
                                    ΔRRM_reproducible$Ranges_coord))

# Remove events belonging to ADAR control 
ADAR_MATR3_overlap <- intersect(ADARctrl_reproducible$Ranges_coord,
                                  MATR3_wt_reproducible$Ranges_coord)

ADAR_ΔRRM_overlap <- intersect(ADARctrl_reproducible$Ranges_coord,
                                   ΔRRM_reproducible$Ranges_coord)

################################ Overlap upon DRRM and WT
MATR3_ΔRRM_overlap_only <- setdiff(
  intersect(MATR3_wt_reproducible$Ranges_coord,
            ΔRRM_reproducible$Ranges_coord),
            ADARctrl_reproducible$Ranges_coord)

reprod_MATR3_ΔRRM_df <-  MATR3_wt_reproducible[MATR3_wt_reproducible$Ranges_coord %in% MATR3_ΔRRM_overlap_only, ]


```

```{r}

################################ DRRM control unique
reprod_ΔRRM <- setdiff(ΔRRM_reproducible$Ranges_coord, union(ADAR_ΔRRM_overlap, MATR3_ΔRRM_overlap_only))

reprod_uniq_ΔRRM_df <- ΔRRM_reproducible[ΔRRM_reproducible$Ranges_coord %in% reprod_ΔRRM, ]

################################ WT unique
reprod_MATR3 <- setdiff(MATR3_wt_reproducible$Ranges_coord, union(ADAR_MATR3_overlap, MATR3_ΔRRM_overlap_only)) 

reprod_uniq_MATR3_df <- MATR3_wt_reproducible[MATR3_wt_reproducible$Ranges_coord %in% reprod_MATR3, ]

################################ ADAR control unique
reprod_ADARctrl <- setdiff(ADARctrl_reproducible$Ranges_coord, union(ADAR_MATR3_overlap, ADAR_ΔRRM_overlap))

reprod_uniq_ADAR_df <- ADARctrl_reproducible[ADARctrl_reproducible$Ranges_coord %in% reprod_ADARctrl, ]
```


## Convert specific site df into GRanges and annotate
```{r}
df_convert_gr <- function(df) {
  df2 <- df %>%
 # Split Ranges_coord into chr and numeric position
  tidyr::separate(Ranges_coord, into = c("seqnames", "ranges"), sep = "_") %>%
  mutate(
    ranges = as.numeric(ranges),
    score = round(Mean, 2)) %>%
  dplyr::select(seqnames, ranges, score)

  gr <- makeGRangesFromDataFrame(
    df2,
    seqnames.field = "seqnames",
    start.field    = "ranges",
    end.field      = "ranges",
    keep.extra.columns = TRUE)
  return(gr)}
############################################ All events 
ADARctrl_reproducible_gr <- df_convert_gr(ADARctrl_reproducible)
MATR3_wt_reproducible_gr <- df_convert_gr(MATR3_wt_reproducible)
ΔRRM_reproducible_gr <- df_convert_gr(ΔRRM_reproducible)

############################################ Reprod events
reprod_uniq_ΔRRM_gr <- df_convert_gr(reprod_uniq_ΔRRM_df)
reprod_uniq_MATR3_gr <- df_convert_gr(reprod_uniq_MATR3_df)
reprod_uniq_ADAR_gr <- df_convert_gr(reprod_uniq_ADAR_df)
reprod_MATR3_ΔRRM_gr <- df_convert_gr(reprod_MATR3_ΔRRM_df)

# Combine all reproducible GRanges
m3specific_sites <- reprod_uniq_MATR3_gr
```


```{r}
annotate_genenames <- function(gr, gns) {
  
  hits <- GenomicRanges::findOverlaps(gr, gns, type = "within", ignore.strand = FALSE)
  gr2 <- gr[queryHits(hits)]
  strand(gr2) <- strand(gns)[subjectHits(hits)]
  
  available_cols <- intersect(c("gene_name", "gene_id", "gene_type", "tag"), colnames(mcols(gns)))
  
  mcols(gr2) <- cbind(
    mcols(gr2),
    mcols(gns)[subjectHits(hits), available_cols, drop = FALSE]
  )
  
  as.data.frame(gr2) %>%
    dplyr::select(any_of(c("gene_name", "gene_id", "seqnames", "start", "end", 
                           "strand", "score", "gene_type", "tag", "region"))) %>%
    dplyr::distinct()}

############## specific sites
m3specific_sites2 <- annotate_genenames(m3specific_sites, gns)
#high_conf_sites2 <- high_conf_sites2 %>%
#  mutate(GeneID = sub("\\..*", "", gene_id))
# saveRDS(m3specific_sites2, "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/m3specific_sites2.rds")
```

# Genes of interest list (from collaborators)
```{r}
# Gene of interest
#mmusculus <- useMart("ENSEMBL_MART_ENSEMBL", dataset="mmusculus_gene_ensembl")

#listFilters(bm)
#listAttributes(bm)

genes_of_interest <- c(
  "Rbm20","Matr3","Xist","Ttn","Cacna1a","Cacna1b","Cacna1c","Cacna1d",
  "Ldb3","Pdlim5","Camk2a","Camk2d","Camk2g","Rbm14","Rbm24","Rbm39",
  "Rbm8a","Rbpms","Rbpms2","Mbnl1","Mbnl2","Nppa","Nppb","Myh6","Myh7",
  "Hnrnpu","Srsf3","Tnnt2","Des","Acta1","Actn2","Mybpc1","Mybpc3","Myocd")

collagen_genes <- c(
  "Col1a1","Col1a2","Col2a1","Col3a1","Col4a1","Col4a2","Col4a3","Col4a4",
  "Col4a5","Col4a6","Col5a1","Col5a2","Col5a3","Col6a1","Col6a2","Col6a3",
  "Col6a4","Col6a5","Col6a6","Col7a1","Col8a1","Col8a2","Col9a1","Col9a2",
  "Col9a3","Col10a1","Col11a1","Col11a2","Col12a1","Col13a1","Col14a1",
  "Col15a1","Col16a1","Col17a1","Col18a1","Col19a1","Col20a1","Col22a1",
  "Col23a1","Col24a1","Col25a1","Col26a1","Col27a1","Col28a1","Colgalt1",
  "Colgalt2","Colq")
```


## Count events per region per dataset
```{r}
region_anno_events <- function(gr, regions_list) {
  
  # initialize metadata
  region    <- rep(NA_character_, length(gr))
  region_id <- rep(NA_character_, length(gr))
  
  # loop over regions in GRangesList
  for (x in names(regions_list)) {
    hits <- GenomicRanges::findOverlaps(gr, regions_list[[x]], type="any", ignore.strand=FALSE)
    
    region[queryHits(hits)[is.na(region[queryHits(hits)])]] <- x
  }
  
  # add metadata back to gr
  mcols(gr)$region    <- region
  mcols(gr)$region_id <- region_id
  
  return(gr)
}

################################ specific sites
m3specific_sites2_gr <- makeGRangesFromDataFrame(m3specific_sites2,
    seqnames.field = "seqnames",
    start.field    = "start",
    end.field      = "end",
    keep.extra.columns = TRUE)
m3specific_sites2_gr_anno    <- region_anno_events(m3specific_sites2_gr, regions_list)
gns <- region_anno_events(gns, regions_list)

# saveRDS(gns, file = "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/gns_mm39.rds")
#as.data.frame(high_conf_sites2_gr_anno)

gene_anno <- as.data.frame(gns) %>%
  dplyr::select(gene_id = gene_id, gene_name = gene_name, gene_type = gene_type) %>%
  distinct() %>% 
  mutate(GeneID = sub("\\..*", "", gene_id)) # remove version suffix from gene_id

# saveRDS(gene_anno, file = "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/geneanno_df_mm39.rds")
```


# Create list of control 
```{r}
all_control_list <- list(
  only_ADARctrl = reprod_uniq_ADAR_df, # Background
  only_ΔRRMctrl = reprod_uniq_ΔRRM_df) # Different enrichment, different specificity

############## control annotation function
control_annotate_func <- function(df, gns, regions_list) {

  gr <- df_convert_gr(df)                 
  df_gene <- annotate_genenames(gr, gns)  # returns a DF

  gr2 <- makeGRangesFromDataFrame(df_gene,
    seqnames.field="seqnames", start.field="start", end.field="end",
    keep.extra.columns=TRUE)
  
  GenomeInfoDb::seqlevelsStyle(gr2) <- "UCSC"

  gr2 <- region_anno_events(gr2, regions_list)

  as.data.frame(gr2) %>%
    mutate(GeneID = sub("\\..*", "", gene_id))}

control_anno_gr <- lapply(all_control_list,
                          control_annotate_func,
                          gns = gns,
                          regions_list = regions_list)

#saveRDS(control_anno_gr, file = "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/HT_control_list.rds")
```


## Genomic region distribution of Matrin3-specific sites 
```{r}
region_colors <- c(
  "UTR5"   = "#3288BD", 
  "CDS"    = "#99D594", 
  "Intron" = "#FDAE61", 
  "UTR3"   = "#D53E4F")  # Neutral grey

# set class order
class_order <- c("M3-specific", "ADAR specific", "ΔRRM specific")

# set region order manually if you want full control
# change this to match your real annotation names exactly
region_order <- c("UTR5", "CDS", "UTR3", "exon", "Intron")

# if your region names are different, for example:
# region_order <- c("five_prime_utr", "cds", "three_prime_utr", "exon", "intron", "Others")

# count number of each region in M3-specific sites
m3specific_tx_region_counts <- as.data.frame(m3specific_site_gr_anno) %>%
  filter(!is.na(region)) %>%                    # drop unassigned sites
  group_by(region) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(percentage = count / sum(count) * 100,
         class = "M3-specific")

# count number of each region in controls
all_control_region_counts_list <- setNames(
  lapply(names(all_control_list_HT), function(cond) {
    df <- all_control_list_HT[[cond]]
    
    df %>%
      mutate(region = ifelse(is.na(region) | region == "", "Others", region)) %>%
      group_by(region) %>%
      summarise(count = n(), .groups = "drop") %>%
      mutate(
        percentage = count / sum(count) * 100,
        class = cond)}),
  names(all_control_list_HT))

all_control_region_counts <- bind_rows(all_control_region_counts_list) %>%
  mutate(class = recode(
    class,
    "only_ADARctrl" = "ADAR specific",
    "only_ΔRRMctrl" = "ΔRRM specific"))

# combine

regions_count_data <- bind_rows(m3specific_tx_region_counts,
                                all_control_region_counts) %>%
  mutate(
    class  = factor(class,  levels = class_order),
    region = factor(region, levels = region_order)) %>%
  filter(!is.na(region)) %>%           
  arrange(class, region)
```


```{r}
##################### Stacked barplot
ggplot(regions_count_data, aes(x = class, y = percentage, fill = region)) +
  geom_col(color = "black", linewidth = 0.25, width = 0.6) +
  geom_text(
    aes(label = ifelse(percentage >= 3, paste0(round(percentage, 1), "%"), "")),
    position = position_stack(vjust = 0.5),
    size = 5.5,
    fontface = "bold") +
  scale_fill_manual(values = region_colors, drop = FALSE) +
  labs(
    title = "Genomic region distribution",
    x = NULL, y = "Proportion of editing sites", 
    fill = "Region") +
  theme_bw(base_size = 14) +
  theme_thesis()

##################### Group barplot
ggplot(regions_count_data, aes(x = class, y = count, fill = region)) +
  geom_col(
    color = "black", linewidth = 0.25, width = 0.75,
    position = position_dodge2(width = 0.8, preserve = "single")) +
  geom_text(
    aes(label = count),
    position = position_dodge2(width = 0.8, preserve = "single"),
    vjust = -0.4,
    size = 3) +
  scale_fill_manual(values = region_colors, drop = FALSE) +
  labs(
    title = "Genomic region distribution",
    x = NULL, y = "Number of editing sites", fill = "Region") +
  theme_bw(base_size = 14) +
  theme_thesis()
```


# Number of high confident site in each target genes
```{r}
m3specific_sites_per_GOI <- m3specific_sites_per_genes %>%
  filter(Target_genes %in% c("GOI", "Collagen"))

ggplot(m3specific_sites_per_GOI, aes(x = reorder(gene_name, -count), y = count, fill = Target_genes)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Number of M3-specific editing sites per GOI",
    x = "Gene",
    y = "Number of M3-specific editing sites",
    fill = "Gene set"
  ) +
  scale_fill_manual(values = c("GOI" = "#F5B355", "Collagen" = "cornflowerblue")) +
  theme_minimal() + 
  theme_thesis()
```

# GO term analysis of genes containing Matr3-specific sites 
```{r go-analysis func}
plot_go_enrichment <- function(ego_obj, set_name, n_terms = 6) {

  plot_df <- as.data.frame(ego_obj) %>%
    tidyr::separate(GeneRatio, into = c("gene_count", "bg_count"),
                    sep = "/", convert = TRUE) %>%
    mutate(GeneRatio_num = gene_count / bg_count) %>%
    filter(
      !is.na(GeneRatio_num), is.finite(GeneRatio_num),
      !is.na(p.adjust), !is.na(Count)) %>%
    group_by(ONTOLOGY) %>%
    slice_min(p.adjust, n = n_terms, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      term_label = str_wrap(Description, width = 35),
      ONTOLOGY   = dplyr::recode(ONTOLOGY,
        "BP" = "Biological Process",
        "CC" = "Cellular Component",
        "MF" = "Molecular Function"),
      ONTOLOGY = factor(ONTOLOGY,
        levels = c("Biological Process", "Cellular Component", "Molecular Function")))

  ggplot(plot_df,
    aes(x = GeneRatio_num,
        y = reorder(term_label, GeneRatio_num),
        color = p.adjust,
        size  = Count)) +
    geom_point(alpha = 0.7) +
    facet_grid(ONTOLOGY ~ ., scales = "free_y", space = "free_y") +
    scale_color_continuous(low = "red", high = "blue", trans = "reverse") +
    scale_size_continuous(name = "Gene count") +
    labs(
      title = paste("GO enrichment —", set_name),
      x     = "Gene ratio (target genes / total genes in term)",
      y     = NULL,
      color = "Adjusted\np-value") +
    theme_light(base_size = 14) +
    theme(
      strip.text       = element_text(size = 14, face = "bold"),
      strip.background = element_rect(fill = "grey20"),
      axis.text        = element_text(color = "black", size = 12),
      plot.title       = element_text(hjust = 0.5, face = "bold"),
      plot.margin      = margin(10, 10, 10, 10))
}
```

## GO enrichment of genes containing MATR3-specific sites {.tabset}
```{r}
# Get rid of suffix in gene_id to match with OrgDb keys
m3specific_sites2 <- m3specific_sites2 %>%
  mutate(GeneID = sub("\\..*", "", gene_id))

# Run GO enrichment separately for each ontology
ego_m3_specific <- enrichGO(
  gene          = unique(m3specific_sites2$GeneID),
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENSEMBL",
  ont           = "ALL",
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.05,
  readable      = TRUE)

# Build plots
ego_m3_specific_plot <- plot_go_enrichment(
  ego_obj   = ego_m3_specific,
  set_name  = "specific targets",
  n_terms   = 8)

ego_m3_specific_plot

```


# Biotype enrichment of top targets
```{r}
highconf_sites_biotype_count <- m3specific_site %>%
  group_by(gene_type) %>%
  summarise(n_genes = n(), 
            .groups = "drop") %>%
  arrange(desc(n_genes))

ggplot(highconf_sites_biotype_count, aes(x= reorder(gene_type, n_genes), y = n_genes)) +
  geom_bar(stat = "identity", 
          fill = "steelblue",
          alpha = 0.8) + 
  theme_light(base_size = 14) + 
  labs(
    title = "Gene Biotypes among Matrin3-specific targets", 
    x = "Gene Biotype",
    y = "Number of genes") + coord_flip()
```

# Export into bedGraph format for IGV visualization
```{r}
seqlevelsStyle(m3specific_sites2_gr_anno) <- "UCSC"

m3specific_sites2_gr_anno_bg <- m3specific_sites2_gr_anno
start(m3specific_sites2_gr_anno_bg) <- start(m3specific_sites2_gr_anno_bg) - 1
end(m3specific_sites2_gr_anno_bg)   <- start(m3specific_sites2_gr_anno_bg)       # 1bp

# export(m3specific_sites2_gr_anno_bg, con = "../m3specific_sites.bedGraph", format = "bedGraph")
```


# Session info
```{r}
sessionInfo()
```


