---
title: "MATR3-specific editing sites identification"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
author:
  - name: "Kim Nguyen"
format:
  html:
    code-fold: TRUE
    code-overflow: scroll
    code-summary: "Show code"
    code-tools: true
    code-line-numbers: true
    toc: true
    toc-depth: 5
    toc-location: right
    toc-expand: true
    number-sections: true
    theme: sandstone
    fontsize: 11pt
    linestretch: 1.5
    fig-format: png
    cap-location: margin
    body-width: 900px
    sidebar-width: 300px
    page-layout: full
    crossref:
      fig-title: Fig
    embed-resources: true
    link-external-newwindow: true
    smooth-scroll: true
    css: css.css
    execute:
      echo: true
      warning: false
---

# Library
```{r setup, message=FALSE}
#BiocManager::install("BindingSiteFinder")
#install.packages("ComplexUpset")
# --- 1. Core BioConductor & Genomic Infrastructure ---
# Essential for handling GRanges, TxDb objects, and sequence data
library(GenomicRanges)
library(GenomicFeatures)
library(GenomeInfoDb)
library(rtracklayer)
library(Biostrings)
library(AnnotationDbi)
library(txdbmaker)         # Replaces older MakeTxDb functions
library(VariantAnnotation)

# --- 2. Organism Specific (Mus musculus mm39) ---
library(org.Mm.eg.db)
library(biomaRt)

# --- 3. Tidyverse & Data Wrangling ---
# 'tidyverse' automatically loads: ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr, forcats
library(tidyverse) 
library(magrittr)   # For advanced piping (%<>%, %T>%)

# --- 4. Functional Enrichment & Analysis ---
library(clusterProfiler)
library(gprofiler2)

# --- 5. Visualization & Aesthetics ---
# General Plotting Enhancements
library(ggrepel)        # Prevent label overlap
library(ggrastr)        # Rasterize heavy plots (essential for large CLIP/RNA-seq datasets)
library(ggpointdensity) # High-density scatter plots
library(ggforce)        # Specialized geoms
library(ggpp)           # Annotations
library(shadowtext)     # Better text readability

# Color Palettes
library(RColorBrewer)
library(Polychrome)
library(colorRamp2)

# Set Logic & Intersections (Venn/UpSet)
library(ggvenn)
library(eulerr)
library(UpSetR)
# library(ComplexUpset) # Uncomment if you need the ggplot-native version

# Specialized Plots (Networks, Alluvial, Correlations)
library(networkD3)
library(ggalluvial)
library(corrplot)
library(GGally)

# Layout & Reporting
library(gridExtra)      # Arrange multiple plots
library(patchwork)     # (Recommended alternative to gridExtra for ggplot2)
library(magick)         # Image processing
library(kableExtra)     # Publication-quality tables
library(plotly)         # Interactive plots
library(markdown)
```
Genomic ranges: https://biodatascience.github.io/compbio/bioc/ranges.html 

# Process data
## BED files import
```{r}
# Folder path
folder <- "/Users/tinkienoob/Master_thesis/MATR3/HT_BedGraph_files"

# List all .bg files
files <- list.files(folder, pattern = "\\.bg$", full.names = TRUE)

# Import all bedGraph files into a named list
bg_list <- lapply(files, function(f) {
  gr <- import(f, format = "bedGraph")
  # Filter out duplicates and keep only single-nucleotide resolution
  gr <- gr[!duplicated(gr$NA..19)]
  # Standardize to 1bp site (the actual editing coordinate)
  ranges(gr) <- IRanges(start = end(gr), end = end(gr))
  return(gr)
})
names(bg_list) <- gsub("\\.txt\\.bg$", "", basename(files))

#n_distinct(unique(bg_list$Silke_HT_1_A2G$NA..19))
```

# Process TXT files (containing coverage/score metrics)
```{r}
# List all .txt files 
txt_files <- list.files(folder, pattern = "\\.txt$", full.names = TRUE)

txt_list <- lapply(txt_files, function(f) {
  read.table(f, header = TRUE) %>%
    dplyr::rename(start = Edit_coord) %>%
    mutate(end = start, strand = "*") %>%
    makeGRangesFromDataFrame(keep.extra.columns = TRUE)
})

names(txt_list) <- gsub("\\.txt$", "", basename(txt_files))
```

## Genomic features annotation
### Make annotation database from gtf file
annotation file is downloaded from: https://www.gencodegenes.org/mouse/release_M37.html
The gencode.vM37.basic.annotation.gtf file is designed for the mouse genome assembly GRCm39 (mm39). Released around May 2025, this GENCODE dataset (vM37) 

makeTxDbFromGFF builds a structural map including Only the strictly necessary coordinates to define gene structure (Chromosome, Start, End, Strand, IDs)
while rtracklayer::import reads the full descriptive content in the GTF file (e.g., gene_id, gene_name, gene_biotype, tag) becomes a column in the metadata (mcols).

```{r message=FALSE}
annoFile = "/Users/tinkienoob/Master_thesis/MATR3/mm39_anno/gencode.vM37.basic.annotation.gtf" 

# Build TxDb from GFF3
annoDb <- txdbmaker::makeTxDbFromGFF(file = annoFile, format = "gtf")

# Import full annotation for metadata
annoInfo <- rtracklayer::import(annoFile, format = "gtf")

# Save the TxDb object to file
AnnotationDbi::saveDb(annoDb, "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/annoDb.Db")

# Extract genes as GRanges
gns <- genes(annoDb)

# Strip version suffixes from GENCODE IDs (e.g., .18) ---
#gns$gene_id <- sub("\\..*", "", gns$gene_id)
#annoInfo$gene_id <- sub("\\..*", "", annoInfo$gene_id)

# Match gene IDs between TxDb and Gtf
idx <- match(gns$gene_id, annoInfo$gene_id)

# Combine metadata from TxDb + Gtf
meta <- cbind(elementMetadata(gns),
              elementMetadata(annoInfo)[idx,])
meta <- meta[,!duplicated(colnames(meta))]

elementMetadata(gns) <- meta
```

# Create annotation for transcript regions
```{r}
  # Define regions from TxDb
  # Count the overlaps of each binding site for each region of the transcript. 
  cdseq <- cds(annoDb) 
  intrns <- unlist(intronsByTranscript(annoDb), use.names=TRUE)  
  utrs3 <- unlist(threeUTRsByTranscript(annoDb), use.names=TRUE) 
  utrs5 <- unlist(fiveUTRsByTranscript(annoDb), use.names=TRUE) 
  regions_list <- GRangesList(UTR5 = utrs5, UTR3 = utrs3, CDS = cdseq, Intron = intrns)
  seqlevelsStyle(regions_list) <- "UCSC"
  
  # Extract all transcripts (as GRanges)
transcripts_id <- transcripts(annoDb, columns = c("tx_id", "tx_name", "gene_id"))
seqlevelsStyle(transcripts_id) <- "UCSC"
  #view(as.data.frame(regions_list))

saveRDS(regions_list, file = "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/mm39_region_list.rds")
```

# Create intron length table
```{r}
# transcript lookup
transcripts_id <- transcripts(annoDb, columns = c("tx_id", "tx_name", "gene_id"))
seqlevelsStyle(transcripts_id) <- "UCSC"

tx_df <- tibble(
  tx_id   = as.character(mcols(transcripts_id)$tx_id),
  tx_name = as.character(mcols(transcripts_id)$tx_name),
  gene_id = as.character(unlist(mcols(transcripts_id)$gene_id))) %>%
  distinct()

# recover tx_id from names(intrns)
intrns_df <- tibble::tibble(
  seqnames = as.character(seqnames(intrns)),
  start    = start(intrns),
  end      = end(intrns),
  strand   = as.character(strand(intrns)),
  width    = width(intrns),
  tx_id    = as.character(names(intrns))) %>%
  left_join(tx_df, by = "tx_id")

# Convert into Granges
introns_gr <- GRanges(
  seqnames = intrns_df$seqnames,
  ranges   = IRanges(start = intrns_df$start, end = intrns_df$end),
  strand   = intrns_df$strand,
  tx_id    = intrns_df$tx_id,
  tx_name  = intrns_df$tx_name,
  gene_id  = intrns_df$gene_id
)

saveRDS(introns_gr, file = "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/mm39_introns_gr.rds")
```

## Annotate strand for each txt GRanges
```{r}
# Merge TXT and BedGraph data
# We ensure site-specific scores are attached to genomic coordinates
gr_list <- lapply(names(txt_list), function(nm) {
  txt_gr <- txt_list[[nm]]
  bg_gr  <- bg_list[[nm]]
  
  # Find overlaps (identical coordinates)
  hits <- findOverlaps(txt_gr, bg_gr, type = "equal")
  res <- txt_gr[queryHits(hits)]
  res$score <- bg_gr$score[subjectHits(hits)]
  res$Ranges_coord <- bg_gr$NA..19[subjectHits(hits)]
  
  # Annotate strand from GTF
  strand_hits <- findOverlaps(res, gns, ignore.strand = TRUE)
  # Assign strand of the first gene hit (typical for HyperTRIBE)
  strand(res)[queryHits(strand_hits)] <- strand(gns)[subjectHits(strand_hits)]
  
  return(res)
})

# Keep same names
names(gr_list) <- names(txt_list)
# checking unique genomic ranges (chrom, start, end, strand)
#table(duplicated(granges(gr_list$Silke_DRRM_HT_1_A2G)))

saveRDS(gr_list, file = "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/GRanges_M3_HT_allsamples.rds")
```

# Strategy to identify true MATR3-dependent editing from background/noise events.
```{r}
save_thesis_plot <- function(plot, filename,
                             width = 10, height = 6.5,
                             dpi = 300, bg = "white") {
  ggsave(
    filename = filename,
    plot = plot,
    width = width,
    height = height,
    units = "in",
    dpi = dpi,
    bg = bg
  )
}


condition_map <- c(
  "Silke_M3_HT_1_A2G" = "MATR3", "Silke_M3_HT_2_A2G" = "MATR3", "Silke_M3_HT_3_A2G" = "MATR3", "Silke_M3_HT_4_A2G" = "MATR3",
  "Silke_M3d_HT_1_A2G" = "ΔRRMcontrol", "Silke_M3d_HT_2_A2G" = "ΔRRMcontrol", "Silke_M3d_HT_3_A2G" = "ΔRRMcontrol", "Silke_M3d_HT_4_A2G" = "ΔRRMcontrol",
  "Silke_HT_1_A2G" = "ADARcontrol", "Silke_HT_2_A2G" = "ADARcontrol", "Silke_HT_3_A2G" = "ADARcontrol", "Silke_HT_4_A2G" = "ADARcontrol"
)
```

## 1st GLOBAL CUTOFF: Globale filter
### Checking read distribution
Most genes have very low counts, but a few genes have very high counts (up to 16k).
That’s why filtering by a threshold is important—it removes a lot of weak/noisy signals at the bottom tail.
```{r}
#quantile(df_list$Silke_HT_1_A2G$Total_count, probs = seq(0,1, by = .1))

par(mfrow=c(2,2)) # Loop through all data frames

for (x in names(gr_list)) {
  
  df <- gr_list[[x]]
  probs <- seq(0, 100, 10) # percentiles for x-axis: 0,10,…100 
  
  q <- quantile(df$Total_count, probs = seq(0, 1, 0.1))
  
  plot(probs, q, type = "b", pch = 16, lwd = 2, col = "#0E54B6FF", # type="b": Plots both points and lines connecting them.
       main = paste("Quantile distribution of count:", x),
       xlab = "Percentile (%)", ylab = "read counts",
       ylim = c(0, max(q)))
  
  grid(col = "grey80", lty = 3)
  text(probs, q, labels = round(q, 2), pos = 3, cex = 1, col = "#BD241EFF")
  abline(h = q["20%"], col = "#BD241EFF", lty = 2, lwd = 2)
}
```


```{r, results: asis}
coverage_df <- imap_dfr(gr_list, ~{
  as.data.frame(.x) %>%
    mutate(Sample = .y)
}) %>%
  mutate(
    Condition = condition_map[Sample],
    Replicate = sub(".*_([1-4])_A2G$", "rep\\1", Sample),
    Condition = factor(Condition, levels = c("MATR3", "ΔRRMcontrol", "ADARcontrol")),
    Replicate = factor(Replicate, levels = c("rep1", "rep2", "rep3", "rep4")))

rep_colors <- setNames(
  colorRampPalette(c("#08306B", "#C6DBEF"))(4),
  c("rep1", "rep2", "rep3", "rep4"))

conditions <- unique(coverage_df$Condition)

for (cond in conditions) {
  
  cat("### ", cond, "\n\n", sep = "")
  
  p <- ggplot(
    coverage_df %>% filter(Condition == cond),
    aes(x = Total_count, color = Replicate)) +
    stat_ecdf(geom = "point", size = 0.8) +
    geom_vline(xintercept = c(5, 10, 15), linetype = "dashed", color = "grey50") +
    scale_color_manual(values = rep_colors, drop = FALSE) +
    theme_bw() +
    labs(
      title = cond,
      x = "Read coverage",
      y = "Cumulative fraction of editing sites",
      color = "Replicates") +
    coord_cartesian(xlim = c(0, 100)) +
    theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(color = "black", size = 17),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 15),
    legend.position = "right"
  )
  print(p)
  
  save_thesis_plot(
    p,
    paste0("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png/threshold_optimize/coverage_ecdf_", cond, ".png"),
    width = 8,
    height = 6
  )
  
  cat("\n\n")
}
```

### Filter editing sites with sufficient coverages
keep only sites with sufficient coverage (e.g. ≥ 5 reads = cut 20% dataset off)
```{r}
count_events <- function(gr_list, condition_map) {

  # create empty df   
count_events <- data.frame(
    Condition = character(),
    Total_unique_events = numeric(),
    stringsAsFactors = FALSE
  )

# for conditions in condition_map
for (cond in unique(condition_map)) {
  # Each name of grange in the list = 1 replicate 
    reps <- names(condition_map[condition_map == cond])
    
    # Extract data, convert to DF, and bind rows 
    combined_before <- dplyr::bind_rows(lapply(reps, function(rep) {
      as.data.frame(gr_list[[rep]])
    }))
    
    # Count unique sites (based on Ranges_coord)
    Total_unique_events <- dplyr::n_distinct(combined_before$Ranges_coord)
    
    # Store results
    count_events <- rbind(count_events, data.frame(
      Condition = cond,
      Total_unique_events = Total_unique_events
    ))
  }
  
  return(count_events)
}

count_events_df <- count_events(gr_list, condition_map)
```


### Loop for different cut-off levels for coverage (reads)
```{r}
cond_order <- c("MATR3", "ADARcontrol", "ΔRRMcontrol")
coverage_cutoff <- c(0, 5, 10, 15)

# Create lists to store filtered data and counts
coverage_cutoff_list <- list()
coverage_count_list  <- list()

for (k in coverage_cutoff) {
  
  # Filter each replicate based on coverage cutoff
  df_list_filtered <- lapply(names(gr_list), function(x) {
    gr <- gr_list[[x]]
    df <- as.data.frame(gr)
    
    if (k == 0) {
      df_filtered <- df
    } else {
      df_filtered <- df %>%
        dplyr::filter(Total_count >= k)
    }
    
    df_filtered %>%
      dplyr::select(Name, strand, Type, width, Ranges_coord, G_count, C_count, Total_count, score) %>%
      dplyr::distinct()
  })
  
  names(df_list_filtered) <- names(gr_list)
  coverage_cutoff_list[[as.character(k)]] <- df_list_filtered
  
  # Count total unique sites per condition
  counts <- count_events(df_list_filtered, condition_map)
  
  # Add stage label
  counts$Stage <- ifelse(
    k == 0,
    "No coverage cutoff",
    paste0("Coverage \u2265 ", k)
  )
  
  coverage_count_list[[as.character(k)]] <- counts
}

# Combine all results
coverage_count_list1 <- do.call(rbind, coverage_count_list)

# Order condition classes
coverage_count_list1$Condition <- factor(
  coverage_count_list1$Condition,
  levels = cond_order
)

# Order stages for plotting
coverage_count_list1$Stage <- factor(
  coverage_count_list1$Stage,
  levels = c(
    "No coverage cutoff",
    "Coverage \u2265 5",
    "Coverage \u2265 10",
    "Coverage \u2265 15"
  )
)

# Build plot object
coverage_optimize <- ggplot(
  coverage_count_list1,
  aes(x = Condition, y = Total_unique_events, fill = Stage)
) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.85) +
  geom_text(
    aes(label = scales::comma(Total_unique_events)),
    position = position_dodge(width = 0.9),
    vjust = 0,
    hjust = 0,
    size = 4.5,
    angle = 65
  ) +
  scale_fill_manual(values = c(
    "No coverage cutoff" = "#9E9E9E",
    "Coverage \u2265 5" = "#08306B",
    "Coverage \u2265 10" = "#6785AB",
    "Coverage \u2265 15" = "#96AFCB"
  )) +
  labs(
    title = "Editing sites across coverage cutoffs",
    x = "",
    y = "Number of editing sites",
    fill = "Coverage cutoff"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    legend.position = "right",
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(color = "black", size = 12, face = "bold"),
    axis.text.y = element_text(size = 12)
  ) +
  expand_limits(y = max(coverage_count_list1$Total_unique_events) * 1.3)

# Show plot
coverage_optimize

save_thesis_plot(
  coverage_optimize,
  filename = "/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png/threshold_optimize/coverage_cutoff_optimize.png",
  width = 8.5,
  height = 6
)
```

## 2nd GLOBAL CUTOFF: How strong are the sites we kept
How is score (editing score) per site calculated from bigWig coverage? 
score = (G/C count depend on +/- strand)/ Total count (coverage at that genomic position) * 100
## Editing score bins by condition {.tabset}
```{r}
## Editing score bins by condition {.tabset}
#| results: asis

# Collect events after 1st cutoff
df_list_tc5 <- coverage_cutoff_list[["5"]]

score_df <- imap_dfr(df_list_tc5, ~{
  .x %>%
    mutate(Sample = .y)
}) %>%
  mutate(
    Condition = condition_map[Sample],
    Replicate = sub(".*_([1-4])_A2G$", "rep\\1", Sample),
    Condition = factor(Condition, levels = c("MATR3", "ΔRRMcontrol", "ADARcontrol")),
    Replicate = factor(Replicate, levels = c("rep1", "rep2", "rep3", "rep4"))
  )

conditions <- levels(score_df$Condition)

for (cond in conditions) {
  
  cat("### ", cond, "\n\n", sep = "")
  
  p <- ggplot(
  score_df %>% filter(Condition == cond),
  aes(x = score, color = Replicate)
) +
  stat_ecdf(geom = "point", size = 0.8, alpha = 0.6) +
  geom_vline(xintercept = c(1, 5, 10), linetype = "dashed", color = "grey50") +
  scale_color_manual(values = rep_colors, drop = FALSE) +
  scale_x_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 10)
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  labs(
    title = cond,
    x = "Editing score (%)",
    y = "Cumulative fraction of sites",
    color = "Replicates"
  ) +
  theme_bw(base_size =20) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(color = "black", size = 17),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 15),
    legend.position = "right"
  )
  
  print(p)
  
  save_thesis_plot(
    p,
    paste0("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png/threshold_optimize/editscore_dis", gsub("[^A-Za-z0-9]+", "_", cond), ".png"),
    width = 8,
    height = 6
  )
  
  cat("\n\n")
}
```

### Loop for different cut-off levels
```{r}
edit_cutoff <- c(0, 1, 3, 5, 7, 9)

# Create lists to store filtered data and counts
edit_cutoff_list <- list()
edit_count_list  <- list()

for (k in edit_cutoff) {
  
  # Filter each replicate based on editing score cutoff
  df_list_filtered <- lapply(names(df_list_tc5), function(x) {
    gr <- df_list_tc5[[x]]
    df <- as.data.frame(gr)
    
    if (k == 0) {
      df_filtered <- df
    } else {
      df_filtered <- df %>%
        dplyr::filter(score >= k)
    }
    
    df_filtered %>%
      dplyr::distinct()
  })
  
  # use the correct names
  names(df_list_filtered) <- names(df_list_tc5)
  
  # store filtered replicate-level data
  edit_cutoff_list[[as.character(k)]] <- df_list_filtered
  
  # count total unique events per condition
  counts <- count_events(df_list_filtered, condition_map)
  
  # add stage label
  counts$Stage <- ifelse(
    k == 0,
    "No editing score cutoff",
    paste0("Score \u2265 ", k, "%")
  )
  
  # store summary counts
  edit_count_list[[as.character(k)]] <- counts
}

# Combine all count tables
edit_count_list1 <- do.call(rbind, edit_count_list)

# Set factor order for conditions
edit_count_list1$Condition <- factor(
  edit_count_list1$Condition,
  levels = cond_order
)

# Set factor order for stages
edit_count_list1$Stage <- factor(
  edit_count_list1$Stage,
  levels = c(
    "No editing score cutoff",
    "Score \u2265 1%",
    "Score \u2265 3%",
    "Score \u2265 5%",
    "Score \u2265 7%",
    "Score \u2265 9%"
  )
)

# Plot
editscore_cutoff <- ggplot(
  edit_count_list1,
  aes(x = Condition, y = Total_unique_events, fill = Stage)
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = scales::comma(Total_unique_events)),
    position = position_dodge(width = 0.9),
    vjust = 0,
    hjust = 0,
    size = 4.5,
    angle = 65
  ) +
  scale_fill_manual(values = c(
    "No editing score cutoff" = "#9E9E9E",
    "Score \u2265 1%" = "#08306B",
    "Score \u2265 3%" = "#375A8B",
    "Score \u2265 5%" = "#6785AB",
    "Score \u2265 7%" = "#96AFCB",
    "Score \u2265 9%" = "#C6DBEF"
  )) +
  theme_bw(base_size = 14) +
  labs(
    title = "Editing sites after coverage \u2265 5 filtering across editing score cutoffs",
    x = "",
    y = "Number of editing sites",
    fill = "Cutoffs"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    legend.position = "right",
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(color = "black", size = 12, face = "bold"),
    axis.text.y = element_text(size = 12)
  ) +
  expand_limits(y = max(edit_count_list1$Total_unique_events) * 1.15)

save_thesis_plot(
  editscore_cutoff,
  filename = "/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png/threshold_optimize/editscore_cutoff_optimize.png",
  width = 8.5,
  height = 6
)

df_list_tc5_e5 <- edit_cutoff_list[["5"]]
df_list_tc5_e1 <- edit_cutoff_list[["1"]]

editscore_cutoff
```

# Mean editing score of reproducible events
I calculate this using rowSums(!is.na(...)). This counts how many replicates actually detected the site. This becomes our 'Reproducibility Score'—if a site has a frequency of 1, it’s a singleton/noise. If it’s 4, it’s highly robust.

Mean Score: We calculate the average editing percentage across the active replicates. This gives us a single representative editing value for that genomic location

## ADAR control
```{r}
ADAR_ctrl_list <- df_list_tc5_e5[grep("Silke_HT_", names(df_list_tc5_e5))]

ADARctrl_scores <- purrr::imap(ADAR_ctrl_list, ~{
  .x %>%
    dplyr::select(Ranges_coord, score) %>%
    group_by(Ranges_coord) %>%
    summarise(score = mean(score, na.rm = TRUE), .groups = "drop") %>%
    dplyr::mutate(Condition = paste0("ADARcontrol_rep", gsub("Silke_HT_|_A2G", "", .y)))
}) %>%
  bind_rows()

# Pivot to wider format (one column per replicate)
ADARctrl_scores_count <- ADARctrl_scores %>%
  pivot_wider(
    id_cols = Ranges_coord,
    names_from = Condition,
    values_from = score) %>%
  mutate(Frequency = rowSums(!is.na(across(starts_with("ADARcontrol_rep"))))) %>%
  mutate(Mean = rowMeans(across(starts_with("ADARcontrol_rep")), na.rm = TRUE))
```

## MATR3 WT
```{r}
MATR3_list <- df_list_tc5_e5[grep("Silke_M3_HT_", names(df_list_tc5_e5))]

MATR3_wt_scores <- purrr::imap(MATR3_list, ~{
  .x %>%
    dplyr::select(Ranges_coord, score) %>%
    group_by(Ranges_coord) %>%
    summarise(score = mean(score, na.rm = TRUE), .groups = "drop") %>%
    mutate(Condition = paste0("Matr3_wt_rep", gsub("Silke_M3_HT_|_A2G", "", .y)))
}) %>%
  bind_rows()

# Pivot to wider format (one column per replicate)
MATR3_wt_scores_count <- MATR3_wt_scores %>%
  pivot_wider(
    id_cols = Ranges_coord,
    names_from = Condition,
    values_from = score) %>%
  mutate(Frequency = rowSums(!is.na(across(starts_with("Matr3_wt_rep"))))) %>%
  mutate(Mean = rowMeans(across(starts_with("Matr3_wt_rep")), na.rm = TRUE))

```

## Deplete RRM
```{r}
ΔRRM_list <- df_list_tc5_e5[grep("Silke_M3d_HT_", names(df_list_tc5_e5))]

ΔRRM_scores <- purrr::imap(ΔRRM_list, ~{
  .x %>%
    dplyr::select(Ranges_coord, score) %>%
    group_by(Ranges_coord) %>%
    summarise(score = mean(score, na.rm = TRUE), .groups = "drop") %>%
    mutate(Condition = paste0("ΔRRM_rep", gsub("Silke_M3d_HT_|_A2G", "", .y)))
}) %>%
  bind_rows()

# Pivot to wider format (one column per replicate)
ΔRRM_scores_count <- ΔRRM_scores %>%
  pivot_wider(
    id_cols = Ranges_coord,
    names_from = Condition,
    values_from = score) %>%
  mutate(Frequency = rowSums(!is.na(across(starts_with("ΔRRM_rep"))))) %>%
  mutate(Mean = rowMeans(across(starts_with("ΔRRM_rep")), na.rm = TRUE))
```

## How many events reproduce in more than 2 replicates?
```{r}
# Custom function to generate high-quality UpSet plots with color-coded reproducibility
plot_reproducible_upset <- function(upset_list, filename, main_label, highlight_color = "#2171B5") {
  
  # 1. Transform list to UpSet format
  upset_data <- UpSetR::fromList(upset_list)
  set_names <- names(upset_list)
  
  # 2. Generate all combinations for intersections (degree >= 2)
  shared_combos <- unlist(lapply(2:length(set_names), function(x) {
    combn(set_names, x, simplify = FALSE)
  }), recursive = FALSE)
  
  # 3. Create queries for UpSetR
  highlight_queries <- lapply(shared_combos, function(combo) {
    list(query = UpSetR::intersects, params = combo, color = highlight_color, active = TRUE)
  })
  
  # 4. Render and Save Plot
  png(filename = filename, width = 2600, height = 2200, res = 300, bg = "white")
  
  # Explicit call to UpSetR::upset to prevent namespace errors
  p <- UpSetR::upset(
    upset_data,
    nsets = length(set_names),
    order.by = "freq",
    mainbar.y.label = main_label,
    sets.x.label = "Editing sites / Replicate",
    sets.bar.color = "#08306B",
    main.bar.color = "#08306B",   # Color for singletons (degree = 1)
    queries = highlight_queries,
    text.scale = c(2.0, 1.6, 1.6, 1.4, 2.0, 2.2),
    number.angles = 60
  )
  p$Main_bar$grobs[[6]]$children[[4]]$gp$col <- rep("black", 3)
  p$Main_bar$grobs[[6]]$children[[4]]$hjust <- rep(0, 3)
  
  print(p)
  dev.off()

}

```

### Control (HT)
```{r}
ADARctrl_upsetlist <- list(
  ADARctrl_r1 = ADAR_ctrl_list$Silke_HT_1_A2G$Ranges_coord,
  ADARctrl_r2 = ADAR_ctrl_list$Silke_HT_2_A2G$Ranges_coord,
  ADARctrl_r3 = ADAR_ctrl_list$Silke_HT_3_A2G$Ranges_coord,
  ADARctrl_r4 = ADAR_ctrl_list$Silke_HT_4_A2G$Ranges_coord
)

plot_reproducible_upset(ADARctrl_upsetlist, "/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png/threshold_optimize/ADAR_HT_upset.png", "Shared editing sites (HT)")
```


## MATR3 WT
```{r}
MATR3_upsetlist <- list(
        MATR3_r1 = unique(MATR3_list$Silke_M3_HT_1_A2G$Ranges_coord),
        MATR3_r2 = unique(MATR3_list$Silke_M3_HT_2_A2G$Ranges_coord),
        MATR3_r3 = unique(MATR3_list$Silke_M3_HT_3_A2G$Ranges_coord),
        MATR3_r4 = unique(MATR3_list$Silke_M3_HT_4_A2G$Ranges_coord))

plot_reproducible_upset(MATR3_upsetlist, "/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png/threshold_optimize/MATR3_HT_upset.png", "Shared editing sites (M3_HT)")
```

### Depleted RRM (M3d_HT)
```{r}
ΔRRM_upsetlist <- list(ΔRRM_r1 = unique(ΔRRM_list$Silke_M3d_HT_1_A2G$Ranges_coord),
                          DRRM_r2 = unique(ΔRRM_list$Silke_M3d_HT_2_A2G$Ranges_coord),
                          ΔRRM_r3 = unique(ΔRRM_list$Silke_M3d_HT_3_A2G$Ranges_coord),
                          ΔRRM_r4 = unique(ΔRRM_list$Silke_M3d_HT_4_A2G$Ranges_coord))

plot_reproducible_upset(ΔRRM_upsetlist, "/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png/threshold_optimize/ΔRRM_HT_upset.png", "Shared editing sites (M3d_HT)")
```

```{r}
reproducible_events <- function(df) {
  
  df_reproducible <- df[df$Frequency >= 2,]
  
  return(df_reproducible)}

ADARctrl_reproducible <- reproducible_events(ADARctrl_scores_count) 
MATR3_wt_reproducible <- reproducible_events(MATR3_wt_scores_count)
ΔRRM_reproducible <- reproducible_events(ΔRRM_scores_count)

reprod_list <- list(
  ADARctrl = ADARctrl_reproducible,
  MATR3 = MATR3_wt_reproducible,
  ΔRRMctrl = ΔRRM_reproducible
)
```

# MATR3 with editing % < 5 reproducibility
```{r}
# use coverage-filtered object before edit cutoff
MATR3_c5_list <- df_list_tc5[grep("Silke_M3_HT_", names(df_list_tc5))]

# Keep unique genomic sites per replicate, then split by score
MATR3_c5_list_e4 <- purrr::imap(MATR3_c5_list, ~{
  as.data.frame(.x) %>%
    dplyr::select(Ranges_coord, score) %>%
    group_by(Ranges_coord) %>%
    summarise(score = mean(score, na.rm = TRUE), .groups = "drop") %>%
    filter(score < 5) %>%
    mutate(rep = paste0("rep", gsub("Silke_M3_HT_|_A2G", "", .y)))
})

MATR3_c5_e4_upsetlist <- list(
        MATR3_r1 = unique(MATR3_c5_list_e4$Silke_M3_HT_1_A2G$Ranges_coord),
        MATR3_r2 = unique(MATR3_c5_list_e4$Silke_M3_HT_2_A2G$Ranges_coord),
        MATR3_r3 = unique(MATR3_c5_list_e4$Silke_M3_HT_3_A2G$Ranges_coord),
        MATR3_r4 = unique(MATR3_c5_list_e4$Silke_M3_HT_4_A2G$Ranges_coord))

# Prepare list for UpSetR
MATR3_c5_e4_upset <- fromList(MATR3_c5_e4_upsetlist)

#WT_upset_filtered <- WT_upset[rowSums(WT_upset) >= 2, ]

png(
  filename = "MATR3_c5_e4_upset.png",
  width = 2400,
  height = 1800,
  res = 300,
  bg = "white")

upset(
  MATR3_c5_e4_upset,
  nsets = length(MATR3_c5_e4_upsetlist), 
  order.by = "freq",
  mainbar.y.label = "Shared editing sites (MATR3_HT)",
  sets.x.label = "Editing sites/  Replicate",
  sets.bar.color = "#08306B",
  main.bar.color = "#6785AB",
  text.scale = c(1.2, 1.2, 1, 1, 1, 1)
)

dev.off()
```


# Count events after each cut-off 
```{r}
count_events <- function(gr_list, condition_map) {
  
count_events <- data.frame(
    Condition = character(),
    Total_unique_events = numeric(),
    stringsAsFactors = FALSE
  )

for (cond in unique(condition_map)) {
    reps <- names(condition_map[condition_map == cond])
    combined_before <- data.frame()
    
    for (rep in reps) {
      gr <- gr_list[[rep]]
      df <- as.data.frame(gr)
      combined_before <- dplyr::bind_rows(combined_before, df)
    }
    
    # Count unique sites (based on Ranges_coord)
    Total_unique_events <- dplyr::n_distinct(combined_before$Ranges_coord)
    
    # Store results
    count_events <- rbind(count_events, data.frame(
      Condition = cond,
      Total_unique_events = Total_unique_events
    ))
  }
  
  return(count_events)
}

reprod_conds <- c("ADARctrl" = "ADARcontrol",
                  "MATR3" = "MATR3",
                  "ΔRRMctrl" = "ΔRRMcontrol")
                  
total_counts <- count_events(gr_list, condition_map)

# calculate all events after 1st cutoff
counts_after_tc5 <- count_events(df_list_tc5, condition_map)

# editing cutoffs ≥5%
counts_after_e5 <- count_events(df_list_tc5_e5, condition_map)

# reprod cutoff ≥2 replicates 
counts_after_r2 <- count_events(reprod_list, reprod_conds)

HT_colors <- setNames(
  c("#BD241EFF", "#F5B355", "cornflowerblue"),
  c("ADARcontrol", "MATR3", "ΔRRMcontrol")
)
```

```{r}
total_counts <- count_events(gr_list, condition_map)

# calculate all events after 1st cutoff
counts_after_tc5 <- count_events(df_list_tc5, condition_map)

# editing cutoffs ≥5%
counts_after_e5 <- count_events(df_list_tc5_e5, condition_map)

# reprod cutoff ≥2 replicates 
counts_after_r2 <- count_events(reprod_list, reprod_conds)
```

```{r}
barplot_count <- bind_rows(
  total_counts     %>% mutate(Stage = "No cutoff"),
  counts_after_tc5 %>% mutate(Stage = "Coverage ≥ 5"),
  counts_after_e5  %>% mutate(Stage = "Score ≥ 5%"),
  counts_after_r2  %>% mutate(Stage = "≥ 2 replicates")
)

barplot_count$Stage <- factor(
  barplot_count$Stage,
  levels = c(
    "No cutoff",
    "Coverage ≥ 5",
    "Score ≥ 5%",
    "≥ 2 replicates"
  )
)

barplot_count$Condition <- factor(
  barplot_count$Condition,
  levels = cond_order
)

compare_cutoff <- ggplot(
  barplot_count,
  aes(x = Condition, y = Total_unique_events, fill = Stage)
) +
  geom_bar(
    stat = "identity",
    position = "dodge",
    width = 0.8
  ) +
  geom_text(
    aes(label = scales::comma(Total_unique_events)),
    position = position_dodge(width = 0.9),
    vjust = 0,
    hjust = 0,
    size = 5.5,
    angle = 65
  ) +
  scale_fill_manual(values = c(
    "No cutoff" = "#9E9E9E",
    "Coverage ≥ 5" = "#08306B",
    "Score ≥ 5%" = "#6785AB",
    "≥ 2 replicates" = "#96AFCB"
  )) +
  scale_y_continuous(
    limits = c(0, max(barplot_count$Total_unique_events) * 1.4)
  ) +
  theme_bw(base_size = 15) +
  labs(
    title = "Editing sites retained across sequential filtering",
    x = "",
    y = "Number of editing sites",
    fill = "Filtering step"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(color = "black", size = 14, face = "bold"),
    axis.text.y = element_text(size = 14),
    legend.position = "right",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14)
  )

save_thesis_plot(
  compare_cutoff,
  filename = "cutoff_comparison_barplot.png",
  width = 8.5,
  height = 6
)

compare_cutoff
```

# Remove sites look alike background noise to collect high confident sites
The negative control (no MATR3) captures non-specific signal (background): mapping artifacts, sequencing errors, contaminating gDNA, or editing triggered by Hypertribe. If a site appears robustly in the negative control, it’s likely not specifically induced by MATR3. Therefore, removing negative-control and overlapping among negative control and WT and dRRM sites reduces false positives.
```{r}
ggven_reproducible_events <- list(
  ADARcontrol = ADARctrl_reproducible$Ranges_coord, 
  MATR3_wt = MATR3_wt_reproducible$Ranges_coord,
  ΔRRMcontrol = ΔRRM_reproducible$Ranges_coord)

png(
  filename = "/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png/threshold_optimize/reproducible-venndiagram.png",
  width = 2400,
  height = 1800,
  res = 300,
  bg = "white")

plot(eulerr::euler(ggven_reproducible_events),
       fills = list(fill = c("#BD241EFF", "#2171B5", "#F5B355"), alpha = 0.5),
       quantities = TRUE,
       stroke_size = 1, # control border size
       set_name_size = 1, # set name size
       text_size = 2)

dev.off()
```


# Collect high confident events
Why do we collect DRRM events?

If a site shows up in both WT and dRRM, but not in the negative control, it means:
Editing at that site depends on MATR3 being present (because it doesn’t happen in negative control without MATR3).
But it does not depend on the RRM domain specifically, since even without RRM (dRRM mutant), the site is still edited.
This suggests MATR3 might recruit ADAR to that RNA site through other parts of the protein (like zinc fingers, low-complexity domains, or protein–protein interactions) rather than through its RRM domain.
```{r}
custom_colors <- c(
  "ADARcontrol" = "grey",      
  "MATR3_wt" = "#2171B5",           # highlight WT unique
  "ΔRRMcontrol" = "grey",         
  "ADARcontrol&MATR3_wt" = "grey",
  "ADARcontrol&ΔRRMcontrol" = "grey",
  "MATR3_wt&ΔRRMcontrol" = "grey", 
  "ADARcontrol&MATR3_wt&ΔRRMcontrol" = "grey"
)

png(
  filename = "/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png/threshold_optimize/Matr3-specific-euler.png",
  width    = 8,
  height   = 5,
  units    = "in",
  res      = 600,
  bg       = "white"
)

plot(
  eulerr::euler(ggven_reproducible_events),
  
  fills = list(
    fill  = custom_colors,
    alpha = 0.5
  ),
  
  # ── Label sizes ────────────────────────────────────────────────
  labels = list(
    fontsize = 16          # set name labels (e.g. "MATR3", "ADAR")
  ),
  
  quantities = list(
    fontsize = 16,         # ← increase this for overlap numbers
    fontface = "bold"      # optional: make numbers bold
  ),
  
  stroke_size = 1
)

dev.off()

```

```{r}
# Get overlap among all 3 datasets
overlap_upon_all <- Reduce(intersect, list(ADARctrl_reproducible$Ranges_coord, 
                                    MATR3_wt_reproducible$Ranges_coord,
                                    ΔRRM_reproducible$Ranges_coord))

# Remove events belonging to Negative control 
ADAR_MATR3_overlap <- intersect(ADARctrl_reproducible$Ranges_coord,
                                  MATR3_wt_reproducible$Ranges_coord)

ADAR_ΔRRM_overlap <- intersect(ADARctrl_reproducible$Ranges_coord,
                                   ΔRRM_reproducible$Ranges_coord)

```

```{r}
################################ Overlap upon DRRM and WT
MATR3_ΔRRM_overlap_only <- setdiff(
  intersect(MATR3_wt_reproducible$Ranges_coord,
            ΔRRM_reproducible$Ranges_coord),
            ADARctrl_reproducible$Ranges_coord)

reprod_MATR3_ΔRRM_df <-  MATR3_wt_reproducible[MATR3_wt_reproducible$Ranges_coord %in% MATR3_ΔRRM_overlap_only, ]

################################ DRRM unique
reprod_ΔRRM <- setdiff(ΔRRM_reproducible$Ranges_coord, union(ADAR_ΔRRM_overlap, MATR3_ΔRRM_overlap_only))

reprod_uniq_ΔRRM_df <- ΔRRM_reproducible[ΔRRM_reproducible$Ranges_coord %in% reprod_ΔRRM, ]

################################ WT unique
reprod_MATR3 <- setdiff(MATR3_wt_reproducible$Ranges_coord, union(ADAR_MATR3_overlap, MATR3_ΔRRM_overlap_only)) 

reprod_uniq_MATR3_df <- MATR3_wt_reproducible[MATR3_wt_reproducible$Ranges_coord %in% reprod_MATR3, ]

################################ Negative control unique
reprod_ADARctrl <- setdiff(ADARctrl_reproducible$Ranges_coord, union(ADAR_MATR3_overlap, ADAR_ΔRRM_overlap))

reprod_uniq_ADAR_df <- ADARctrl_reproducible[ADARctrl_reproducible$Ranges_coord %in% reprod_ADARctrl, ]
```


## Convert high confident site df into GRanges
```{r}
df_convert_gr <- function(df) {
  df2 <- df %>%
 # Split Ranges_coord into chr and numeric position
  tidyr::separate(Ranges_coord, into = c("seqnames", "ranges"), sep = "_") %>%
  mutate(
    ranges = as.numeric(ranges),
    score = round(Mean, 2)) %>%
  dplyr::select(seqnames, ranges, score)

  gr <- makeGRangesFromDataFrame(
    df2,
    seqnames.field = "seqnames",
    start.field    = "ranges",
    end.field      = "ranges",
    keep.extra.columns = TRUE)
  return(gr)}
############################################ All events 
ADARctrl_reproducible_gr <- df_convert_gr(ADARctrl_reproducible)
MATR3_wt_reproducible_gr <- df_convert_gr(MATR3_wt_reproducible)
ΔRRM_reproducible_gr <- df_convert_gr(ΔRRM_reproducible)

############################################ Reprod events
reprod_uniq_ΔRRM_gr <- df_convert_gr(reprod_uniq_ΔRRM_df)
reprod_uniq_MATR3_gr <- df_convert_gr(reprod_uniq_MATR3_df)
reprod_uniq_ADAR_gr <- df_convert_gr(reprod_uniq_ADAR_df)
reprod_MATR3_ΔRRM_gr <- df_convert_gr(reprod_MATR3_ΔRRM_df)

# Combine all reproducible GRanges
m3specific_sites <- reprod_uniq_MATR3_gr
```

# Which genes that reproducible events occur? 
```{r}
annotate_genenames <- function(gr, gns) {
  
  hits <- GenomicRanges::findOverlaps(gr, gns, type = "within", ignore.strand = FALSE)
  gr2 <- gr[queryHits(hits)]
  strand(gr2) <- strand(gns)[subjectHits(hits)]
  
  available_cols <- intersect(c("gene_name", "gene_id", "gene_type", "tag"), colnames(mcols(gns)))
  
  mcols(gr2) <- cbind(
    mcols(gr2),
    mcols(gns)[subjectHits(hits), available_cols, drop = FALSE]
  )
  
  as.data.frame(gr2) %>%
    dplyr::select(any_of(c("gene_name", "gene_id", "seqnames", "start", "end", 
                           "strand", "score", "gene_type", "tag", "region"))) %>%
    dplyr::distinct()
}

############## high confident sites
m3specific_sites2 <- annotate_genenames(m3specific_sites, gns)
#high_conf_sites2 <- high_conf_sites2 %>%
#  mutate(GeneID = sub("\\..*", "", gene_id))
saveRDS(m3specific_sites2, "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/m3specific_sites2.rds")
```

# Genes of interest
```{r}
# Gene of interest
#mmusculus <- useMart("ENSEMBL_MART_ENSEMBL", dataset="mmusculus_gene_ensembl")

#listFilters(bm)
#listAttributes(bm)

genes_of_interest <- c(
  "Rbm20","Matr3","Xist","Ttn","Cacna1a","Cacna1b","Cacna1c","Cacna1d",
  "Ldb3","Pdlim5","Camk2a","Camk2d","Camk2g","Rbm14","Rbm24","Rbm39",
  "Rbm8a","Rbpms","Rbpms2","Mbnl1","Mbnl2","Nppa","Nppb","Myh6","Myh7",
  "Hnrnpu","Srsf3","Tnnt2","Des","Acta1","Actn2","Mybpc1","Mybpc3","Myocd"
)

collagen_genes <- c(
  "Col1a1","Col1a2","Col2a1","Col3a1","Col4a1","Col4a2","Col4a3","Col4a4",
  "Col4a5","Col4a6","Col5a1","Col5a2","Col5a3","Col6a1","Col6a2","Col6a3",
  "Col6a4","Col6a5","Col6a6","Col7a1","Col8a1","Col8a2","Col9a1","Col9a2",
  "Col9a3","Col10a1","Col11a1","Col11a2","Col12a1","Col13a1","Col14a1",
  "Col15a1","Col16a1","Col17a1","Col18a1","Col19a1","Col20a1","Col22a1",
  "Col23a1","Col24a1","Col25a1","Col26a1","Col27a1","Col28a1","Colgalt1",
  "Colgalt2","Colq"
)
```


## Count events per region per dataset
```{r}
region_anno_events <- function(gr, regions_list) {
  
  # initialize metadata
  region    <- rep(NA_character_, length(gr))
  region_id <- rep(NA_character_, length(gr))
  
  # loop over regions in GRangesList
  for (x in names(regions_list)) {
    hits <- GenomicRanges::findOverlaps(gr, regions_list[[x]], type="any", ignore.strand=FALSE)
    
    region[queryHits(hits)[is.na(region[queryHits(hits)])]] <- x
  }
  
  # add metadata back to gr
  mcols(gr)$region    <- region
  mcols(gr)$region_id <- region_id
  
  return(gr)
}

################################ high confident sites
m3specific_sites2_gr <- makeGRangesFromDataFrame(m3specific_sites2,
    seqnames.field = "seqnames",
    start.field    = "start",
    end.field      = "end",
    keep.extra.columns = TRUE)
m3specific_sites2_gr_anno    <- region_anno_events(m3specific_sites2_gr, regions_list)
gns <- region_anno_events(gns, regions_list)
saveRDS(gns, file = "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/gns_mm39.rds")
#as.data.frame(high_conf_sites2_gr_anno)

gene_anno <- as.data.frame(gns) %>%
  dplyr::select(gene_id = gene_id, gene_name = gene_name, gene_type = gene_type) %>%
  distinct() %>% 
  mutate(GeneID = sub("\\..*", "", gene_id)) # remove version suffix from gene_id
saveRDS(gene_anno, file = "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/geneanno_df_mm39.rds")
```

# Export into bedGraph format for IGV visualization
```{r}
seqlevelsStyle(m3specific_sites2_gr_anno) <- "UCSC"

m3specific_sites2_gr_anno_bg <- m3specific_sites2_gr_anno
start(m3specific_sites2_gr_anno_bg) <- start(m3specific_sites2_gr_anno_bg) - 1
end(m3specific_sites2_gr_anno_bg)   <- start(m3specific_sites2_gr_anno_bg)       # 1bp

export(m3specific_sites2_gr_anno_bg, con = "/Users/tinkienoob/Master_thesis/MATR3/m3specific_sites.bedGraph", format = "bedGraph")
```


# Create list of control 
```{r}
all_control_list <- list(
  only_ADARctrl = reprod_uniq_ADAR_df, # Background
  only_ΔRRMctrl = reprod_uniq_ΔRRM_df # Different enrichment, different specificity
)

############## control annotation function
control_annotate_func <- function(df, gns, regions_list) {

  gr <- df_convert_gr(df)                 
  df_gene <- annotate_genenames(gr, gns)  # returns a DF

  gr2 <- makeGRangesFromDataFrame(df_gene,
    seqnames.field="seqnames", start.field="start", end.field="end",
    keep.extra.columns=TRUE
  )
  GenomeInfoDb::seqlevelsStyle(gr2) <- "UCSC"

  gr2 <- region_anno_events(gr2, regions_list)

  as.data.frame(gr2) %>%
    mutate(GeneID = sub("\\..*", "", gene_id))
}

control_anno_gr <- lapply(all_control_list,
                          control_annotate_func,
                          gns = gns,
                          regions_list = regions_list)

saveRDS(control_anno_gr, file = "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/HT_control_list.rds")
```

# Number of high confident target genes 
```{r}
# Count number of high confident editing sites of genes
m3specific_sites_per_genes <- as.data.frame(m3specific_sites2_gr_anno) %>%
  group_by(gene_id, gene_name) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(desc(count)) %>%
  mutate(
    Target_genes = case_when(
      gene_name %in% genes_of_interest ~ "GOI",
      gene_name %in% collagen_genes ~ "Collagen",
      TRUE ~ "Other")) 
    
# Get TRUE or FALSE if genes are GOI
m3specific_sites_per_genes2 <- tibble(
  Target_genes = c("GOI", "Collagen"),
  Total_target = c(length(genes_of_interest), length(collagen_genes)),
  highconf_target = c(
    sum(m3specific_sites_per_genes$gene_name %in% genes_of_interest),
    sum(m3specific_sites_per_genes$gene_name %in% collagen_genes)
  )
)

m3specific_sites_per_genes2_long <- m3specific_sites_per_genes2 %>%
  mutate(
    not_included = Total_target - highconf_target
  ) %>%
  pivot_longer(cols = c(highconf_target, not_included),
               names_to = "Status",
               values_to = "Count") %>%
  mutate(Status = recode(Status, 
                         "highconf_target" = "high confident", 
                         "not_included" = "not included"),
         Percent = Count / Total_target * 100)

ggplot(m3specific_sites_per_genes2_long, aes(x = Target_genes, y = Percent, fill = Status)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(Percent, 1), "%")), 
            position = position_stack(vjust = 0.5), size = 5) +
  labs(
    title = "Percentage of Target Genes Detected in High-confidence Sites for MATR3",
    x = "Target gene set",
    y = "Percentage of genes"
  ) +
  scale_fill_manual(values = c("high confident" = "#F5B355", "not included" = "grey")) +
  theme_minimal(base_size = 10)
```


## Region distribution of Matrin3-specific sites vs. control
```{r}
region_colors <- c(
  "UTR5"   = "#3288BD", 
  "CDS"    = "#99D594", 
  "Intron" = "#FDAE61", 
  "UTR3"   = "#D53E4F", 
  "Others"  = "#E0E0E0"  # Neutral grey
)
# set class order
class_order <- c("M3-specific", "ADAR specific", "ΔRRM specific")
# set region order manually if you want full control
# change this to match your real annotation names exactly
region_order <- c("UTR5", "CDS", "UTR3", "exon", "Intron", "Others")

# if your region names are different, for example:
# region_order <- c("five_prime_utr", "cds", "three_prime_utr", "exon", "intron", "Others")

# count number of each region in M3-specific sites
m3specific_tx_region_counts <- as.data.frame(m3specific_site_gr_anno) %>%
  mutate(region = ifelse(is.na(region), "Others", region)) %>%
  group_by(region) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(
    percentage = count / sum(count) * 100,
    class = "M3-specific")

# count number of each region in controls
all_control_region_counts_list <- setNames(
  lapply(names(all_control_list_HT), function(cond) {
    df <- all_control_list_HT[[cond]]
    
    df %>%
      mutate(region = ifelse(is.na(region) | region == "", "Others", region)) %>%
      group_by(region) %>%
      summarise(count = n(), .groups = "drop") %>%
      mutate(
        percentage = count / sum(count) * 100,
        class = cond
      )
  }),
  names(all_control_list_HT))

all_control_region_counts <- bind_rows(all_control_region_counts_list) %>%
  mutate(class = recode(
    class,
    "only_ADARctrl" = "ADAR specific",
    "only_ΔRRMctrl" = "ΔRRM specific"
  ))

# combine
regions_count_data <- bind_rows(
  m3specific_tx_region_counts,
  all_control_region_counts)

# make sure missing regions still appear if absent in some classes
regions_count_data <- regions_count_data %>%
  mutate(
    class = factor(class, levels = class_order),
    region = ifelse(region %in% region_order, region, "Others"),
    region = factor(region, levels = region_order)
  ) %>%
  arrange(class, region)
```


```{r}
##################### Stacked barplot
region_stacked <- ggplot(regions_count_data, aes(x = class, y = percentage, fill = region)) +
  geom_col(color = "black", linewidth = 0.25, width = 0.6) +
  geom_text(
    aes(label = ifelse(percentage >= 3, paste0(round(percentage, 1), "%"), "")),
    position = position_stack(vjust = 0.5),
    size = 5.5,
    fontface = "bold") +
  scale_fill_manual(values = region_colors, drop = FALSE) +
  labs(
    title = "Genomic region distribution",
    x = NULL, y = "Proportion of editing sites", 
    fill = "Region") +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    legend.position = "right",
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(color = "black", size = 14, face = "bold"),
    axis.text.y = element_text(size = 14)
  )

##################### Group barplot
region_group <- ggplot(regions_count_data, aes(x = class, y = count, fill = region)) +
  geom_col(
    color = "black", linewidth = 0.25, width = 0.75,
    position = position_dodge2(width = 0.8, preserve = "single")) +
  geom_text(
    aes(label = count),
    position = position_dodge2(width = 0.8, preserve = "single"),
    vjust = -0.4,
    size = 3) +
  scale_fill_manual(values = region_colors, drop = FALSE) +
  labs(
    title = "Genomic region distribution",
    x = NULL, y = "Number of editing sites", fill = "Region") +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(face = "bold", size = 12),
    legend.position = "right") 

# save_thesis_plot(
#   region_stacked,
#   filename = file.path("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
#                        "Genomic_region_distribution_stacked.png"),
#   width = 7.5,
#   height = 6,
#   dpi = 300,
#   bg = "white")
# 
# save_thesis_plot(
#   region_group,
#   filename = file.path("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
#                        "Genomic_region_distribution_group.png"),
#   width = 7.5,
#   height = 6,
#   dpi = 300,
#   bg = "white")

region_stacked
region_group
```

To answer the question of whether the genomic distribution of high-confidence targets enrich in certain regions due to Matrin3 binding or by chance compared to controls (ADAR random editing sites), we can perform a statistical test such as Fisher's exact test or Chi-squared test. These tests will help us determine if there is a significant association between the class (high-confidence targets vs. controls) and the genomic region distribution.

introns occupy a large fraction of pre-mRNA sequence and are present in the nucleus while transcripts are being processed.
HyperTRIBE editing happens where the fusion protein (MATR3-ADAR) has access to RNA.
Pre-mRNA in the nucleus still contains introns.
Mature cytoplasmic mRNA mostly lacks introns.
So if most edits are in introns, it supports the idea that a big fraction of editing is happening before splicing, i.e., on nuclear transcripts.

because all your datasets are HyperTRIBE-style editing readouts from the same system/tissue, they share strong global biases:
introns are long → more chances to hit intronic bases
coverage/expression drives detection → highly expressed nuclear transcripts dominate
ADAR has sequence/context preferences that apply in every condition

To answer the question, whether those high confidence editing sites are enriched in intronic region by chance or due to Matrin3 binding, we can compare the proportion of genes with intronic editing sites in the high-confidence gene set vs. a control set (e.g., ADAR control background). We can use a contingency table and perform a Fisher's exact test to see if there is a significant association between being a high-confidence gene and having intronic editing sites.

High-confidence MATR3 target genes are more likely than ADAR background control genes to contain at least one intronic editing site. Compared with ADAR background control genes, high-confidence MATR3 targets showed a higher frequency of genes containing at least one intronic editing site, suggesting that the observed intronic enrichment is unlikely to be explained solely by background ADAR editing.

https://rpkgs.datanovia.com/rstatix/reference/fisher_test.html



Why? picking the longest transcript per gene, which is a common choice because it covers all exons, UTRs, and possible intronic regions that MATR3 could bind. 

# Number of high confident site in each target genes
```{r}
m3specific_sites_per_GOI <- m3specific_sites_per_genes %>%
  filter(Target_genes %in% c("GOI", "Collagen"))

ggplot(m3specific_sites_per_GOI, aes(x = reorder(gene_name, -count), y = count, fill = Target_genes)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Number of M3-specific editing sites per GOI",
    x = "Gene",
    y = "Number of M3-specific editing sites",
    fill = "Gene set"
  ) +
  scale_fill_manual(values = c("GOI" = "#F5B355", "Collagen" = "cornflowerblue")) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 80, hjust = 1, font.size = 0.5))
```

# GO term analysis of genes containing Matr3-specific sites 
```{r}
plot_go_enrichment <- function(ego_obj, set_name, n_terms = 6) {  # remove n_show_genes argument
  
  plot_df <- as.data.frame(ego_obj) %>%
    tidyr::separate(GeneRatio, into = c("gene_count", "bg_count"), sep = "/", convert = TRUE) %>%
    mutate(
      GeneRatio_num = gene_count / bg_count,
      ONTOLOGY = as.character(ONTOLOGY),
      geneID = as.character(geneID)
    ) %>%
    filter(
      !is.na(GeneRatio_num),
      is.finite(GeneRatio_num),
      !is.na(p.adjust),
      !is.na(Count),
      !is.na(Description),
      !is.na(ONTOLOGY)
    ) %>%
    group_by(ONTOLOGY) %>%
    slice_min(p.adjust, n = n_terms, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      term_label = str_wrap(Description, width = 32),
      
      # ---- REMOVED: gene_label generation ----
      
      ONTOLOGY = dplyr::recode(ONTOLOGY,
        "BP" = "Biological Process",
        "CC" = "Cellular Component",
        "MF" = "Molecular Function"
      )
    )
  
  plot_df$ONTOLOGY <- factor(plot_df$ONTOLOGY, 
                              levels = c("Biological Process", 
                                         "Cellular Component", 
                                         "Molecular Function"))
  
  ggplot(
    plot_df,
    aes(
      x = GeneRatio_num,
      y = reorder(term_label, GeneRatio_num),
      color = p.adjust,
      size = Count)) +
    geom_point(alpha = 0.6) +
    facet_grid(ONTOLOGY ~ ., scales = "free_y", space = "free_y") +
    scale_color_continuous(low = "red", high = "blue", trans = "reverse") +
    scale_size_continuous(name = "Number of\nMATR3 Targets") + # rename size legend
    labs(
      title = paste("Functional Enrichment of Matrin3", set_name, "genes"),
      x = "Ratio of MATR3 Targets to Total Genes in Term",
      y = NULL,
      color = "Adjusted\np-value") +
    theme_light(base_size = 15) +
    theme(
    strip.text   = element_text(size = 18, face = "bold"),
    axis.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(color = "black", size = 16),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.background = element_rect(fill = "grey20"),
    plot.margin = margin(10, 10, 10, 10))  # reduced right margin since no labels
  
}
```

## GO enrichment of genes containing MATR3-specific sites {.tabset}
```{r}
# Run GO enrichment separately for each ontology
ego_m3_specific <- enrichGO(
  gene          = unique(m3specific_site$GeneID),
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENSEMBL",
  ont           = "ALL",
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

# Build plots
ego_m3_specific_plot <- plot_go_enrichment(
  ego_obj   = ego_m3_specific,
  set_name  = "specific targets",
  n_terms   = 8
)

# # Save each ontology plot as PNG
# ggsave(
#   filename = file.path(
#     "/Users/tinkienoob/Master_thesis/MATR3/Deseq2/GO_deseq2_png/GO_MATR3_specific_targets.png"),
#   plot = ego_m3_specific_plot,
#   width = 9,
#   height = 15,
#   dpi = 600
# )
```

## GO enrichment of control-specific sites {.tabset}
```{r}
#| results: asis
control_anno_gr <- lapply(control_anno_gr, function(df) {
  df %>%
    mutate(GeneID = sub("\\..*", "", gene_id)) # remove version suffix from gene_id
})

ctrl_display_names <- c(
  "only_ADARctrl" = "ADAR specific",
  "only_ΔRRMctrl" = "ΔRRM specific"
)

ego_ctrl_list  <- list()
ego_ctrl_plots <- list()

for (ctrl in names(ctrl_display_names)) {

  display <- ctrl_display_names[[ctrl]]
  cat("### ", display, "\n\n", sep = "")

  genes_ctrl <- unique(control_anno_gr[[ctrl]]$GeneID)
  genes_ctrl <- genes_ctrl[!is.na(genes_ctrl) & genes_ctrl != ""]

  ego <- enrichGO(
    gene          = genes_ctrl,
    OrgDb         = org.Mm.eg.db,
    keyType       = "ENSEMBL",
    ont           = "ALL",
    pAdjustMethod = "BH",
    qvalueCutoff  = 0.05,
    readable      = TRUE
  )

  ego_ctrl_list[[ctrl]] <- ego

  if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) {
    p <- plot_go_enrichment(
      ego_obj  = ego,
      set_name = display,
      n_terms  = 8
    )
    ego_ctrl_plots[[ctrl]] <- p
    print(p)

    # save_thesis_plot(
    #   p,
    #   filename = file.path(
    #     "/Users/tinkienoob/Master_thesis/MATR3/Deseq2/GO_deseq2_png",
    #     paste0("GO_", gsub(" ", "_", display), ".png")),
    #   width = 9, height = 15
    # )
  } else {
    cat("No significant GO terms found for", display, "\n\n")
  }

  cat("\n\n")
}
```

# Biotype enrichment of top targets
```{r}
highconf_sites_biotype_count <- m3specific_site %>%
  group_by(gene_type) %>%
  summarise(n_genes = n(), 
            .groups = "drop") %>%
  arrange(desc(n_genes))

ggplot(highconf_sites_biotype_count, aes(x= reorder(gene_type, n_genes), y = n_genes)) +
  geom_bar(stat = "identity", 
          fill = "steelblue",
          alpha = 0.8) + 
  theme_light(base_size = 14) + 
  labs(
    title = "Gene Biotypes among Matrin3-specific targets", 
    x = "Gene Biotype",
    y = "Number of genes") + coord_flip()
```


