---
title: "pureCLIP"
format: html
editor: visual
---

```{r message=FALSE, warning=FALSE}
library(knitr)
library(dplyr)
library(purrr)
library(rtracklayer)
library(ggplot2)
library(tidyverse)
library(BindingSiteFinder)
library(GenomicRanges)
library(GenomeInfoDb)
library(GenomicFeatures)
library(rtracklayer)
library(GenomicAlignments)
library(eulerr)
out <- "/Users/tinkienoob/Master_thesis/MATR3/pureCLIP_ENA_iCLIP_M3_brain/"
```

# pureCLIP

Like Hypertribe data, pureCLIP data is also needed to filter to collect high confidence binding motifs. https://www.bioconductor.org/packages/release/bioc/vignettes/BindingSiteFinder/inst/doc/vignette.html

iCLIP captures the exact nucleotide where the protein crosslinked to the RNA. PureCLIP is designed to report this single-nucleotide resolution. Therefore, in a standard BED file , a single nucleotide at position: 0-based start, 1-based end.

For iCLIP (crosslinking sites), threshold is added to filter/ distinguish true/ high-affinity binding sites from low-affinity/ background noises.

# Manage the input data

```{r}
# Read high confidence editing sites
m3specific_site <- readRDS("/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/m3specific_sites2.rds")
rank_M3tx_df <- readRDS("/Users/tinkienoob/Master_thesis/MATR3/rank_M3_target_manual.rds") %>%
  mutate(GeneID = sub("(.+)\\..*", "\\1", gene_id))
load("/Users/tinkienoob/Master_thesis/MATR3/Deseq2/dds_df_full/fulldds_DESeq2.RData")

mydds_res_list <- list(
  "Matr3" = res_WT_vs_M3,
  "dKO" = res_WT_vs_dKO,
  "RBM20" = res_WT_vs_RBM20
)

save_thesis_plot <- function(plot, filename,
                             width = 10, height = 6.5,
                             dpi = 300, bg = "white") {
  ggsave(
    filename = filename,
    plot = plot,
    width = width,
    height = height,
    units = "in",
    dpi = dpi,
    bg = bg
  )
}

```

## Build the BSFDataSet

```{r}
M3_pureclip_file <- "/Users/tinkienoob/Master_thesis/MATR3/pureCLIP_ENA_iCLIP_M3_brain/iCLIP_M3_pureclip_crosslink_sites.bed"

# read pureclip peaks
peaks <- rtracklayer::import(con = M3_pureclip_file, 
                             format = "BED", 
                             extraCols=c("additionalScores" = "character"))

# clean pureclip peaks columns
peaks$additionalScores <- NULL
#peaks$name <- NULL

# Optionally keep only standard chromosomes and drop chrM
peaks <- keepStandardChromosomes(peaks, pruning.mode = "coarse") %>% 
     dropSeqlevels(., "chrM", pruning.mode = "coarse")
```

# crosslink sites check?

```{r}
# Import hypertribe bed files for cross-checking
folder <- "/Users/tinkienoob/Master_thesis/MATR3/HT_Bed_peaks"

# List all .bg files
files <- list.files(folder, pattern = "\\.bedGraph$", full.names = TRUE)

# Import all BED files into a named list
HT_bed_list <- lapply(files, function(file) {
  rtracklayer::import(con = file, format = "bedGraph")
})

# Name each element after the file (without suffix)
names(HT_bed_list) <- gsub("\\.bedGraph$", "", basename(files))
```

```{r}
M3_pureclip_df <- read.table(M3_pureclip_file, header=FALSE) %>%
  mutate(
    # V7 is the 7th column where PureCLIP stores the score strings
    CrosslinkScore = as.numeric(str_extract(V7, "(?<=score_CL=)[0-9\\.]+")),
    CoverageScore  = as.numeric(str_extract(V7, "(?<=score_E=)[0-9\\.]+")),
    BindingScore   = as.numeric(str_extract(V7, "(?<=score_B=)[0-9\\.]+"))
  ) %>%
  dplyr::select(-V7) %>%
  distinct()
colnames(M3_pureclip_df)[1:6] <- c("seqnames", "start", "end", "name", "score", "strand")
```

```{r}
hist(
  M3_pureclip_df$BindingScore,
  breaks = 50,
  main = "Distribution of MATR3 Binding Scores",
  xlim = c(0, 500)
)

# Removing sites with very low probability scores (noise)
M3clean_pureclip_df <- M3_pureclip_df[M3_pureclip_df$BindingScore >= 2, ]
M3_pureclip_df
```

# Top 20% binding sites with highest binding score

```{r}
# top 20% threshold
top20_cutoff <- quantile(M3clean_pureclip_df$BindingScore, probs = 0.80, na.rm = TRUE)

M3clean_pureclip_df <- M3clean_pureclip_df %>%
  mutate(
    score_group = if_else(BindingScore >= top20_cutoff, "Top20", "Bottom80")
  )

table(M3clean_pureclip_df$score_group)

######################### Convert to GRanges
pureclip_gr <- GRanges(
  seqnames = M3clean_pureclip_df$seqnames,
  ranges   = IRanges(start = M3clean_pureclip_df$start + 1, end = M3clean_pureclip_df$end),
  strand   = M3clean_pureclip_df$strand
)

mcols(pureclip_gr)$BindingScore   <- M3clean_pureclip_df$BindingScore
mcols(pureclip_gr)$CrosslinkScore <- M3clean_pureclip_df$CrosslinkScore
mcols(pureclip_gr)$CoverageScore  <- M3clean_pureclip_df$CoverageScore
mcols(pureclip_gr)$score_group    <- M3clean_pureclip_df$score_group
```

```{r}
regions_list <- readRDS("/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/mm39_region_list.rds")

annotate_regions_from_list <- function(query_gr, region_list, 
                                       priority = c("CDS", "UTR3", "UTR5", "Intron", "ncRNA", "Other")) {
  region_hits <- lapply(names(region_list), function(reg) {
    hits <- findOverlaps(query_gr, region_list[[reg]], ignore.strand = FALSE)
    tibble(
      query_idx = queryHits(hits),
      region = reg
    )
  }) %>% bind_rows()

  # if one site overlaps multiple regions, keep one by priority
  region_hits <- region_hits %>%
    mutate(region = as.character(region),
           region = if_else(region %in% priority, region, "Other"),
           region = factor(region, levels = priority)) %>%
    arrange(query_idx, region) %>%
    distinct(query_idx, .keep_all = TRUE)

  anno_df <- tibble(query_idx = seq_along(query_gr)) %>%
    left_join(region_hits %>% mutate(region = as.character(region)), by = "query_idx") %>%
    mutate(region = replace_na(region, "Other"))

  anno_df$region
}

mcols(pureclip_gr)$region <- annotate_regions_from_list(
  query_gr = pureclip_gr,
  region_list = regions_list,
  priority = c("CDS", "UTR3", "UTR5", "Intron", "ncRNA", "Other")
)

table(mcols(pureclip_gr)$region, useNA = "ifany")
```

```{r}
region_comp_df <- as.data.frame(mcols(pureclip_gr)) %>%
  dplyr::count(score_group, region, name = "n") %>%
  group_by(score_group) %>%
  mutate(percent = 100 * n / sum(n)) %>%
  ungroup()

ggplot(region_comp_df, aes(x = score_group, y = percent, fill = region)) +
  geom_col(width = 0.7) +
  theme_bw(base_size = 12) +
  labs(
    x = NULL,
    y = "Percentage of pureCLIP sites",
    fill = "Region",
    title = "Region composition of top 20% pureCLIP binding-score sites"
  ) +
  scale_fill_brewer(palette = "Spectral")
```

```{r}
score_region_df <- as.data.frame(mcols(pureclip_gr)) %>%
  mutate(region = factor(region, levels = c("Intron", "UTR3", "UTR5", "CDS", "Other")))

ggplot(score_region_df, aes(x = region, y = BindingScore, fill = region)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  coord_cartesian(ylim = quantile(score_region_df$BindingScore, c(0.05, 0.95), na.rm = TRUE)) +
  theme_bw(base_size = 12) +
  labs(
    x = NULL,
    y = "BindingScore",
    title = "BindingScore distribution across transcript regions"
  )
```

# Import bigWig

```{r}
bw_path <- "/Users/tinkienoob/Master_thesis/MATR3/matr3_iCLIP/BamToBigWig"

bw.plus <- list.files(bw_path, pattern = "plus.bw$", full.names = TRUE, recursive = TRUE)

bw.minus <- list.files(bw_path, pattern = "minus.bw$", full.names = TRUE, recursive = TRUE)

```

```{r}
# read prepared annotation
gns1 <- readRDS("/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/gns_mm39.rds")

region_anno <- readRDS("/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/mm39_region_list.rds")
```

# Peaks from pureclip

```{r}
# make metadata for BindingSiteFinder
meta <- data.frame(
  id = seq_along(bw.plus), 
  condition = factor("WT"),
  clPlus = bw.plus,
  clMinus = bw.minus
)

meta
```

# run BindingSiteFinder in automatic mode

```{r}
# create BindingSiteFinder object
bds_object <- BSFDataSetFromBigWig(ranges = peaks, 
                           meta = meta, 
                           silent = FALSE)
```

# Make inital BS allowing BSF to determine automatic parameters

```{r}
bds_automatic <- BSFind(bds_object, 
                 anno.genes = gns1, 
                 anno.transcriptRegionList = region_anno, 
                 cutoff.geneWiseFilter = 0,
                 repro.nReps = 2,
                 repro.cutoff = 0.05)

# save automatic binding sites
saveRDS(bds_automatic, paste0(out, "bds_automatic.rds"))
```

# Binding site summary

```{r}
bds_automatic <- readRDS(paste0(out, "bds_automatic.rds"))
bds_automatic
estimateBsWidthPlot(bds_automatic)
```

# Visualize steps and filters

```{r}
processingStepsFlowChart(bds_automatic)
```

# Pre-filtering of crosslink sites

```{r}
png(paste0(out, "globalfilter_bs5.png"), width = 1400, height = 800, res = 300)
pureClipGlobalFilterPlot(bds_selected_bs5)
dev.off()
```

# Gene-wise filter

The gene-wise filter cutoff estimated in the previous step (cutoff.geneWiseFilter) is applied to keep only the top percent of crosslink sites, ranked by their PureCLIP score. Many duplicates/ overlaps occurs due to dense single 1 nucleotide crosslink sites. The presence of duplicates can indicate that the algorithm is correctly identifying clusters of crosslinking events, which may represent true binding sites. However, it can also suggest that the parameters for merging crosslinks into binding sites may need to be adjusted to better capture the underlying biology without over-merging distinct sites.

```{r}
duplicatedSitesPlot(bds_automatic)
```

```{r}
png(paste0(out, "summaryplot_bs5.png"), width = 1600, height = 800, res = 250)
makeBsSummaryPlot(bds_selected_bs5)
dev.off()

```

# Compute binding sites with different widths

```{r}
bds1 <- makeBindingSites(object = bds_object, bsSize = 3)
bds2 <- makeBindingSites(object = bds_object, bsSize = 5)
bds3 <- makeBindingSites(object = bds_object, bsSize = 7)
bds4 <- makeBindingSites(object = bds_object, bsSize = 9)
bds5 <- makeBindingSites(object = bds_object, bsSize = 13)
bds6 <- makeBindingSites(object = bds_object, bsSize = 19)

# summarize in list
l = list(`1. bsSize = 3` = bds1, `2. bsSize = 5` = bds2, 
         `3. bsSize = 7` = bds3, `4. bsSize = 9` = bds4,
         `5. bsSize = 13` = bds5, `6. bsSize = 19` = bds6)

# plot comparison with rangeCoveragePlot() is averaging crosslink coverage across all binding sites in each set.
```

```{r}
png(paste0(out, "rangeCoverage.png"), width = 1600, height = 1000, res = 250)
rangeCoveragePlot(l, width = 20, show.samples = TRUE, subset.chromosome = "chr3")
dev.off()

```

# Merge crosslink sites into binding sites

Region-to-fit width → How many nucleotides wide the candidate region is when BindingSiteFinder tries to merge nearby crosslinks.

Region-to-fit #N → Number of regions that need that width to fit the crosslinks. Most binding evidence is concentrated in 1–3 nt windows.

```{r}
mergeCrosslinkDiagnosticsPlot(bds_automatic)
```

```{r}
png(paste0(out, "regionenrich_bs5.png"), width = 1400, height = 800, res = 250)
transcriptRegionSpectrumPlot(bds_selected_bs5)
dev.off()

```

RBP is likely involved in pre-mRNA processing or alternative splicing rather than translation or mRNA stability (which would favor UTRs).

# Ensure reproducibility among replicates

```{r}
reproducibilityFilterPlot(bds_automatic)
```

```{r}
png(paste0(out, "reprodplot_bs5.png"), width = 1300, height = 1000, res = 250)
reproducibilitySamplesPlot(bds_selected_bs5)
dev.off()
```

# Genomic target idenitfication

```{r}
geneOverlapsPlot(bds_automatic)
```

```{r}
targetGeneSpectrumPlot(bds_automatic)
```

# Transcript region identification

```{r}
p <- transcriptRegionSpectrumPlot(bds_selected_bs5) 
p

df <- p$data
library(ggplot2)
library(scales)

# use your table directly (more robust than p$data)
df <- data.frame(
  region = c("Intron", "3'UTR", "CDS", "5'UTR"),
  value  = c(9491, 447, 24, 7)
)

pp <- ggplot(df, aes(x = value, y = reorder(region, value))) +
  
  geom_col(fill = "#7A3B5D", width = 0.7) +
  
  geom_text(
    aes(label = comma(value)),
    color = "#D55E00",      # strong contrast (better than default orange)
    hjust = -0.15,
    size = 6,               # BIG for thesis
    fontface = "bold"
  ) +
  
  scale_x_continuous(
    labels = comma,
    expand = expansion(mult = c(0, 0.2))
  ) +
  
  labs(
    title = "transcriptRegionSpectrumPlot()",
    x = "#N Binding sites",
    y = "Transcript region"
  ) +
  
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    plot.subtitle = element_text(size = 14),
    
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 14),
    axis.text.y = element_text(face = "bold"),
    
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

save_thesis_plot(pp, "/Users/tinkienoob/Master_thesis/MATR3/pureCLIP_ENA_iCLIP_M3_brain/regionrich_bs5.png", 
                 width = 8, height = 6)

pp
```

```{r}
bindingSiteDefinednessPlot(bds_automatic, by = "transcript")
```

# Binding site strength

```{r}
globalScorePlot(bds_automatic)
```

```{r}
bindingSiteCoveragePlot(bds_automatic, plotIdx = 8, flankPos = 100, autoscale = TRUE)
```

```{r}
pSites <- getRanges(bds_object)
bindingSiteCoveragePlot(bds_automatic, plotIdx = 8, flankPos = 20, autoscale = TRUE,
                        customRange = pSites, customRange.name = "pSites", shiftPos = -10)
```

# Merge replicate signal

```{r}
bdsMerge <- collapseReplicates(bds_object)[1:1000]
covTotal <- coverageOverRanges(bdsMerge, returnOptions = "merge_positions_keep_replicates")

covRep <- coverageOverRanges(bds_object[1:1000], returnOptions = "merge_positions_keep_replicates")

df <- cbind.data.frame(mcols(covTotal), mcols(covRep)) %>%
    mutate(rep1 = round(`1_WT`/ WT, digits = 2) * 100,
           rep2 = round(`2_WT`/ WT, digits = 2) * 100,
           rep3 = round(`3_WT`/ WT, digits = 2) * 100,
           rep4 = round(`4_WT`/ WT, digits = 2) * 100,
           rep5 = round(`5_WT`/ WT, digits = 2) * 100) %>%
    tibble::rowid_to_column("BsID") %>%
    dplyr::select(BsID, rep1, rep2, rep3, rep4, rep5) %>%
    pivot_longer(-BsID) %>%
    group_by(BsID) %>%
    arrange(desc(value), .by_group = TRUE) %>%
    mutate(name = factor(name, levels = name)) %>%
    group_by(name) %>%
    arrange(desc(value), .by_group = TRUE) %>%
    mutate(BsID = factor(BsID, levels = BsID))


ggplot(df, aes(x = BsID, y = value, fill = name)) +
    geom_col(position = "fill", width = 1) +
    theme_bw() +
    scale_fill_brewer(palette = "Set3") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 7)) +
    labs(x = "Binding site ID",
         y = "Percentage",
         fill = "Replicate"
         ) +
    scale_y_continuous(labels = scales::percent)
```

```{r skip-this-chunk, eval = FALSE}
# Customize pipeline with specific parameters
bds_selected_width <- BSFind(bds_object, 
                 anno.genes = gns1, 
                 anno.transcriptRegionList = region_anno, 
                 bsSize = 5,
                 cutoff.geneWiseFilter = 0.6,
                 repro.nReps = 2,
                 repro.cutoff = 0.05,
                 overlaps.geneAssignment = "frequency",
                 overlaps.TranscriptRegions = "frequency",
                 est.maxBsWidth = 29,
                 veryQuite = FALSE)

saveRDS(bds_selected_width, paste0(out, "bds_selected_width.rds"))
```

```{r}
bds_selected_bs5 <- readRDS(paste0(out, "bds_selected_width.rds"))
bds_selected_bs5
# Visualize steps and filters
bs5_flowchart <- processingStepsFlowChart(bds_selected_bs5)

save_thesis_plot(bs5_flowchart, "/Users/tinkienoob/Master_thesis/MATR3/pureCLIP_ENA_iCLIP_M3_brain/BS5_flowchart.png", 
                 width = 6, height = 6)
```

```{r}
bindingSiteCoveragePlot(bds_selected_bs5, plotIdx = 8, flankPos = 20, autoscale = TRUE,
                        customRange = pSites, customRange.name = "pSites", shiftPos = -10)
```

```{r}
# export binding sites as .bed
# exportToBED(bds_selected_bs5, con = paste0(out, "BindingSites_bs5.bed"))
# exportToBED(bds_automatic, con = paste0(out, "BindingSites_auto.bed"))
```

# Overlap binding sites with genes from gns

```{r}

bs5_gr <- getRanges(bds_selected_bs5)
# Find overlaps: binding sites that fall in gene bodies
bs5_hits <- GenomicRanges::findOverlaps(bs5_gr, gns1, ignore.strand = TRUE)

# 3) Extract unique gene IDs overlapped by any binding site
gene_ids_bs5 <- unique(names(gns1)[subjectHits(bs5_hits)])
#gene_ids_bs5 <- sub("\\..*$", "", gene_ids_bs5)
```

# Overlap of genes between M3-specific editing with binding sites

```{r}
bs_es_genes_overlap <- list(
  #`iCLIP binding-site\nenriched genes` = unique(gene_ids_bs5),
 # `Top M3 targets`   = unique(rank_M3tx_df$gene_id),
  `HyperTRIBE\nMATR3-specific targets` = unique(m3specific_site$gene_id),
 `MATR3-DEGs` = unique(mydds_res_list$Matr3 %>% filter(regulated == TRUE) %>% pull(gene_id))
)

fit <- eulerr::euler(bs_es_genes_overlap)

png("/Users/tinkienoob/Master_thesis/MATR3/Deseq2/es_deg_overlap.png",
    width = 10, height = 6, units = "in", res = 800, bg = "white")

plot(fit,
  fills     = list(fill = c("#009E73", "dodgerblue3"), alpha = 0.5),
  edges     = list(col = "white"),
  stroke_size = 1,

  # ── set/circle label size ─────────────────────────────────────────────
  labels    = list(
    fontsize = 18,          # size of "Binding Sites" / "Editing Sites" labels
    fontface = "bold"
  ),

  # ── count/quantity label size ─────────────────────────────────────────
  quantities = list(
    fontsize = 20,          # size of the numbers inside circles
    fontface = "plain"
  )
)

dev.off()

```

```{r}
bs_tier_overlap_list <- list(
  `iCLIP binding-site enriched genes` = unique(gene_ids_bs5),
  `Tier 1` = unique(rank_M3tx_df$gene_id[rank_M3tx_df$rank == "Tier 1"]),
  `Tier 2` = unique(rank_M3tx_df$gene_id[rank_M3tx_df$rank == "Tier 2"]),
  `Tier 3` = unique(rank_M3tx_df$gene_id[rank_M3tx_df$rank == "Tier 3"])
)

fit_bs_tiers <- eulerr::euler(bs_tier_overlap_list)
fit_bs_tiers

png("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png/BS_Tier_euler.png",
    width = 10, height = 7, units = "in", res = 600, bg = "white")

plot(
  fit_bs_tiers,
  fills = list(fill = c("#009E73", tier_colors), alpha = 0.55),
  edges = list(col = "white", lwd = 1),
  quantities = list(font = 2, cex = 1.2),
  labels = list(font = 2, cex = 1.2),
  percent = TRUE
)

dev.off()
```

```{r}
tier_colors <- c(
  "Tier 1" = "#CD534C", 
  "Tier 2" = "#EFC050", 
  "Tier 3" = "#45B7D1",
  "Others" = "grey80")

# Overlap binding sites with Hypertribe peaks
# overlap_bs_es_genes <- Reduce(
#   intersect,
#   list(
#     bs_es_genes_overlap$`Genes with Bs5`,
#     bs_es_genes_overlap$`MATR3-specific targets`,
#     bs_es_genes_overlap$`MATR3-DEGs`))

# overlapping genes = bs ∩ es
overlap_bs_es <- intersect(
  unique(gene_ids_bs5),
  unique(m3specific_site$gene_id)
)

# total genes per tier
total_tier <- rank_M3tx_df %>%
  filter(rank %in% c("Tier 1", "Tier 2", "Tier 3")) %>%
  dplyr::count(rank, name = "total")

# overlap genes per tier
overlap_tier <- tibble(gene_id = overlap_bs_es) %>%
  left_join(
    rank_M3tx_df %>% dplyr::select(gene_id, rank),
    by = "gene_id"
  ) %>%
  filter(rank %in% c("Tier 1", "Tier 2", "Tier 3")) %>%
  dplyr::count(rank, name = "overlap")

# merge and compute percentage
bar_df <- total_tier %>%
  left_join(overlap_tier, by = "rank") %>%
  mutate(
    overlap = replace_na(overlap, 0),
    pct     = round(100 * overlap / total, 1),
    label   = paste0(overlap, "/", total, " (", pct, "%)"),
    rank    = factor(rank, levels = c("Tier 3", "Tier 2", "Tier 1"))  # bottom→top
  )

p_bar <- ggplot(bar_df) +

  # ── background bar = total ───────────────────────────────────────────────
  geom_col(aes(x = rank, y = total, fill = rank),
           alpha = 0.30, width = 0.8) +

  # ── foreground bar = overlap ─────────────────────────────────────────────
  geom_col(aes(x = rank, y = overlap, fill = rank),
           alpha = 1.00, width = 0.8) +

  # ── percentage label to the right of background bar ─────────────────────
  geom_text(aes(x = rank, y = total, label = label),
            hjust   = 0.2,
            fontface = "bold",
            size    = 5) +

  scale_fill_manual(values = tier_colors) +

  coord_flip(clip = "off") +

  scale_y_continuous(
    expand = expansion(mult = c(0, 0.28))
  ) +

  labs(
    title = "",
    x     = NULL,
    y     = "Number of genes",
    fill  = NULL
  ) +

  theme_bw(base_size = 12) +
  theme(
    legend.position = "none",
    strip.text   = element_text(size = 18, face = "bold"),
    axis.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(color = "black", size = 16),
    axis.text.y = element_text(color = "black", size = 16, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16)
  ) 

save_thesis_plot(p_bar, "/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png/BS_Tier_barplot.png", 
                 width = 6.5, height = 6)

p_bar       
```

```{r}
# Total genes per tier
total_tier_df <- rank_M3tx_df %>%
  dplyr::count(rank, name = "total_genes")

# Overlap genes per tier
overlap_tier_df <- tibble(GeneID = overlap_bs_es_genes) %>%
  left_join(rank_M3tx_df, by = "GeneID") %>%
  filter(!is.na(rank)) %>%
  dplyr::count(rank, name = "overlap_genes")

# Percentage table for labels
perc_tier_df <- total_tier_df %>%
  left_join(overlap_tier_df, by = "rank") %>%
  mutate(
    overlap_genes = replace_na(overlap_genes, 0),
    percentage = round(100 * overlap_genes / total_genes, 1),
    label = paste0(percentage, "%")
  )

ggplot(total_tier_df, aes(x = rank, y = total_genes)) +
  geom_col(aes(fill = rank), alpha = 0.35) +
  geom_col(
    data = overlap_tier_df,
    aes(x = rank, y = overlap_genes, fill = rank)) +
  coord_flip(clip = "off") +
  theme_bw() +
  labs(
    title = "Overlap genes relative to total genes in each tier",
    x = NULL,
    y = "Number of genes") +
  scale_fill_manual(values = tier_colors) +
  geom_text(
    data = perc_tier_df,
    aes(
      x = rank,
      y = total_genes + max(total_tier_df$total_genes) * 0.03,
      label = paste0(overlap_genes, "/", total_genes, " (", percentage, "%)")),
    hjust = 0.1,
    size = 3.5) +
  theme(
    legend.position = "none",
    strip.text   = element_text(size = 18, face = "bold"),
    axis.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(color = "black", size = 16),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16)
  ) +
  expand_limits(y = max(total_tier_df$total_genes) * 1.22)
```

# Session Info

```{r, message=FALSE, warnings=FALSE}
sessionInfo()
```
