---
title: "Find top Matrin3 target transcripts"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
author:
  - name: "Kim Nguyen"
format:
  html:
    code-fold: TRUE
    code-overflow: scroll
    code-summary: "Show code"
    code-tools: true
    code-line-numbers: true
    toc: true
    toc-depth: 5
    toc-location: right
    toc-expand: true
    number-sections: true
    theme: sandstone
    fontsize: 11pt
    linestretch: 1.5
    fig-format: png
    cap-location: margin
    body-width: 900px
    sidebar-width: 300px
    page-layout: full
    crossref:
      fig-title: Fig
    embed-resources: true
    link-external-newwindow: true
    smooth-scroll: true
    css: css.css
    execute:
      echo: true
      warning: false
editor_options: 
  chunk_output_type: inline
---

 The mRNA targets of RBPs were identified using HyperTRIBE. De novo motif discovery analysis was conducted using the XSTREME tool from the MEME suite. Differential gene expression analysis comparing RNA-seq data was performed using DESeq2. The GO term enrichment analysis was carried out using the R package gProfiler2 and visualized with the Cytoscape plugin, EnrichmentMap 
 
# Library
```{r setup, message=FALSE, warning=FALSE}
# --- Core Data Manipulation ---
library(tidyverse)
library(magrittr)
library(DT)
library(kableExtra)

# Bioinformatics & Bioconductor
library(GenomicRanges)
library(rtracklayer)
library(DESeq2)
library(clusterProfiler)
library(biomaRt)
library(org.Mm.eg.db)
library(Rsamtools)
library(ggpmisc)

# Visualization
library(ggrepel)
library(ggvenn)
library(eulerr)
library(patchwork) 
library(RColorBrewer)
library(ggpubr)
library(cowplot)
library(ggstatsplot)

# --- Path ---
outpath  <- "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/"
```

# Load data from previous analysis (M3-specific)
```{r import-data, message=FALSE, warning=FALSE}
############## M3-specific sites
m3specific_site <- readRDS("/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/m3specific_sites2.rds") %>%
  mutate(GeneID = sub("\\..*", "", gene_id)) # remove version suffix from gene_id

############## Annotation
annoDb <- AnnotationDbi::loadDb("/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/annoDb.Db")
gns    <- readRDS("/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/gns_mm39.rds")
regions_list <- readRDS("/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/mm39_region_list.rds")
gene_anno_df <- readRDS("/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/geneanno_df_mm39.rds")
all_control_list_HT <- readRDS("/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/HT_control_list.rds")
```


```{r setup-function, message=FALSE, warning=FALSE}
# Standardize all plots for your thesis
theme_thesis <- function(base_size = 16) {
  theme_bw(base_size = base_size) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = base_size + 2),
      axis.title = element_text(face = "bold", size = base_size),
      axis.text = element_text(color = "black", size = base_size - 1),
      legend.title = element_text(face = "bold", size = base_size - 1),
      legend.text = element_text(size = base_size - 2),
      strip.text = element_text(face = "bold", size = base_size - 1)
    )
}

# --- Color Palettes ---
region_colors <- c(
  "UTR5"   = "#3288BD", 
  "CDS"    = "#99D594", 
  "Intron" = "#FDAE61", 
  "UTR3"   = "#D53E4F", 
  "Others" = "#E0E0E0"
)

tier_colors <- c(
  "Tier1" = "#CD534C", 
  "Tier2" = "#EFC050", 
  "Tier3" = "#45B7D1",
  "Others" = "grey80"
)
```
Genomic ranges: https://biodatascience.github.io/compbio/bioc/ranges.html 
Because 1 editing site can be annotated to multiple transcripts of the same gene, I will use gene_id as the main identifier for annotation and downstream analysis.

## Genomic features annotation
### Make annotation database from gtf file
annotation file is downloaded from: https://www.gencodegenes.org/mouse/release_M37.html
The gencode.vM37.basic.annotation.gtf file is designed for the mouse genome assembly GRCm39 (mm39). Released around May 2025, this GENCODE dataset (vM37) 

makeTxDbFromGFF builds a structural map including Only the strictly necessary coordinates to define gene structure (Chromosome, Start, End, Strand, IDs)
while rtracklayer::import reads the full descriptive content in the GTF file (e.g., gene_id, gene_name, gene_biotype, tag) becomes a column in the metadata (mcols).

```{r message=FALSE}
annoFile = "/Users/tinkienoob/Master_thesis/MATR3/mm39_anno/gencode.vM37.basic.annotation.gtf" 

# Build TxDb from GFF3
annoDb <- txdbmaker::makeTxDbFromGFF(file = annoFile, format = "gtf")

# Import full annotation for metadata
annoInfo <- rtracklayer::import(annoFile, format = "gtf")

# Save the TxDb object to file
AnnotationDbi::saveDb(annoDb, "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/annoDb.Db")

# Extract genes as GRanges
gns <- genes(annoDb)

# Strip version suffixes from GENCODE IDs (e.g., .18) ---
#gns$gene_id <- sub("\\..*", "", gns$gene_id)
#annoInfo$gene_id <- sub("\\..*", "", annoInfo$gene_id)

# Match gene IDs between TxDb and Gtf
idx <- match(gns$gene_id, annoInfo$gene_id)

# Combine metadata from TxDb + Gtf
meta <- cbind(elementMetadata(gns),
              elementMetadata(annoInfo)[idx,])
meta <- meta[,!duplicated(colnames(meta))]

elementMetadata(gns) <- meta
```
## Count events per region per dataset
```{r}
region_anno_events <- function(gr, regions_list) {
  
  # initialize metadata
  region    <- rep(NA_character_, length(gr))
  region_id <- rep(NA_character_, length(gr))
  
  # loop over regions in GRangesList
  for (x in names(regions_list)) {
    hits <- GenomicRanges::findOverlaps(gr, regions_list[[x]], type="any", ignore.strand=FALSE)
    
    region[queryHits(hits)[is.na(region[queryHits(hits)])]] <- x
  }
  
  # add metadata back to gr
  mcols(gr)$region    <- region
  mcols(gr)$region_id <- region_id
  
  return(gr)
}

################################ M3-specific sites annotation
m3specific_site_gr <- makeGRangesFromDataFrame(m3specific_site,
    seqnames.field = "seqnames",
    start.field    = "start",
    end.field      = "end",
    keep.extra.columns = TRUE)
m3specific_site_gr_anno <- region_anno_events(m3specific_site_gr, regions_list)

# saveRDS(m3specific_site_gr_anno, "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/m3spec_Granges.rds")
```

# Calculate Region Lengths per Gene
```{r}
# Map lengths to the dataframe, and get gene_id for each transcript
tx2gene <- AnnotationDbi::select(annoDb, keys = keys(annoDb, "TXNAME"), 
                                 columns = c("TXNAME", "GENEID"), keytype = "TXNAME") %>%
  dplyr::rename(tx_name = TXNAME, gene_id = GENEID) %>%
  filter(!is.na(gene_id))

# CDS
cds_by_gene <- cdsBy(annoDb, by = "gene")
cds_reduced <- GenomicRanges::reduce(cds_by_gene)
cds_len     <- sum(width(cds_reduced)) # Named integer vector by GENEID

# 5' UTR
utr5_by_gene <- fiveUTRsByTranscript(annoDb, use.names = TRUE)
utr5_gr      <- unlist(utr5_by_gene)
utr5_gr$gene_id <- tx2gene$gene_id[match(names(utr5_gr), tx2gene$tx_name)]
utr5_reduced <- GenomicRanges::reduce(split(utr5_gr, utr5_gr$gene_id))
utr5_len     <- sum(width(utr5_reduced))

# 3' UTR
utr3_by_gene <- threeUTRsByTranscript(annoDb, use.names = TRUE)
utr3_gr      <- unlist(utr3_by_gene)
utr3_gr$gene_id <- tx2gene$gene_id[match(names(utr3_gr), tx2gene$tx_name)]
utr3_reduced <- GenomicRanges::reduce(split(utr3_gr, utr3_gr$gene_id))
utr3_len     <- sum(width(utr3_reduced))

# Introns
intron_by_gene <- intronsByTranscript(annoDb, use.names = TRUE)
introns_gr     <- unlist(intron_by_gene)
introns_gr$gene_id <- tx2gene$gene_id[match(names(introns_gr), tx2gene$tx_name)]
introns_reduced <- GenomicRanges::reduce(split(introns_gr, introns_gr$gene_id))
introns_len     <- sum(width(introns_reduced))
```

# Region length df in gene level 
```{r}
# Create a dataframe of all transcript names
all_genes <- unique(c(names(utr5_len), names(cds_len), names(utr3_len), names(introns_len)))

region_lengths_df <- data.frame(
  gene_id = all_genes,
  UTR5 = as.numeric(utr5_len[all_genes]),
  CDS  = as.numeric(cds_len[all_genes]),
  UTR3 = as.numeric(utr3_len[all_genes]),
  Intron = as.numeric(introns_len[all_genes]))

# transpose df 
region_length_df <- region_lengths_df %>%
  replace(is.na(.), 0) %>%
  pivot_longer(cols = c(UTR5, CDS, UTR3, Intron), names_to = "region", values_to = "region_length_kb") %>%
  mutate(region_length_kb = region_length_kb / 1000) %>%
  filter(region_length_kb > 0)
```


## Region distribution of Matrin3-specific sites vs. control
```{r}
region_colors <- c(
  "UTR5"   = "#3288BD", 
  "CDS"    = "#99D594", 
  "Intron" = "#FDAE61", 
  "UTR3"   = "#D53E4F"
)

# set class order
class_order <- c("M3-specific", "ADAR specific", "ΔRRM specific")
# set region order manually if you want full control
# change this to match your real annotation names exactly
region_order <- c("UTR5", "CDS", "Intron", "UTR3")

m3_region_counts <- as.data.frame(m3specific_site_gr_anno) %>%
  mutate(
    region = ifelse(
      is.na(region) | region == "",
      "Others",
      region
    )
  ) %>%
  filter(
    region %in% c("UTR5", "UTR3", "CDS", "Intron")
  ) %>%
  group_by(region) %>%
  summarise(
    count = n(),
    .groups = "drop"
  ) %>%
  mutate(
    percentage = count / sum(count) * 100,
    class = "M3-specific"
  )

control_region_counts <- bind_rows(

  lapply(names(all_control_list_HT), function(cond) {

    df <- all_control_list_HT[[cond]]

    df %>%
      mutate(
        region = ifelse(
          is.na(region) | region == "",
          "Others",
          region
        )
      ) %>%
      filter(
        region %in% c("UTR5", "UTR3", "CDS", "Intron")
      ) %>%
      group_by(region) %>%
      summarise(
        count = n(),
        .groups = "drop"
      ) %>%
      mutate(
        percentage = count / sum(count) * 100,
        class = cond
      )

  })

) %>%
  mutate(
    class = recode(
      class,
      "only_ADARctrl" = "ADAR specific",
      "only_ΔRRMctrl" = "ΔRRM specific"
    )
  )

regions_count_data <- bind_rows(
  m3_region_counts,
  control_region_counts
)

regions_count_data <- regions_count_data %>%
  mutate(
    class = factor(
      class,
      levels = class_order
    ),
    region = factor(
      region,
      levels = region_order
    )
  ) %>%
  arrange(class, region)

region_stacked_raw <- ggplot(
  regions_count_data,
  aes(
    x = class,
    y = percentage,
    fill = region
  )
) +
  geom_col(
    color = "black",
    linewidth = 0.25,
    width = 0.6
  ) +
  geom_text(
    aes(
      label = ifelse(
        percentage >= 3,
        paste0(round(percentage, 1), "%"),
        ""
      )
    ),
    position = position_stack(vjust = 0.5),
    size = 5.5,
    fontface = "bold"
  ) +
  scale_fill_manual(
    values = region_colors,
    drop = FALSE
  ) +
  labs(
    title = "",
    x = NULL,
    y = "Proportion of editing sites",
    fill = "Region"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "right",
    legend.title = element_text(
      size = 13,
      face = "bold"
    ),
    legend.text = element_text(size = 12),
    axis.title = element_text(
      face = "bold",
      size = 14
    ),
    axis.text.x = element_text(
      color = "black",
      size = 14,
      face = "bold"
    ),
    axis.text.y = element_text(size = 14)
  )

region_stacked_raw

save_thesis_plot(
  region_stacked_raw,
  filename = "/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png/Region_distribution_raw.png",
  width = 7.5,
  height = 6
)

```


```{r}
# count number of each region in M3-specific sites
m3_site_counts <- as.data.frame(m3specific_site_gr_anno) %>%
  mutate(
    region = ifelse(
      is.na(region) | region == "",
      "Others",
      region
    )
  ) %>%
  filter(
    region %in% c("UTR5", "UTR3", "CDS", "Intron")
  ) %>%
  group_by(region) %>%
  summarise(
    count = n(),
    .groups = "drop"
  )


m3_genes <- unique(
  as.data.frame(m3specific_site_gr_anno)$gene_id
)

m3_region_lengths <- region_length_df %>%
  filter(
    gene_id %in% m3_genes,
    region %in% c("UTR5", "UTR3", "CDS", "Intron")
  ) %>%
  group_by(region) %>%
  summarise(
    total_region_kb = sum(region_length_kb, na.rm = TRUE),
    .groups = "drop"
  ) 

m3specific_region_density <- m3_site_counts %>%
  left_join(
    m3_region_lengths,
    by = "region"
  ) %>%
  mutate(
    editing_density = count / total_region_kb,
    percentage = editing_density / sum(editing_density) * 100,
    class = "M3-specific"
  )

# count number of each region in controls
control_density_df <- bind_rows(

  lapply(names(all_control_list_HT), function(cond) {

    df <- all_control_list_HT[[cond]]

    site_counts <- df %>%
      mutate(
        region = ifelse(
          is.na(region) | region == "",
          "Others",
          region
        )
      ) %>%
      filter(
        region %in% c("UTR5", "UTR3", "CDS", "Intron")
      ) %>%
      group_by(region) %>%
      summarise(
        count = n(),
        .groups = "drop"
      ) 


    genes_in_control <- unique(df$gene_id)

    region_lengths <- region_length_df %>%
      filter(
        gene_id %in% genes_in_control,
        region %in% c("UTR5", "UTR3", "CDS", "Intron")
      ) %>%
      group_by(region) %>%
      summarise(
        total_region_kb = sum(region_length_kb, na.rm = TRUE),
        .groups = "drop"
      )

    site_counts %>%
      left_join(region_lengths, by = "region") %>%
      mutate(
        editing_density = count / total_region_kb,
         percentage = editing_density / sum(editing_density) * 100,
        class = cond
      )

  })

) %>%
  mutate(
    class = recode(
      class,
      "only_ADARctrl" = "ADAR specific",
      "only_ΔRRMctrl" = "ΔRRM specific"
    )
  )

# combine
m3_region_density_df <- bind_rows(
  m3specific_region_density,
  control_density_df
)

# make sure missing regions still appear if absent in some classes
m3_region_density_df <- m3_region_density_df %>%
  mutate(
    class = factor(class, levels = class_order),
    region = ifelse(region %in% region_order, region, "Others"),
    region = factor(region, levels = region_order)
  ) %>%
  arrange(class, region)
```


```{r}
##################### Stacked barplot = normalized_density = editing_density / sum(editing_density)
region_stacked <- ggplot(m3_region_density_df, aes(x = class, y = editing_density, fill = region)) +
  geom_col(color = "black", position = "fill", linewidth = 0.25, width = 0.6) +
  geom_text(
    aes(label = ifelse(percentage >= 3, paste0(round(percentage, 1), "%"), "")),
    position = position_fill(vjust = 0.5),
    size = 5.5,
    fontface = "bold") +
  scale_fill_manual(values = region_colors, drop = FALSE) +
  labs(
    title = "",
    x = NULL, y = "Proportion of normalized editing density", 
    fill = "Region") +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    legend.position = "right",
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(color = "black", size = 14, face = "bold"),
    axis.text.y = element_text(size = 14)
  )

region_stacked

save_thesis_plot(
  region_stacked,
  filename = "/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png/Region_distribution.png",
  width = 7.5, height = 6
)
```

```{r fisher-regional-distribution}

# Fisher's exact test: for each region, test whether its proportion of sites
# differs between M3-specific (Tier 1) and each comparison group (Tier 2, Tier 3).
# 2x2 table per region per comparison:
#               | in region | not in region |
#   M3-specific |     a     |       b       |
#   Group G     |     c     |       d       |

regions_of_interest <- c("UTR5", "UTR3", "CDS", "Intron")

# Helper: get per-region raw counts from a data frame
get_region_counts <- function(df) {
  df %>%
    mutate(region = ifelse(is.na(region) | region == "", "Others", region)) %>%
    filter(region %in% regions_of_interest) %>%
    count(region, name = "n") %>%
    complete(region = regions_of_interest, fill = list(n = 0))
}

# Tier 1 counts
tier1_counts <- get_region_counts(as.data.frame(m3specific_site_gr_anno))
tier1_total  <- sum(tier1_counts$n)

# Comparison groups: Tier 2 (ADAR-specific) and Tier 3 (ΔRRM-specific)
comparison_groups <- list(
  "Tier2 (ADAR specific)"  = all_control_list_HT[["only_ADARctrl"]],
  "Tier3 (ΔRRM specific)"  = all_control_list_HT[["only_ΔRRMctrl"]]
)

# Run 2x2 Fisher test for each region × comparison group
fisher_results <- lapply(names(comparison_groups), function(grp_name) {

  grp_counts <- get_region_counts(comparison_groups[[grp_name]])
  grp_total  <- sum(grp_counts$n)

  lapply(regions_of_interest, function(reg) {

    a <- tier1_counts$n[tier1_counts$region == reg]       # Tier1 in region
    b <- tier1_total - a                                   # Tier1 not in region
    c <- grp_counts$n[grp_counts$region == reg]            # Group in region
    d <- grp_total - c                                     # Group not in region

    tbl <- matrix(c(a, b, c, d), nrow = 2, byrow = TRUE,
                  dimnames = list(
                    c("M3-specific", grp_name),
                    c("in_region", "not_in_region")
                  ))

    ft <- fisher.test(tbl)

    data.frame(
      region     = reg,
      comparison = paste0("M3-specific vs ", grp_name),
      count_M3   = a,
      count_grp  = c,
      odds_ratio = ft$estimate,
      p_value    = ft$p.value
    )
  }) %>% bind_rows()

}) %>% bind_rows()

# BH correction across all tests
fisher_results <- fisher_results %>%
  mutate(
    p_adj        = p.adjust(p_value, method = "BH"),
    significance = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ "ns"
    )
  ) %>%
  arrange(comparison, region)

print(fisher_results)
```

# GO term analysis of genes containing Matr3-specific sites
```{r}
plot_go_enrichment <- function(ego_obj, set_name, n_terms = 6) {  # remove n_show_genes argument
  
  plot_df <- as.data.frame(ego_obj) %>%
    tidyr::separate(GeneRatio, into = c("gene_count", "bg_count"), sep = "/", convert = TRUE) %>%
    mutate(
      GeneRatio_num = gene_count / bg_count,
      ONTOLOGY = as.character(ONTOLOGY),
      geneID = as.character(geneID)
    ) %>%
    filter(
      !is.na(GeneRatio_num),
      is.finite(GeneRatio_num),
      !is.na(p.adjust),
      !is.na(Count),
      !is.na(Description),
      !is.na(ONTOLOGY)
    ) %>%
    group_by(ONTOLOGY) %>%
    slice_min(p.adjust, n = n_terms, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      term_label = str_wrap(Description, width = 32),
      
      # ---- REMOVED: gene_label generation ----
      
      ONTOLOGY = dplyr::recode(ONTOLOGY,
        "BP" = "Biological Process",
        "CC" = "Cellular Component",
        "MF" = "Molecular Function"
      )
    )
  
  plot_df$ONTOLOGY <- factor(plot_df$ONTOLOGY, 
                              levels = c("Biological Process", 
                                         "Cellular Component", 
                                         "Molecular Function"))
  
  ggplot(
    plot_df,
    aes(
      x = GeneRatio_num,
      y = reorder(term_label, GeneRatio_num),
      color = p.adjust,
      size = Count)) +
    geom_point(alpha = 0.6) +
    facet_grid(ONTOLOGY ~ ., scales = "free_y", space = "free_y") +
    scale_color_continuous(low = "red", high = "blue", trans = "reverse") +
    scale_size_continuous(name = "Number of\nMATR3 Targets") + # rename size legend
    labs(
      title = paste("Functional Enrichment of Matrin3", set_name, "genes"),
      x = "Ratio of MATR3 Targets to Total Genes in Term",
      y = NULL,
      color = "Adjusted\np-value") +
    theme_light(base_size = 19) +
    theme(
    strip.text   = element_text(size = 20, face = "bold"),
    axis.title = element_text(face = "bold", size = 20),
    axis.text.x = element_text(color = "black", size = 18),
    axis.text.y = element_text(color = "black", size = 20),
    legend.title = element_text(size = 18, face = "bold"),
    legend.text = element_text(size = 18),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.background = element_rect(fill = "grey20"),
    plot.margin = margin(10, 10, 10, 10))  # reduced right margin since no labels
  
}
```

## GO enrichment of genes containing MATR3-specific sites {.tabset}
```{r}
# Run GO enrichment separately for each ontology
ego_m3_specific <- enrichGO(
  gene          = unique(m3specific_site$GeneID),
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENSEMBL",
  ont           = "ALL",
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

# Build plots
ego_m3_specific_plot <- plot_go_enrichment(
  ego_obj   = ego_m3_specific,
  set_name  = "specific targets",
  n_terms   = 7
)

# # Save each ontology plot as PNG
ggsave(
  filename = file.path(
    "/Users/tinkienoob/Master_thesis/MATR3/Deseq2/GO_deseq2_png/GO_MATR3_specific_targets.png"),
  plot = ego_m3_specific_plot,
  width = 11,
  height = 13,
  dpi = 600
)
```

## Overlap with APPRIS Transcripts
APPRIS selects a single CDS variant for each gene as the 'PRINCIPAL' isoform based on the range of protein features: https://appris.bioinfo.cnio.es/#/downloads. 

Why APPRIS principal 1? A transcript is tagged PRINCIPAL:1 if it has the strongest structural evidence, regardless of what the other transcripts of the same gene are.
PRINCIPAL:1 = very strong evidence
PRINCIPAL:2 = strong
PRINCIPAL:3 = moderate
PRINCIPAL:4 = weak
PRINCIPAL:5 = very weak / default
But all PRINCIPAL:1–5 are still considered the “main isoform” for that gene.
```{r}
appris_anno <- read.table("/Users/tinkienoob/Master_thesis/MATR3/mm39_anno/appris_data_principal_m39.txt", 
                          header = FALSE, sep = "\t", stringsAsFactors = FALSE)

appris_anno <- appris_anno %>%
  rename(
    gene_name = V1,
    gene_id = V2,
    tx_id = V3,
    appris_class = V4,
    appris_tag = V5) 

appris_only_principal <- appris_anno %>%
  filter(grepl("PRINCIPAL", appris_tag)) 

```

* Overlap APPRIS principal transcripts with M3-specific targets
```{r}
# Get all transcripts with gene_id and tx_id for overlap
all_transcripts <- transcripts(annoDb, columns = c("tx_name", "gene_id")) 

# remove version suffix from tx_name to match APPRIS tx_id
all_transcripts$tx_id <- sub("\\..*", "", all_transcripts$tx_name)

overlap_hits <- GenomicRanges::findOverlaps(
  query = m3specific_site_gr, 
  subject = all_transcripts, 
  type = "any", 
  ignore.strand = FALSE
)

# Extract the specific transcript IDs that overlap your editing sites
m3spec_tx_ids <- all_transcripts$tx_id[subjectHits(overlap_hits)]
```

```{r}
## Get principal isoform 
principal_m3spec_sets <- list(
  Principal_isoforms = unique(appris_only_principal$tx_id),
  Matr3_specific_targets = unique(m3spec_tx_ids))

png(
  filename = "/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png/APPRIS_overlap_M3.png",
  width    = 8,
  height   = 5,
  units    = "in",
  res      = 600,
  bg       = "white")

plot(
  eulerr::euler(principal_m3spec_sets),
  
  fills = list(
    fill  = c("grey", "#2171B5"),
    alpha = 0.5),
  
  labels = list(
    fontsize = 18,
    fontface = "bold"),
  
  quantities = list(
    fontsize = 18,
    fontface = "bold",
    col      = "black"),
  
  stroke_size = 1.5)

dev.off()

plot(
  eulerr::euler(principal_m3spec_sets),
  
  fills = list(
    fill  = c("grey", "#2171B5"),
    alpha = 0.5),
  
  labels = list(
    fontsize = 18,
    fontface = "bold"),
  
  quantities = list(
    fontsize = 18,
    fontface = "bold",
    col      = "black"),
  
  stroke_size = 1.5)
```

```{r}
# Overlapping genes between APPRIS principal genes and MATR3-specific targets
overlap_principal_tx <- intersect(
  appris_only_principal$tx_id,
  unique(m3spec_tx_ids))

# Count how many overlapping genes belong to p:1, p:2, p:3, p:4
overlap_appris_count <- appris_only_principal %>%
  filter(tx_id %in% overlap_principal_tx) %>%
  mutate(
    principal_group = case_when(
      str_detect(appris_tag, "PRINCIPAL:1") ~ "P:1",
      str_detect(appris_tag, "PRINCIPAL:2") ~ "P:2",
      str_detect(appris_tag, "PRINCIPAL:3") ~ "P:3",
      str_detect(appris_tag, "PRINCIPAL:4") ~ "P:4",
      TRUE ~ "other")) %>%
  filter(principal_group %in% c("P:1", "P:2", "P:3", "P:4")) %>%
  dplyr::distinct(tx_id, principal_group) %>%   # avoid counting same gene twice
  dplyr::count(principal_group)

overlap_appris_count

overlap_principal_genes <- appris_only_principal %>%
  filter(tx_id %in% overlap_principal_tx)

overlap_appris_count <- overlap_principal_genes %>%
  mutate(
    principal_group = case_when(
      str_detect(appris_tag, "PRINCIPAL:1") ~ "P:1",
      str_detect(appris_tag, "PRINCIPAL:2") ~ "P:2",
      str_detect(appris_tag, "PRINCIPAL:3") ~ "P:3",
      str_detect(appris_tag, "PRINCIPAL:4") ~ "P:4",
      TRUE ~ "other"
    )
  ) %>%
  filter(principal_group %in% c("P:1", "P:2", "P:3", "P:4")) %>%
  dplyr::distinct(tx_id, principal_group) %>%  
  dplyr::count(principal_group)

# Plot the distribution
appris_dis <- ggplot(overlap_appris_count, aes(x = principal_group, y = n, fill = principal_group)) +
  geom_col(width = 0.7, color = "black", size = 0.3) +
  geom_text(aes(label = n), vjust = -0.5, size = 5, fontface = "bold") +
  scale_x_discrete(limits = c("P:1", "P:2", "P:3", "P:4")) +
  scale_fill_brewer(palette = "Blues") + # A clean palette for thesis figures
  labs(
    title = "",
    x = "",
    y = "Number of transcripts"
  ) +
  theme_thesis() +
  coord_cartesian(ylim = c(0, max(overlap_appris_count$n) * 1.15)) + # Dynamically scale Y-axis
  theme(legend.position = "none") # Legend is redundant here

save_thesis_plot(
  appris_dis,
  filename = "/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png/APPRIS_pc_distribution.png",
  width = 6, height = 6
)

appris_dis
```


```{r}
# appris_dis <- ggplot(overlap_appris_count, aes(x = principal_group, y = n, fill = principal_group)) +
#   geom_col(width = 0.7) +
#   geom_text(aes(label = n), vjust = -0.3, size = 5, fontface = "bold") +
#   scale_x_discrete(limits = c("P:1", "P:2", "P:3", "P:4")) +
#   labs(
#     title = "",
#     x = "",
#     y = "Number of transcripts") +
#   theme_bw(base_size = 18) +
#   theme_thesis() +
#   coord_cartesian(ylim = c(0, 8500))
# 
# # save_thesis_plot(
# #   appris_dis,
# #   filename = file.path("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
# #                        "APPRIS_pc_distribution.png"),
# #   width = 6,
# #   height = 6,
# #   dpi = 300,
# #   bg = "white")
# appris_dis
```

```{r region-length-principal-isoforms}
# Claude generated
# Re-derive region lengths using only APPRIS principal isoform transcripts
# from overlap_principal_tx (transcript-level, not gene-level union across all isoforms).
# Produces region_length_df_principal in the same long format as region_length_df.

# Strip version suffix helper
strip_ver <- function(x) sub("\\..*", "", x)

# Get per-transcript lengths from a GRangesList (indexed by tx_name with version)
get_tx_lengths <- function(grl, keep_tx_ids) {
  keep_idx <- strip_ver(names(grl)) %in% keep_tx_ids
  grl_filt <- grl[keep_idx]
  lens     <- sum(width(grl_filt))
  data.frame(
    tx_id_stripped = strip_ver(names(lens)),
    length_bp      = as.numeric(lens),
    stringsAsFactors = FALSE
  )
}

# tx2gene with version-stripped tx_id for joining
tx2gene_stripped <- tx2gene %>%
  mutate(tx_id_stripped = strip_ver(tx_name))

# Attach gene_id and keep the longest principal tx per gene
to_gene_level <- function(df) {
  df %>%
    left_join(tx2gene_stripped %>% select(tx_id_stripped, gene_id),
              by = "tx_id_stripped") %>%
    filter(!is.na(gene_id)) %>%
    group_by(gene_id) %>%
    slice_max(length_bp, n = 1, with_ties = FALSE) %>%
    ungroup()
}

utr5_pr   <- get_tx_lengths(utr5_by_gene,   overlap_principal_tx) %>% to_gene_level()
utr3_pr   <- get_tx_lengths(utr3_by_gene,   overlap_principal_tx) %>% to_gene_level()
intron_pr <- get_tx_lengths(intron_by_gene, overlap_principal_tx) %>% to_gene_level()

# CDS is gene-level (cdsBy by="gene"); filter to principal genes
principal_gene_ids <- appris_only_principal %>%
  filter(tx_id %in% overlap_principal_tx) %>%
  pull(gene_id) %>%
  unique()

all_pr_genes <- unique(c(utr5_pr$gene_id, utr3_pr$gene_id,
                          intron_pr$gene_id, principal_gene_ids))

region_length_df_principal <- data.frame(
  gene_id = all_pr_genes,
  UTR5    = utr5_pr$length_bp[match(all_pr_genes,   utr5_pr$gene_id)],
  CDS     = as.numeric(cds_len[all_pr_genes]),
  UTR3    = utr3_pr$length_bp[match(all_pr_genes,   utr3_pr$gene_id)],
  Intron  = intron_pr$length_bp[match(all_pr_genes, intron_pr$gene_id)],
  stringsAsFactors = FALSE
) %>%
  replace(is.na(.), 0) %>%
  pivot_longer(cols = c(UTR5, CDS, UTR3, Intron),
               names_to  = "region",
               values_to = "region_length_kb") %>%
  mutate(region_length_kb = region_length_kb / 1000) %>%
  filter(region_length_kb > 0)
```


# Define different metrics to evaluate the top targets
## Transcript length (from annotation)
```{r}
# Make a gene length df from gns annotation
gene_length_df <- data.frame(
  gene_id = names(gns),
  gene_length_kb = width(gns)/ 1000 
)

# Add gene length info into main df: 
m3specific_site <- m3specific_site %>%
  left_join(gene_length_df, by = "gene_id") %>%
  mutate(GeneID = sub("\\..*", "", gene_id))

# Get transcript length from range of transcripts
transcripts_id <- transcripts(annoDb, columns = c("tx_name", "gene_id"))
seqlevelsStyle(transcripts_id) <- "UCSC"

transcripts_df <- as.data.frame(transcripts_id)

transcripts_df$gene_id <- as.character(unlist(transcripts_df$gene_id))

transcripts_df <- transcripts_df %>%
  mutate(tx_length = width)  # width is GRanges width

# Keep the longest transcript per gene
longest_transcripts <- transcripts_df %>%
  group_by(gene_id) %>%
  slice_max(order_by = tx_length, n = 1) %>%  # only choose longest transcript per 1 gene
  ungroup() %>% 
  mutate(GeneID = sub("\\..*", "", gene_id)) # remove version suffix from gene_id
```

## Transcript length distribution
```{r}
ggplot(m3specific_site, aes(x = gene_length_kb)) +
  geom_histogram(
    binwidth = 50, 
    fill = "#4C78A8", 
    color = "black", 
    alpha = 0.7) +
  theme_light(base_size = 14) +
  labs(
    title = "Transcript Length of MATR3-specific targets Distribution",
    x = "Transcript Length (kb)",
    y = "Number of transcripts") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

## The longest high confidence transcripts
```{r eval = FALSE}
# remove suffix
longest_transcripts <- longest_transcripts %>%
  mutate(tx_id = sub("\\..*", "", tx_name))

longest_highconf_sets <- list(
  Longest_Transcripts = unique(longest_transcripts$tx_id),
  Matr3_specific_targets = unique(m3spec_tx_ids))

plot(eulerr::euler(longest_highconf_sets),
       fills = list(fill = c("#BD241EFF", "#F5B355"), alpha = 0.5),
       quantities = TRUE,
       stroke_size = 1, # control border size
       set_name_size = 1, # set name size
       text_size = 2)

######################## collect longest high confidence transcripts 
longest_highconf_tx <- intersect(unique(longest_transcripts$gene_id),
                                 unique(m3spec_tx_ids)

```



```{r}
######################## collect principal high confidence transcripts 
principal_m3_tx <- intersect(unique(appris_only_principal$gene_id), 
                                   unique(m3specific_site$GeneID))

principal_m3_tx_df <- m3specific_site %>%
  group_by(GeneID) %>%
  filter(GeneID %in% principal_m3_tx)

# saveRDS(principal_m3_tx_df, file = "/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/matr3_principal_iso.rds")
```


## Number of editing sites in each gene
```{r}
site_counts_df <- m3specific_site %>%
  group_by(gene_id) %>%
  summarise(
    n_sites = n(),
   .groups = "drop"  # removes the extra message
  ) %>%
  left_join(gene_length_df %>% 
            dplyr::select(gene_id, gene_length_kb) %>%
            distinct(),
            by = "gene_id") %>%
  arrange(desc(n_sites))

ggplot(site_counts_df, aes(n_sites)) +
  geom_histogram(binwidth = 2,
                 fill = "#4C78A8", 
                color = "black", 
                alpha = 0.7) +
  labs(
    title = "Number of editing sites/ a gene",
    x = "Number of editing sites",
    y = "Number of genes") +
  coord_cartesian(xlim = c(0,50)) +
  theme_light(base_size = 14)
```


## Editing site density
now I have one longest transcript per gene, I use gene length (or the summed exon length of the gene) to normalize editing sites.
```{r}
# Calculate number of editing sites per longest transcript/ gene
sites_per_kb_df <- site_counts_df %>%
  mutate(sites_per_kb = n_sites / gene_length_kb) %>%
  filter(!is.na(sites_per_kb), gene_length_kb > 0)

ggplot(sites_per_kb_df, aes(x = sites_per_kb)) +
  geom_histogram(
    fill = "#4C78A8",
    color = "black",
    alpha = 0.7) +
  theme_light(base_size = 14) +
  labs(
    x = "Number of sites/ kb gene length",
    y = "Number of genes",
    title = "Editing site density") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```
## Editing scores calculation: mean, median, and max score
```{r}
calc_edit_scores <- m3specific_site %>%
  group_by(gene_id) %>%
  summarise(
    mean_editing   = mean(score, na.rm = TRUE),
    median_editing = median(score, na.rm = TRUE),
    max_editing    = max(score, na.rm = TRUE),
    .groups = "drop"
  )
```

### Median editing % per site
```{r}
# Histogram for median editing %
median_per <- ggplot(calc_edit_scores, aes(x = median_editing)) +
  geom_histogram(binwidth = 2,
                 fill = "#4C78A8", 
                color = "black", 
                alpha = 0.7) +
  theme_light() +
  labs(
    title = "Histogram of Median Editing %",
    x = "Median Editing %",
    y = "Number of genes") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(color = "black", size = 17),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 15),
    legend.position = "right"
  )

median_per

# save_thesis_plot(
#   median_per,
#   filename = file.path("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
#                        "Median_editing_distribution.png"),
#   width = 6,
#   height = 6,
#   dpi = 300,
#   bg = "white")
```

### Mean editing % per site
```{r}
# Histogram for mean editing %
mean_per <- ggplot(calc_edit_scores, aes(x = mean_editing)) +
  geom_histogram(binwidth = 2,
                 fill = "#4C78A8", 
                color = "black", 
                alpha = 0.7) +
  theme_light() +
  labs(
    title = "Histogram of Mean Editing %",
    x = "Mean Editing %",
    y = "Number of genes") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(color = "black", size = 17),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 15),
    legend.position = "right"
  )

mean_per

# save_thesis_plot(
#   mean_per,
#   filename = file.path("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
#                        "Mean_editing_distribution.png"),
#   width = 6,
#   height = 6,
#   dpi = 300,
#   bg = "white")
```

### Max Editing % per gene
```{r}
max_per <- ggplot(calc_edit_scores, aes(x = max_editing)) +
  geom_histogram(binwidth = 2,
                 fill = "#4C78A8", 
                color = "black", 
                alpha = 0.7) +
  theme_light() + 
  labs(
    title = "Histogram of Max Editing %",
    x = "Max Editing %",
    y = "Number of genes") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(color = "black", size = 17),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 15),
    legend.position = "right"
  )

max_per

# save_thesis_plot(
#   max_per,
#   filename = file.path("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
#                        "Max_editing_distribution.png"),
#   width = 6,
#   height = 6,
#   dpi = 300,
#   bg = "white")
```

## Expression level computed by TPM
TPM (Transcripts Per Million) is a within-sample measure of relative transcript abundance.
If a transcript has many editing sites but also very high TPM, that might be due to transcript high expression, not because MATR3 binds it more specifically.
* Input htseq count 
```{r htseq-input, echo=FALSE, message=FALSE, warning=FALSE}
# Input path
countFolder_HT <- "/Users/tinkienoob/Master_thesis/MATR3/htseqcount_HT_nonreverse"

list.files(countFolder_HT)

# List all file containing "counts" in the folder
countFiles_HT <- list.files(countFolder_HT, pattern = "htseq_forward\\.txt$", full.names = TRUE)

# Create sample table/ data frame for DESeq from Htseq-count
## remove prefix and suffix in htseq output name
sampleNames_HT <- gsub(".*/Silke_|_dedup_htseq_forward\\.txt$", "", countFiles_HT)

## define how many conditions and replicates for df
conditions_HT = factor(rep(c("ADARcontrol", "MATR3", "ΔRRMcontrol"), each=4))
replicates_HT = factor(rep(c("rep1", "rep2", "rep3", "rep4"),3))

sampleTable_HT <- data.frame(
  sampleName = sampleNames_HT,
  file = basename(countFiles_HT),
  condition = conditions_HT,
  replicate = replicates_HT
)

####################### Create the DESeqDataSet (without matrix)
dds_HT <- DESeqDataSetFromHTSeqCount(sampleTable = sampleTable_HT,
                                 directory = countFolder_HT,
                                 design = ~ condition)

####################### Running DESeq2 analysis to identify differentially expressed genes
# Deseq data object 
dds_HT <- DESeq(dds_HT)

# Save dds outout in the outpath as rds file
# saveRDS(dds_HT, file = paste0(outpath, "Hypertribe_DESeq2.rds"))
```

* Get normalized read counts and calculate DESeq2 results
```{r}
dds_HT <- readRDS(paste0(outpath, "Hypertribe_DESeq2.rds"))
# Inspect the sizeFactor
dds_HT$sizeFactor

# Get read counts
#A = DESeq2::counts(dds)
#rownames(A) = sub("\\..*", "", rownames(A))

#Get normalized read counts of dds result
dds_count_HT <- DESeq2::counts(dds_HT, normalized = TRUE)

# calculate dds result 
res_HT <- results(dds_HT, alpha = 0.05)

### Information about which variables and tests were used can be found by calling the function mcols on the results object.
#mcols(res_HT)$description
```

* Convert dds result into dataframe and add gene_id column for later merging with annotation
```{r}
# Convert dds result into dataframe
res_HT.df <- as.data.frame(res_HT) %>%
  mutate(GeneID = sub("\\..*", "", rownames(res_HT))) # remove version suffix from gene_id
```

Source - https://stackoverflow.com/q
sweep(x, MARGIN, STATS, FUN="-", check.margin=T, ...)
x is the data
STATS refers to the summary statistics which you wish to 'sweep out'
FUN is the function used to carry out the sweep, "-" is the default

```{r}
################################ get raw count from dds
counts_raw_HT <- counts(dds_HT, normalized = FALSE)

counts_HT_df <- as.data.frame(counts_raw_HT) %>%
  tibble::rownames_to_column("gene_id") %>% 
  mutate(GeneID = sub("\\..*", "", gene_id))

counts_HT_df <- counts_HT_df %>%
  left_join(gene_length_df, by = "gene_id")

calculateTPM <- function(df) {
  # only keep raw counts column
  count_cols <- setdiff(colnames(df), c("gene_id", "GeneID", "gene_length_kb"))
  
  # Step 1: Compute RPK per sample
  df_rpk <- df
  df_rpk[, count_cols] <- df_rpk[, count_cols] / (df_rpk$gene_length_kb)
  
  # Step 2: Compute scaling factors per sample
  scaling_factors <- colSums(df_rpk[, count_cols], na.rm = TRUE) / 1e6 # Calculate column-wise sum rpk/1e6
  
  # Step 3: Compute TPM
  tpm_matrix <- sweep(df_rpk[, count_cols], 2, scaling_factors, "/") # TPM[i,j] = RPK[i,j] / (sum(RPK[,j])/ 1e6 )
  
  # Step 4: Add gene_id back
  tpm_df <- cbind(gene_id = df$gene_id, as.data.frame(tpm_matrix))
  
  return(tpm_df)
}

TPM_calc <- calculateTPM(counts_HT_df) 

TPM_df <- TPM_calc %>%
  mutate(
    Control_meanTPM = rowMeans(dplyr::select(., HT_1:HT_4), na.rm = TRUE),
    WT_meanTPM = rowMeans(dplyr::select(., M3_HT_1:M3_HT_4), na.rm = TRUE),
    KD_meanTPM = rowMeans(dplyr::select(., M3d_HT_1:M3d_HT_4), na.rm = TRUE),
    log10_TPM_WT = log10(WT_meanTPM)) %>%
  dplyr::select(c("gene_id", "Control_meanTPM", "WT_meanTPM", "KD_meanTPM", "log10_TPM_WT")) %>%
  left_join(counts_HT_df %>% dplyr::select(gene_id, GeneID, gene_length_kb), by = "gene_id")
```


# Create a table with different features 
```{r}
# calculate fraction of sites
fraction_cal <- m3specific_site %>%
  group_by(gene_name, gene_id) %>%
  summarise(
    n_sites = n(),
    sites_over10 = sum(score > 10, na.rm = TRUE),
    fraction_over10 = round((sites_over10 / n_sites), 2) * 100,
    .groups = "drop"  # ADD THIS
  )

features_df <- site_counts_df %>%
  left_join(calc_edit_scores, by = "gene_id") %>% # add editing scores
  left_join(TPM_df %>% dplyr::select(gene_id, WT_meanTPM, log10_TPM_WT), by = "gene_id") %>% # add TPM and log10(TPM) 
  left_join(gene_anno_df %>% dplyr::select(gene_id, gene_name), by = "gene_id")  # add gene name

```

## Expression level distribution and cutoff
```{r}
TPM_cutoff <- ggplot(features_df, aes(x = log10_TPM_WT)) +
  geom_histogram(binwidth = 0.1, fill = "#4C78A8", color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + # Cutoff example
  labs(title = "Distribution of log10(TPM) Expression Levels",
       x = "log10(TPM)",
       y = "Number of transcripts") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(color = "black", size = 14, face = "bold"),
    axis.text.y = element_text(size = 14)
  )

TPM_cutoff

# save_thesis_plot(
#   TPM_cutoff,
#   filename = file.path("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
#                        "TPM_cutoff_distribution.png"),
#   width = 6.5,
#   height = 4,
#   dpi = 600,
#   bg = "white")
```
Why log10(TPM) > 0? This means TPM > 1, which is a common threshold to consider a gene as "expressed" in the sample.

### Filter out lowly expressed transcripts (log10(TPM) <= -1) to focus on more reliable editing events
```{r}
features_df2 <- features_df %>%
  filter(log10_TPM_WT > 0)

lowexpress_genes <- features_df %>%
  filter(log10_TPM_WT < 0)
```

# Scatterplot between gene length and max editing %
Editing % reflects how efficiently ADAR edits the RNA at that site, which correlates with how stable or strong the RBP–RNA interaction is => Here I would call it editing efficiency 
High editing % → strong or repeated RBP binding
Low editing % → weak, transient, or noise.

Mean while editing density (#sites per kb) tells you how many contacts occur.
filter(log10_TPM_WT > 0) is equivalent to filter(WT_meanTPM > 1)

```{r}
maxscore_length <- ggplot(features_df2, aes(x = gene_length_kb, y = max_editing)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE) +
  stat_cor(method = "pearson", size = 6) +
  theme_bw() +
  labs(
   title = "",
    x = "Gene length (kb)",
    y = "Maximum Editing %") + 
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(color = "black", size = 16),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16))+
  coord_cartesian(xlim = c(0, 500), ylim = c(0, 100))

maxscore_length

save_thesis_plot(
  maxscore_length,
  filename = file.path("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
                       "Max_editing_gene_length.png"),
  width = 6,
  height = 6,
  dpi = 300,
  bg = "white")
# 
```
The cloud is very dense and scattered ("football" shape). The points are not tightly hugging the line. This visually suggests a weaker correlation than the site count plot, even though the line goes up.

# Scatterplot between Max Editing %/ Editing sites
```{r}
maxscore_count <- ggplot(features_df2, aes(x = n_sites, y = max_editing)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", color = "blue", fill = "lightblue", alpha = 0.2) +
  theme_bw() + 
  stat_cor(method = "pearson", size = 6) +
  coord_cartesian(xlim = c(0, 90),
                  ylim = c(0, 100)) +
  labs(
    title = "",
    x = "Number of Editing sites",
    y = "Maximum Editing %") +
    theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(color = "black", size = 16),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16)
  )

maxscore_count

save_thesis_plot(
  maxscore_count,
  filename = file.path("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
                       "Max_editing_site_count.png"),
  width = 6,
  height = 6,
  dpi = 300,
  bg = "white")

```

# Scatterplot between Expression level and Numbers of editing site
```{r}
#| echo: true
tpm_count <- ggplot(features_df2, aes(x = log10_TPM_WT, y = n_sites)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_bw() +
  labs(
    title = "",
    x = "log10(TPM)",
    y = "Number of Editing Sites") +
  stat_cor(method = "pearson", size = 6) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(color = "black", size = 16),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16)
  )

save_thesis_plot(
  tpm_count,
  filename = file.path("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
                       "tpm_count.png"),
  width = 6,
  height = 6,
  dpi = 300,
  bg = "white")

tpm_count
```

# Scatterplot between Expression level and Max editing %
```{r}
tpm_maxedit <- ggplot(features_df2, aes(x = log10_TPM_WT, y = max_editing)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "",
    x = "log10(TPM)",
    y = "Maximum Editing %") +
  theme_bw() + 
  stat_cor(method = "pearson", size = 6) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(color = "black", size = 16),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16)
  )

save_thesis_plot(
  tpm_maxedit,
  filename = file.path("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
                       "tpm_maxedit.png"),
  width = 6,
  height = 6,
  dpi = 300,
  bg = "white")

tpm_maxedit
```
There is no positive correlation between gene expression and editing efficiency.
=> Being a highly expressed gene does not make it more likely to be edited or high-efficiency editing.

## Scatterplot between Editing sites/ gene length (kb)
```{r}
length_count_w_outlier <- ggplot(features_df2, aes(x = gene_length_kb, y = n_sites)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  # geom_point(data = subset(features_df2, features_df2$gene_length_kb >= quantile(features_df2$gene_length_kb, 0.995)), color = "red", size = 3, alpha = 0.4) + # Highlight outliers in red
  # geom_point(data = subset(features_df2, features_df2$n_sites >= quantile(features_df2$n_sites, 0.995)), color = "red", size = 3, alpha = 0.4) + # Highlight outliers in red
  theme_bw() +
  labs(
    title = "BEFORE outlier removal",
    x = "Gene Length (kb)",
    y = "Number of Editing Sites") +
  stat_cor(method = "pearson", size = 6) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(color = "black", size = 16),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16)
  )

length_count_w_outlier

# save_thesis_plot(
#   length_count_wth_outlier,
#   filename = file.path("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
#                        "Gene_length_count_before_removal.png"),
#   width = 6,
#   height = 6,
#   dpi = 300,
#   bg = "white")
```

# Correcting for Size to Find True Biological Signal
### Set outlier thresholds
```{r}

# Check how many outliers at different thresholds
threshold_comparison <- data.frame()

for(q in c(0.9, 0.95, 0.99, 0.995, 1.0)){
  
  length_t <- quantile(features_df2$gene_length_kb, q)
  editing_t <- quantile(features_df2$n_sites, q)
  
  outliers <- features_df2 %>%
    filter(gene_length_kb >= length_t | n_sites >= editing_t)
  
  clean <- features_df2 %>%
    filter(gene_length_kb < length_t & n_sites < editing_t)
  
  # Fit model on clean data
  model_clean <- lm(n_sites ~ gene_length_kb, data = clean)
  r2 <- summary(model_clean)$r.squared
  
  threshold_comparison <- rbind(threshold_comparison, data.frame(
    quantile = q,
    length_cutoff_kb = round(length_t, 1),
    editing_cutoff = round(editing_t, 0),
    n_outliers = nrow(outliers),
    pct_removed = round(nrow(outliers)/nrow(features_df2)*100, 2),
    r_squared = round(r2, 4)
  ))
}

print(threshold_comparison)

# plot_threshold <- function(q) {
#   length_t  <- quantile(features_df2$gene_length_kb, q, na.rm = TRUE)
#   editing_t <- quantile(features_df2$n_sites, q, na.rm = TRUE)
#   
#   clean <- features_df2 %>%
#     filter(gene_length_kb < length_t & n_sites < editing_t)
#   
#   ggplot(clean, aes(x = gene_length_kb, y = n_sites)) +
#     geom_point(alpha = 0.6) +
#     geom_smooth(method = "lm", se = TRUE) +
#     ggpubr::stat_cor(method = "pearson", size = 5) +
#     theme_bw() +
#     labs(
#       title = paste("After outlier removal: q =", q),
#       x = "Gene length (kb)",
#       y = "Number of editing sites"
#     ) +
#     theme(
#       plot.title = element_text(hjust = 0.5, face = "bold"),
#       axis.title = element_text(face = "bold")
#     )
# }

```
Why don't high editing site counts mean that transcripts are more likely to be potential targets of Matr3? because numbers of editing site are driven by length of genes (the longer the genes, the more opportunities that ADAR enzyme can cause more editing)

```{r}
# Define 99.5th percentile thresholds
length_threshold  <- quantile(features_df2$gene_length_kb, 0.995)
editing_threshold <- quantile(features_df2$n_sites, 0.995)

ggplot(features_df2, aes(x = gene_length_kb, y = n_sites)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_point(data = subset(features_df2, features_df2$gene_length_kb >= length_threshold),
             color = "red", size = 3, alpha = 0.2) + # Highlight outliers in red
  geom_point(data = subset(features_df2, features_df2$n_sites >= editing_threshold),
             color = "red", size = 3, alpha = 0.2) + # Highlight outliers in red
  labs(
    title = "AFTER outlier removal",
    x = "Gene Length (kb)",
    y = "Number of Editing Sites") +
  stat_cor(method = "pearson", size = 6) +
  theme_bw() + 
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(color = "black", size = 16),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16)
  )

```
* Fit linear model to predict expected number of editing sites based on gene length
```{r}
model_len_sites <- lm(n_sites ~ gene_length_kb, data = features_df2)
# out: expected sites = 1.22061 + (0.06854 x gene length in kb)

features_df3 <- features_df2 %>%
  mutate(
    expected_sites = predict(model_len_sites, newdata = .),
    # Residual = Actual Length - Expected Length
    residuals = n_sites - expected_sites
  ) %>%
  dplyr::select(c("gene_id", "gene_name", "n_sites", "gene_length_kb", "max_editing", "residuals")) %>%
  arrange(desc(residuals))

#view(top_target_df2)
################################## Create new df without outliers for residual values
upper_outliers <- features_df3 %>%
  filter(
    gene_length_kb >= length_threshold |
    n_sites        >= editing_threshold, 
    residuals > 0) %>%
  mutate(outlier_type = "top")

# Collect bottom outliers with extremely low editing site density
lower_outliers <- features_df3 %>%
  filter(gene_length_kb >= length_threshold,
         residuals < 0) %>%
  mutate(outlier_type = "bot")
  
# All outliers with high editing site counts and low editing site counts but long gene length
outliers_len_sites <- bind_rows(upper_outliers, lower_outliers)

# Collect potential transcripts having high density editing sites
upper_targets <- features_df3 %>%
  filter(residuals > 0) %>% 
  arrange(desc(n_sites))

# Long transcripts below the regression line which have low density editing sites
lower_targets <- features_df3 %>%
  filter(residuals <= 0) %>%
  arrange(desc(n_sites))
```
The "big cloud" in the bottom-left corner (short genes, few sites) represents the vast majority of the transcriptome. It is tempting to dismiss them as "background," but they often contain highly specific.


* Remove outliers and re-plot the scatterplot to see the true relationship between gene length and editing site number
```{r}
#| echo: true
# Make scatterplot between gene length and editing site number WITHOUT outliers
m3target_no_outlier_df <- features_df2 %>%
  filter(gene_length_kb < length_threshold & n_sites < editing_threshold)

length_count_no_outliers <- ggplot(m3target_no_outlier_df,
       aes(x = gene_length_kb, y = n_sites)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  theme_bw() +
  labs(
    title = "AFTER outlier removal",
    x = "Gene Length (kb)",
    y = "Number of Editing Sites") +
  stat_cor(method = "pearson", size = 6) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(color = "black", size = 16),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16)
  )

# save_thesis_plot(
#   length_count_no_outliers,
#   filename = file.path("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
#                        "genelng_count_no_outliers.png"),
#   width = 6,
#   height = 6,
#   dpi = 300,
#   bg = "white")

length_count_no_outliers
```
Top outliners = more highly expressed than predicted but poorly edited (because ADAR enzyme is probably less active on them)
Bot outliner = strongly edited eventhough low expression
99.5th percentile cutoff means:

upper_outliers = genes with extremely HIGH n_sites 
               OR extremely LONG transcripts
               → These are genuine MATR3 targets 
                 but distort regression line
               → Should go to Tier 1 or Tier 2 
                 based on max_editing

lower_outliers = genes with extremely LOW editing density
               → Very long transcripts with almost no sites
               → Should go to Tier 3
               
* Fit linear model on the clean data without outliers to get expected number of editing sites based on gene length, then calculate residuals to find transcripts with higher or lower editing site density than expected
```{r}
clean_model_len_sites <- lm(n_sites ~ gene_length_kb, data = m3target_no_outlier_df)

# Apply the model to the FULL dataset (assuming 'features_df2' is your full dataframe)
tier_classify_df <- m3target_no_outlier_df %>%
  mutate(
    expected_sites = predict(clean_model_len_sites, newdata = m3target_no_outlier_df), # Get expected number of sites based on gene length using the clean model
    
    # Residual = Actual Sites - Expected Sites
    # Positive = denser editing than expected; Negative = sparser editing than expected
    residuals = n_sites - expected_sites) %>%
  mutate(
    # Use case_when to cleanly assign tiers based on your established criteria
    rank = case_when(
      residuals > 0 & max_editing >= 50 ~ "Tier 1",
      residuals > 0 & max_editing < 50  ~ "Tier 2",
      residuals <= 0 ~ "Tier 3",
      TRUE ~ "Others")) %>%
  dplyr::select(gene_id, gene_name, n_sites, gene_length_kb, max_editing, expected_sites, residuals, rank) %>%
  arrange(desc(residuals))

upper_outliers_classified <- upper_outliers %>%
  mutate(
    rank = case_when(
      max_editing >= 50 ~ "Tier 1",
      max_editing < 50  ~ "Tier 2"))

lower_outliers_classified <- lower_outliers %>%
  mutate(rank = "Tier 3")

tier_classify_df <- bind_rows(
  tier_classify_df,
  upper_outliers_classified,
  lower_outliers_classified) %>%
  distinct(gene_id, .keep_all = TRUE)


# Check your distributions to ensure the logic worked as intended
table(tier_classify_df$rank)
```
in short, 
residuals > 0 & max_editing >= 50 ~ "Tier1"
residuals > 0 & max_editing < 50  ~ "Tier2"
residuals < 0 ~ "Tier3"

# Manually move Ttn gene from Tier1 into Tier2 set
```{r}
# Manually assign the Ttn gene to Tier 1, overriding the model
rank_M3tx_manual_df <- tier_classify_df %>%
  mutate(
    rank = ifelse(gene_name == "Ttn", "Tier 1", rank)) 

# Check the distribution to make sure Ttn shifted correctly
table(rank_M3tx_manual_df$rank)

saveRDS(rank_M3tx_manual_df, file = "/Users/tinkienoob/Master_thesis/MATR3/rank_M3_target_manual.rds")
```
* Visualize the relationship between gene length and editing site number colored by tier to see how the 3 tiers distribute in the plot

## Scatterplot between length-adjusted residual and max editing % colored by tier to see how the 3 tiers distribute in the plot
```{r}
#| echo: true

tier_colors <- c(
  "Tier 1" = "#CD534C", 
  "Tier 2" = "#EFC050", 
  "Tier 3" = "#45B7D1",
  "Others" = "grey80")

residual_tier <- ggplot(rank_M3tx_manual_df, aes(x = residuals, y = max_editing, color = rank)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_hline(yintercept = 50, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = tier_colors) +
  stat_quadrant_counts(xintercept = 0, yintercept = 50,
                       size = 6) +
  theme_bw() +
  labs(
    title = "",
    x = "Residual (=Actual site counts - Expected site counts)",
    y = "Maximum Editing %",
    color = "Tier") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(color = "black", size = 16),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16),
    legend.position = "right")
```


```{r}
ggplot(tier_classify_df, aes(x = gene_length_kb, y = n_sites, color = rank)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(values = tier_colors) +
  theme_bw() +
  labs(
    title = "",
    x = "Gene length (kb)",
    y = "Number of editing sites",
    color = "Tier") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(color = "black", size = 16),
    axis.text.y = element_text(color = "black", size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16),
    legend.position = "right")

# save with transparent background
save_thesis_plot(
  residual_tier,
  filename = file.path("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
                       "residual_tier.png"),
  width = 7,
  height = 6,
  dpi = 300,
  bg = "white")

residual_tier
```



# Convert into fasta file for XSTREME analysis
getSeq() then tries to extract sequences for all of them at once, which can exceed R’s recursion/stack limits.
```{r eval = FALSE}
library(Rsamtools)

# genome FASTA (downloaded from GENCODE/Ensembl)
genome_fa_path <- "/Users/tinkienoob/Master_thesis/MATR3/mm39_anno/GRCm39.primary_assembly.genome.fa"

# Create FaFile object for the genome FASTA
fa <- FaFile(genome_fa_path)

fasta_conversion <- function(reprod_gr, prefix, fa){
  
  reprod_gr_resize <- GenomicRanges::resize(reprod_gr, width = 501, fix = "center")

  reprod_seqs <- getSeq(fa, reprod_gr_resize)
  
    names(reprod_seqs) <- !duplicated(paste0(
    prefix, "_",
    seqnames(reprod_gr_resize), ":",
    start(reprod_gr_resize), "-",
    end(reprod_gr_resize), "_",
    strand(reprod_gr_resize), "_",
    seq_along(reprod_gr_resize)
  ))

  writeXStringSet(reprod_seqs, filepath = paste0(prefix, "_es.fasta"))
}

tier1_gr <- m3specific_site_gr_anno[mcols(m3specific_site_gr_anno)$gene_id %in% tier1_tx$gene_id]

# Load control gr
specificADAR_control <- readRDS("/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/specificADAR_control.rds")
specificΔRRM_control <- readRDS("/Users/tinkienoob/Master_thesis/MATR3/Hypertribe_bindingsites_analysis/specificΔRRM_control.rds")

#### tier 1 is high confidence
fasta_conversion(tier1_gr, "tier1", fa)
fasta_conversion(specificADAR_control, "specificADAR_control", fa)
fasta_conversion(specificΔRRM_control, "specificDRRM_control", fa)
```


# Region distribution of editing sites with max editing percentage
```{r}
# collect editing sites with max editing percentage per gene then add region annotation
max_site_df <- as.data.frame(m3specific_site_gr_anno) %>%
  group_by(gene_id) %>%
  slice_max(order_by = score, n = 1, with_ties = FALSE) %>%
  ungroup() 

# add genomic region annotation to the max editing sites
max_site_region_counts <- max_site_df %>%
  distinct(gene_id, region, .keep_all = TRUE) %>% 
  mutate(region = ifelse(is.na(region), "Others", region)) %>%
  group_by(region) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  mutate(percentage = count / sum(count) * 100,
         class = "M3-specific gene")
```


# Region distribution per max editing sites per tier
* For each tier, count the number of max editing sites that fall into each genomic region category, then calculate the percentage distribution of regions within each tier. This will show if certain tiers are enriched for specific regions (e.g., Tier 1 having more CDS sites) compared to others.
```{r}
# Count genomic regions per tier
############################################################
tier_region_maxedit_df <- max_site_df %>%
  left_join(
    rank_M3tx_manual_df %>% dplyr::select(-c(gene_name, gene_length_kb)),
    by = "gene_id") %>%
  mutate(
    gene_id = as.character(gene_id),
    rank    = as.character(rank),
    region  = as.character(region),
    region  = dplyr::if_else(is.na(region) | region == "", "Others", region)
  ) %>%
  filter(!is.na(rank))

tier_region_maxedit_counts <- tier_region_maxedit_df %>%
  dplyr::count(rank, region, name = "count") %>%
  dplyr::group_by(rank) %>%
  dplyr::mutate(percentage = count / sum(count) * 100) %>%
  dplyr::ungroup() 
```

* Region distribution of all control genes with max editing percentage (with Ttn tier1 version)
```{r}
# collect editing sites with max editing %  in controls
all_control_region_maxedit_list <- setNames(
  lapply(names(all_control_list_HT), function(cond) {
    
    df <- all_control_list_HT[[cond]]
    
    df %>%
      group_by(gene_id) %>%
      slice_max(order_by = score, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      mutate(region = ifelse(is.na(region), "Others", region)) %>%
      group_by(region) %>%
      summarise(count = n()) %>%
      arrange(desc(count)) %>%
      mutate(percentage = count / sum(count) * 100,
             rank = cond)}),
  names(all_control_list_HT))

# combine into one df for plotting
control_region_maxedit_counts <- bind_rows(all_control_region_maxedit_list) %>%
  mutate(class = recode(
    rank,
    "only_ADARctrl" = "ADAR specific",
    "only_ΔRRMctrl" = "ΔRRM specific"))
```

the distribution reflects site-level burden and is more strongly influenced by genes with large numbers of editing sites, such as long or highly edited transcripts.

# Compare the region distribution of Matrin3-specific target genes with all control genes (With Ttn tier1 version)
```{r}
tiers <- c("Tier 1", "Tier 2", "Tier 3")

plot_region_tier_df <- bind_rows(
  control_region_maxedit_counts %>%
    filter(class %in% c("ADAR specific", "ΔRRM specific")),
  
  tier_region_maxedit_counts %>%
    mutate(class = rank) %>%  
    filter(class %in% c("Tier 1", "Tier 2", "Tier 3"))) %>%
  mutate(
    class = factor(
      class,
      levels = c("ADAR specific", "ΔRRM specific", "Tier 1", "Tier 2", "Tier 3")),
    region = factor(
      region,
      levels = c("UTR5", "UTR3", "CDS", "Intron", "Others")
    ))

##################### Stacked barplot
ggplot(plot_region_tier_df, aes(x = class, y = percentage, fill = region)) +
  geom_col(color = "black", linewidth = 0.3, width = 0.75) +
  geom_text(
    aes(label = ifelse(percentage >= 3, paste0(round(percentage, 1), "%"), "")),
    position = position_stack(vjust = 0.5),
    size = 4,
    fontface = "bold"
  ) +
  scale_fill_manual(values = region_colors) +
  labs(
    title = "",
    x = NULL,
    y = "Proportion of editing sites",
    fill = "Regions"
  ) +
  coord_cartesian(ylim = c(0, 112), clip = "off") +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    legend.position = "right",
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(color = "black", size = 14, face = "bold", angle = 30, hjust = 1),
    axis.text.y = element_text(size = 14)
  )
```


* For each gene, determine which region has the highest density of editing sites (number of sites / region length in kb) and compare this distribution across tiers and controls to see if certain tiers are enriched for editing in specific regions in terms of region length.
```{r}
# Count editing sites per region per gene
sites_per_region_df <- as.data.frame(m3specific_site_gr_anno) %>%
  mutate(region = ifelse(is.na(region) | region == "", "Others", region)) %>%
  group_by(gene_id, region) %>%
  summarise(n_sites = n(), .groups = "drop")

# Calculate density
density_region_df <- sites_per_region_df %>%
  left_join(region_length_df, by = c("gene_id", "region")) %>%
  mutate(
    site_density = case_when(
      is.na(region_length_kb) ~ NA_real_,     
      TRUE                    ~ n_sites / region_length_kb))

# For each gene, pick the region with the max editing site density
max_density_region_df <- density_region_df %>%
  filter(!is.na(site_density)) %>%
  group_by(gene_id) %>%
  slice_max(order_by = site_density, n = 1, with_ties = FALSE) %>%
  ungroup()

# Count region distribution per tier based on max density region
tier_region_density_counts <- max_density_region_df %>%
  left_join(rank_M3tx_manual_df, by = "gene_id") %>%
  mutate(
    region = ifelse(is.na(region) | region == "", "Others", region),
    rank   = as.character(rank)) %>%
  filter(!is.na(rank)) %>%
  dplyr::count(rank, region, name = "count") %>%
  dplyr::group_by(rank) %>%
  dplyr::mutate(percentage = count / sum(count) * 100) %>%
  dplyr::ungroup() 

```

# Same approach for control samples
```{r}
control_region_density_counts <- bind_rows(
  lapply(names(all_control_list_HT), function(cond) {
    
    df <- all_control_list_HT[[cond]]
    
    # Count sites per region per gene
    sites_per_region <- df %>%
      mutate(region = ifelse(is.na(region) | region == "", "Others", region)) %>%
      group_by(gene_id, region) %>%
      summarise(n_sites = n(), .groups = "drop")
    
    # calculate density
    sites_per_region %>%
      left_join(region_length_df, by = c("gene_id", "region")) %>%
      mutate(
        site_density = case_when(
          is.na(region_length_kb) | region_length_kb == 0 ~ NA_real_,
          TRUE ~ n_sites / region_length_kb)) %>%
      # Pick region with max density per gene
      filter(!is.na(site_density)) %>%
      group_by(gene_id) %>%
      slice_max(order_by = site_density, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      group_by(region) %>%
      summarise(count = n(), .groups = "drop") %>%
      mutate(
        percentage = count / sum(count) * 100,
        class      = cond)})) %>%
  mutate(
    class = recode(
      class,
      "only_ADARctrl" = "ADAR specific",
      "only_ΔRRMctrl" = "ΔRRM specific"))

# combine into one df for plotting
plot_region_density_df <- bind_rows(
  control_region_density_counts %>%
    filter(class %in% c("ADAR specific", "ΔRRM specific")),
  tier_region_density_counts %>%
    mutate(class = rank) %>%
    filter(class %in% c("Tier 1", "Tier 2", "Tier 3"))) %>%
  mutate(
    class = factor(
      class,
      levels = c("ADAR specific", "ΔRRM specific", "Tier 1", "Tier 2", "Tier 3")),
    region = factor(
      region,
      levels = c("UTR5", "UTR3", "CDS", "Intron", "Others")))
```


```{r}
plot_region_density <- ggplot(plot_region_density_df, aes(x = class, y = percentage, fill = region)) +
  geom_col(color = "black", linewidth = 0.3, width = 0.75) +
  geom_text(
    aes(label = ifelse(percentage >= 3, paste0(round(percentage, 1), "%"), "")),
    position = position_stack(vjust = 0.5),
    size = 4,
    fontface = "bold") +
  scale_fill_manual(values = region_colors) +
  labs(
    title = "",
    x = NULL,
    y = "Proportion of regions with \nhighest editing site density",
    fill = "Regions") +
  coord_cartesian(ylim = c(0, 112), clip = "off") +
  theme_bw(base_size = 13) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    legend.position = "right",
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(color = "black", size = 14, face = "bold", angle = 30, hjust = 1),
    axis.text.y = element_text(size = 14))

# save_thesis_plot(
#   plot_region_density,
#   filename = file.path("/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
#                        "region_density_distribution.png"),
#   width = 8,
#   height = 6,
#   dpi = 300,
#   bg = "white")

plot_region_density
```

# Region length distribution among MATR3 target tiers
```{r}
length_plot_df <- region_length_df %>%
  inner_join(rank_M3tx_manual_df %>% dplyr::select(gene_id, rank), by = "gene_id") %>%
  mutate(
    rank = factor(rank, levels = c("Tier 1", "Tier 2", "Tier 3")), 
    region = factor(region, levels = c("UTR5", "CDS", "Intron", "UTR3")) 
  ) %>%
  filter(!is.na(rank))

# 2. Generate the faceted violin plot
region_length_plot <- ggplot(length_plot_df, aes(x = rank, y = log10(region_length_kb), fill = rank)) +
  geom_violin(trim = FALSE, alpha = 0.7, color = "black", linewidth = 0.5) +
  geom_boxplot(width = 0.15, fill = "white", color = "black", outlier.shape = NA, alpha = 0.8) +
  scale_fill_manual(values = tier_colors) +
  facet_wrap(~ region, scales = "free_y", nrow = 1) +
  labs(
    x = "",
    y = "log10(region length (kb))") +
  theme_bw(base_size = 14) +
  theme_thesis() + 
  theme(legend.position = "none",
        axis.text.x = element_text(color = "black", size = 15, face = "bold"))

region_length_plot

```

```{r}
# Annotate editing sites with tier
m3_sites_tier_df <- as.data.frame(m3specific_site_gr_anno) %>%
  left_join(
    rank_M3tx_manual_df %>%
      dplyr::select(gene_id, rank),
    by = "gene_id") %>%
  mutate(
    region = ifelse(
      is.na(region) | region == "",
      "Others",
      region),
    class = rank) %>%
  filter(
    !is.na(class),
    class %in% c("Tier 1", "Tier 2", "Tier 3"),
    region %in% c("UTR5", "UTR3", "CDS", "Intron"))

# Count total editing sites per region within each tier
tier_site_counts <- m3_sites_tier_df %>%
  group_by(class, region) %>%
  summarise(
    total_sites = n(),
    .groups = "drop")

# Calculate total region length within each tier
tier_region_lengths <- rank_M3tx_manual_df %>%
  dplyr::select(gene_id, rank) %>%
  rename(class = rank) %>%
  inner_join(region_length_df, by = "gene_id") %>%
  filter(
    class %in% c("Tier 1", "Tier 2", "Tier 3"),
    region %in% c("UTR5", "UTR3", "CDS", "Intron")) %>%
  group_by(class, region) %>%
  summarise(
    total_region_kb = sum(region_length_kb, na.rm = TRUE),
    .groups = "drop")

# Calculate normalized editing density
# Calculate normalized editing density
tier_density_df <- tier_site_counts %>%
  left_join(
    tier_region_lengths,
    by = c("class", "region")
  ) %>%
  mutate(
    editing_density = total_sites / total_region_kb
  ) %>%
  group_by(class) %>%
  mutate(
    percentage =
      editing_density / sum(editing_density) * 100
  ) %>%
  ungroup()
```



```{r}
# 3. CONTROL DATASETS

control_density_df <- bind_rows(

  lapply(names(all_control_list_HT), function(cond) {

    df <- all_control_list_HT[[cond]]

    # Editing site counts per region
    site_counts <- df %>%
      mutate(
        region = ifelse(
          is.na(region) | region == "",
          "Others",
          region)) %>%
      filter(
        region %in% c("UTR5", "UTR3", "CDS", "Intron")
      ) %>%
      group_by(region) %>%
      summarise(
        total_sites = n(),
        .groups = "drop")

    # Genes present in this control dataset
    genes_in_control <- unique(df$gene_id)

    # Total region length across genes
    region_lengths <- region_length_df %>%
      filter(
        gene_id %in% genes_in_control,
        region %in% c("UTR5", "UTR3", "CDS", "Intron")) %>%
      group_by(region) %>%
      summarise(
        total_region_kb = sum(region_length_kb, na.rm = TRUE),
        .groups = "drop")

    # Density
    site_counts %>%
      left_join(region_lengths, by = "region") %>%
      mutate(
        editing_density = total_sites / total_region_kb,
        percentage = editing_density / sum(editing_density) * 100,
        class = cond)})) %>%
  mutate(
    class = recode(
      class,
      "only_ADARctrl" = "ADAR specific",
      "only_ΔRRMctrl" = "ΔRRM specific")) 

# 4. COMBINE MATR3 + CONTROLS

plot_region_density_df <- bind_rows(
  control_density_df,
  tier_density_df) %>%
  mutate(
    class = factor(
      class,
      levels = c(
        "ADAR specific",
        "ΔRRM specific",
        "Tier 1",
        "Tier 2",
        "Tier 3")),
    region = factor(
      region,
      levels = c("UTR5", "CDS", "Intron", "UTR3")))

# 5. PLOT
plot_region_density <- ggplot(
  plot_region_density_df,
  aes(
    x = class,
    y = percentage,
    fill = region)) +
 geom_col(color = "black", position = "fill",linewidth = 0.3, width = 0.75) +
  geom_text(
    aes(label = ifelse(percentage >= 3, paste0(round(percentage, 1), "%"), "")),
    position = position_fill(vjust = 0.5),
    size = 4,
    fontface = "bold") +
  scale_fill_manual(values = region_colors) +
  labs(
    x = NULL,
    y = "Proportion of normalized editing density",
    fill = "Regions"
  ) +
   theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    legend.position = "right",
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(color = "black", size = 14, face = "bold", angle = 30, hjust = 1),
    axis.text.y = element_text(size = 14))

plot_region_density

# Save plot
# ============================================================

save_thesis_plot(
  plot_region_density,
  filename = file.path(
    "/Users/tinkienoob/Master_thesis/MATR3/Tier_targets_png",
    "tier_region_distribution.png"
  ),
  width = 7.5,
  height = 6,
  dpi = 300,
  bg = "white"
)
```

```{r}
############################################################
# Gene-level regional editing density distribution
# Tier vs controls
############################################################

library(dplyr)
library(ggplot2)
library(ggpubr)

############################################################
# 1. MATR3 TIER GENE-LEVEL DENSITY
############################################################

m3_gene_region_density <- as.data.frame(m3specific_site_gr_anno) %>%
  
  left_join(
    rank_M3tx_manual_df %>%
      dplyr::select(gene_id, rank),
    by = "gene_id"
  ) %>%
  
  mutate(
    region = ifelse(
      is.na(region) | region == "",
      "Others",
      region
    ),
    class = rank
  ) %>%
  
  filter(
    !is.na(class),
    class %in% c("Tier 1", "Tier 2", "Tier 3"),
    region %in% c("UTR5", "UTR3", "CDS", "Intron")
  ) %>%
  
  # count editing sites PER GENE
  group_by(gene_id, class, region) %>%
  summarise(
    total_sites = n(),
    .groups = "drop"
  ) %>%
  
  # add region length PER GENE
  left_join(
    region_length_df %>%
      filter(
        region %in% c("UTR5", "UTR3", "CDS", "Intron")
      ),
    by = c("gene_id", "region")
  ) %>%
  
  mutate(
    editing_density = total_sites / region_length_kb,
    
    # log transform
    log_density = log10(editing_density)
  )

############################################################
# 2. CONTROL GENE-LEVEL DENSITY
############################################################

control_gene_region_density <- bind_rows(

  lapply(names(all_control_list_HT), function(cond) {

    df <- all_control_list_HT[[cond]]

    df %>%
      
      mutate(
        region = ifelse(
          is.na(region) | region == "",
          "Others",
          region
        )
      ) %>%
      
      filter(
        region %in% c("UTR5", "UTR3", "CDS", "Intron")
      ) %>%
      
      # count editing sites PER GENE
      group_by(gene_id, region) %>%
      summarise(
        total_sites = n(),
        .groups = "drop"
      ) %>%
      
      # add region length PER GENE
      left_join(
        region_length_df %>%
          filter(
            region %in% c("UTR5", "UTR3", "CDS", "Intron")
          ),
        by = c("gene_id", "region")
      ) %>%
      
      mutate(
        editing_density = total_sites / region_length_kb,
        log_density = log10(editing_density),
        class = cond
      )

  })

) %>%
  
  mutate(
    class = recode(
      class,
      "only_ADARctrl" = "ADAR specific",
      "only_ΔRRMctrl" = "ΔRRM specific"
    )
  )

############################################################
# 3. COMBINE
############################################################

plot_violin_df <- bind_rows(
  control_gene_region_density,
  m3_gene_region_density
) %>%
  
  mutate(
    class = factor(
      class,
      levels = c(
        "ADAR specific",
        "ΔRRM specific",
        "Tier 1",
        "Tier 2",
        "Tier 3"
      )
    ),
    
    region = factor(
      region,
      levels = c(
        "UTR5",
        "CDS",
        "Intron",
        "UTR3"
      )
    )
  )

############################################################
# 4. STATISTICAL COMPARISONS
############################################################

comparisons <- list(
  c("Tier 1", "Tier 2"),
  c("Tier 2", "Tier 3"),
  c("Tier 1", "Tier 3"))


############################################################
# 5. VIOLIN PLOT
############################################################

plot_region_violin <- ggplot(
  plot_violin_df,
  aes(
    x = class,
    y = log_density,
    fill = class
  )
) +
  
  geom_violin(
    trim = FALSE,
    alpha = 0.7,
    color = "black"
  ) +
  
  geom_boxplot(
    width = 0.12,
    outlier.shape = NA,
    alpha = 0.9
  ) +
  
  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif",
    hide.ns = TRUE,
    size = 4
  ) +
  
  facet_wrap(
    ~region,
    scales = "free_y"
  ) +
  
  labs(
    x = NULL,                                             
   y    = "Editing density (sites / kb, log10 scale)") +
  
  theme_bw(base_size = 14) +
  
  theme(
    strip.text = element_text(
      size = 14,
      face = "bold"
    ),
    
    legend.position = "none",
    
    axis.title = element_text(
      face = "bold",
      size = 14
    ),
    
    axis.text.x = element_text(
      color = "black",
      size = 12,
      face = "bold",
      angle = 30,
      hjust = 1
    ),
    
    axis.text.y = element_text(
      size = 12,
      color = "black"
    )
  )

plot_region_violin
```

# GO compare Tier
```{r}
# remove suffix 
# tier_classify_df$GeneID <- gsub("\\.\\d+$", "", tier_classify_df$gene_id)
# 
# cc_bp <- compareCluster(
#   geneCluster   = tier_classify_df$GeneID,
#   fun           = "enrichGO",
#   OrgDb         = org.Mm.eg.db,
#   keyType       = "ENSEMBL",
#   ont           = "ALL",
#   universe      = gene_anno_df$GeneID,
#   pAdjustMethod = "BH",
#   pvalueCutoff  = 0.05,
#   qvalueCutoff  = 0.05,
#   readable      = TRUE)

```

```{r}
# dotplot_cc_bp <- dotplot(
#   cc_bp,
#   showCategory = 10,
#   title        = "GO Enrichment Comparison across MATR3 Target Tiers",
#   font.size    = 15,
#   label_format = 30) +
#   theme_light(base_size = 15) +
#   theme(
#     plot.title       = element_text(hjust = 0.5, face = "bold", size = 20),
#     strip.text       = element_text(size = 17, face = "bold", color = "white"),
#     strip.background = element_rect(fill = "grey20"),
#     axis.title.x     = element_blank(),
#     axis.text.x      = element_text(color = "black", size = 15, face = "bold"),
#     axis.text.y      = element_text(color = "black", size = 13),
#     legend.title     = element_text(size = 15, face = "bold"),
#     legend.text      = element_text(size = 14),
#     legend.key.size  = unit(0.9, "cm"),
#     panel.grid.minor = element_blank(),
#     plot.margin      = margin(10, 15, 10, 10)) +
#   guides(
#     color = guide_colorbar(title = "Adjusted\np-value"),
#     size  = guide_legend(title = "Gene\nCount"))

# save_thesis_plot(
#   plot = dotplot_cc_bp,
#   filename = "/Users/tinkienoob/Master_thesis/MATR3/Deseq2/GO_deseq2_png/compareCluster_tiers.png",
#   width = 10,
#   height = 12)
```

```{r}
sessionInfo()
```


